---
title: "Templates for All Maps and Graphs + Descriptive Statistics"
description: "This document is meant to act as the location where all final version graphs are meant to be created, and all descriptive statistics put onto those graphs will be calcualted. This synthesizes material from other documents up to this point. "
author: "Sam Lance"
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
    theme: Lux
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r, include =FALSE}
library(dplyr)
library(tidyverse)
library(lubridate)
library(readr)
library(here)
library(ggplot2)
library(stringr)
library(purrr)
library(maps)
library(knitr)
library(kableExtra)
library(sf)
library(RColorBrewer)
library(patchwork)
library(tmap)
library(viridis)
library(sp)
library(rworldmap)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(devtools)
library(janitor)
library(svglite)
```

## Reading In + Cleaning Data

### Reading in Data

```{r, output=FALSE}}
#DISCHARGE
join_dis <- read_csv(here("data","vg_dis_z.csv")) |>
  inner_join(
    read_csv(here("data","base_sites.csv")) |>
      select(site_name, site_lat, site_lon),
    by = "site_name") |> #og data missing these parts, need to join
  mutate(
    zlog_p95_q  = coalesce(zlog_p95_q, z_p95_q),
    zlog_mean_q = coalesce(zlog_mean_q, z_mean_q)
  ) #coalesce falls back to the second option if the first is not available, doing because issue with intermittent flows at 10 sites 

#Sediment 
join_sed = read_csv(here("data","all_site_ssc_z.csv")) |>
  drop_na()



```

### Count by Type

```{r}

count_by_type <- join_dis |>
  group_by(av_type) |>
  summarise(n = n()) |>
  ungroup() 

count_by_type_ssc <- join_sed |>
  group_by(av_type) |>
  summarise(n = n()) |>
  ungroup()

```

### Cleaning Discharge Data

```{r}
#Convert DF to SF Object 
sf_dis <- st_as_sf(join_dis, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_dis, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_dis <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

#Export as CSV 
write_csv(sf_dis, "discharge_all_sites.csv")

```

### Cleaning Sediment Data

```{r}

#Convert DF to SF Object 
sf_sed <- st_as_sf(join_sed, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_sed, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_sed <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

```

### Stats for Paper

```{r}

# Discharge + SSC Overall
dis_stats <- join_dis |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE),
    mean_log_durz = mean(zlog_flood_dur, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_q, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_q, mu = 0)$p.value,
    p_log_durz = t.test(zlog_flood_dur, mu = 0)$p.value
  )

ssc_stats <- join_sed |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_ssc, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_ssc, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_ssc, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_ssc, mu = 0)$p.value
  )

# Summarize Same Way but for Avulsion Types
dis_stats_type <- join_dis |>
  group_by(av_type) |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE),
    mean_log_durz = mean(zlog_flood_dur, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_q, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_q, mu = 0)$p.value,
    p_log_durz = t.test(zlog_flood_dur, mu = 0)$p.value
  )

ssc_stats_type <- join_sed |>
  group_by(av_type) |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_ssc, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_ssc, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_ssc, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_ssc, mu = 0)$p.value
  )
```

### Reading in Full Time Series Data

```{r}

#Load Discharge Data 
discharge_ts <- read_csv(here("data", "vg_dis_z_all_years.csv"))

```

### Pulling out 5 Years Prior to Avulsion

```{r}
#Pull Out 5 Years Prior to Avulsion
pre_avulsion_5 <- discharge_ts |>
  filter(year >= (avulsion_year - 5) & year < avulsion_year) |>
  filter(avulsion_year > 1989)

```

### Conducting Some Proportion Stats

#### Number of Points Above 0 for Each Av Type

```{r}
#Count Number of Points Above 0 fore Each Avulsion Type 
pre_av_point_count <- pre_avulsion_5 |>
  group_by(av_type) %>%
  summarise(
    total_count = n(),
    
    #Calculating the Counts 
    zlog_p95_q_above_zero  = sum(zlog_p95_q > 0,  na.rm = TRUE),
    zlog_mean_q_above_zero = sum(zlog_mean_q > 0, na.rm = TRUE),
    zlog_flood_dur_above_zero = sum(zlog_flood_dur > 0, na.rm = TRUE)) |>
  
  mutate(max_above_prop = zlog_p95_q_above_zero/total_count) |>
  mutate(mean_above_prop = zlog_mean_q_above_zero/total_count) |>
  mutate(flood_dur_above_prop = zlog_flood_dur_above_zero/total_count) |>
  rename(avulsion_type = av_type)

```

#### How Many Sites Have Zscores \> 0 All Five Years

```{r}

#Find How Many Sites Have All ZScores >0

result_sites <- pre_avulsion_5 %>%
  filter(year < avulsion_year) %>%
  group_by(av_type, site_name) %>%
  summarise(
    total_points = n(),
    
    zlog_p95_q_all_above     = all(zlog_p95_q > 0,     na.rm = TRUE),
    zlog_mean_q_all_above    = all(zlog_mean_q > 0,    na.rm = TRUE),
    zlog_flood_dur_all_above = all(zlog_flood_dur > 0, na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  group_by(av_type) %>%
  summarise(
    total_sites = n_distinct(site_name),
    
    zlog_p95_q_sites_all_above     = sum(zlog_p95_q_all_above),
    zlog_mean_q_sites_all_above    = sum(zlog_mean_q_all_above),
    zlog_flood_dur_sites_all_above = sum(zlog_flood_dur_all_above),
    
    zlog_p95_q_prop_all_above     = sum(zlog_p95_q_all_above)     / total_sites,
    zlog_mean_q_prop_all_above    = sum(zlog_mean_q_all_above)    / total_sites,
    zlog_flood_dur_prop_all_above = sum(zlog_flood_dur_all_above) / total_sites,
    
    .groups = "drop"
  ) %>%
  rename(avulsion_type = av_type)

```

#### How Many Sites Have Year Prior to Av \> 2 YP

```{r}

greater_than <- pre_avulsion_5 %>%
  filter(year %in% c(avulsion_year - 2, avulsion_year - 1)) %>%
  group_by(av_type, site_name) %>%
  summarise(
    zlog_p95_q_yr1_gt_yr2     = zlog_p95_q[year == avulsion_year - 1]     > zlog_p95_q[year == avulsion_year - 2],
    zlog_mean_q_yr1_gt_yr2    = zlog_mean_q[year == avulsion_year - 1]    > zlog_mean_q[year == avulsion_year - 2],
    zlog_flood_dur_yr1_gt_yr2 = zlog_flood_dur[year == avulsion_year - 1] > zlog_flood_dur[year == avulsion_year - 2],
    .groups = "drop"
  ) %>%
  group_by(av_type) %>%
  summarise(
    total_sites = n_distinct(site_name),
    
    zlog_p95_q_count_yr1_gt_yr2     = sum(zlog_p95_q_yr1_gt_yr2,     na.rm = TRUE),
    zlog_mean_q_count_yr1_gt_yr2    = sum(zlog_mean_q_yr1_gt_yr2,    na.rm = TRUE),
    zlog_flood_dur_count_yr1_gt_yr2 = sum(zlog_flood_dur_yr1_gt_yr2, na.rm = TRUE),
    
    zlog_p95_q_prop_yr1_gt_yr2     = sum(zlog_p95_q_yr1_gt_yr2,     na.rm = TRUE) / total_sites,
    zlog_mean_q_prop_yr1_gt_yr2    = sum(zlog_mean_q_yr1_gt_yr2,    na.rm = TRUE) / total_sites,
    zlog_flood_dur_prop_yr1_gt_yr2 = sum(zlog_flood_dur_yr1_gt_yr2, na.rm = TRUE) / total_sites,
    
    .groups = "drop"
  ) %>%
  rename(avulsion_type = av_type)

print(greater_than)

```

## Figure 1 - World Map with Avulsion Types

### Establishing Colors and Shapes for All Maps

```{r}
#Loading in World Map Data
world_map <- ne_countries(scale = "medium", returnclass = "sf")

shapes <- c(
  "delta"             = 21,
  "fan"               = 22,
  "intra-lobe"        = 23,
  "retrogradational"  = 24
)

colors <- plasma(4)
av_type_colors <- setNames(colors, c("delta", "fan", "intra-lobe", "retrogradational"))
print(av_type_colors)
combined_colors <- c("All Sites" = "grey50", av_type_colors)
print(names(combined_colors))

# #devtools::install_github("an-bui/calecopal")# filled shapes
# colors <- c(
#   "delta"             = "#0D0887FF",
#   "fan"               = "#9C179EFF",
#   "intra-lobe"        = "#F89441FF",
#   "retrogradational"  = "#F0F921FF"
# )
# breaks <- c(-2, -1, 0, 1, 2, 3, 4, 5)
# colors <- brewer.pal(length(breaks) - 1, "RdYlBu")
# rev_colors <- rev(colors)  # Reverse the color order
# shapes <- c(16, 17, 15, 18)  # Different shapes for avulsion types

```

### Robinson Projection World Map

```{r}

library(tidyverse)
library(sf)
library(ggplot2)

# -------------------------
# Inputs assumed to exist:
# world_map (sf polygons, lon/lat)
# sf_dis (dataframe with site_lon, site_lat, site_name, av_type)
# rev_colors (vector of colors)
# shapes (named vector of shapes for av_type)
# -------------------------

# 1) Select sites to highlight
selected_sites <- sf_dis |>
  filter(site_name %in% c("B9", "R3.2", "JA.25.India.1990", "MANAMBOLO.2004"))

# 2) Ensure factor levels consistent
sf_dis <- sf_dis |>
  mutate(av_type = factor(av_type))

# 3) Named color vector for av_type
av_type_colors <- setNames(av_type_colors, levels(sf_dis$av_type))

# 4) Robinson CRS
crs_robin <- "+proj=robin +lon_0=0 +datum=WGS84 +units=m +no_defs"

# 5) Make points sf (critical for correct projection)
sf_dis_sf <- st_as_sf(
  sf_dis,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

selected_sites_sf <- st_as_sf(
  selected_sites,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

# 6) Ensure world_map has CRS (if not already)
if (is.na(st_crs(world_map))) st_crs(world_map) <- 4326

# 7) Create a "globe" outline and optional graticule
globe_outline <- st_as_sfc(
  st_bbox(c(xmin = -180, ymin = -90, xmax = 180, ymax = 90), crs = st_crs(4326))
) |>
  st_transform(crs_robin)

# Optional graticule
grat <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = seq(-60, 60, by = 30),
  crs = st_crs(4326)
) |>
  st_transform(crs_robin)

# -------------------------
# NEW: Top + bottom boundary (latitude caps)
# -------------------------
lat_min <- -90
lat_max <- 90
lon_seq <- seq(-180, 180, by = 0.25)

make_parallel <- function(lat) {
  st_sfc(
    st_linestring(cbind(lon_seq, rep(lat, length(lon_seq)))),
    crs = 4326
  ) |>
    st_transform(crs_robin)
}

bottom_boundary <- make_parallel(lat_min)
top_boundary    <- make_parallel(lat_max)

# 8) Build plot (Robinson + globe boundary)
world_map_plot <- ggplot() +

  # Light graticule behind everything (optional; remove if you don’t want lines)
  geom_sf(data = grat, color = "grey70", linewidth = 0.25) +

  # Continents
  geom_sf(
    data = world_map,
    fill = "grey70",
    color = "grey60",
    linewidth = 0.25
  ) +

  # All sites
  geom_sf(
    data = sf_dis_sf,
    aes(fill = av_type, shape = av_type, color = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 0.5
  ) +

  # Selected sites: same fill/shape, red outline
  geom_sf(
    data = selected_sites_sf,
    aes(fill = av_type, shape = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 1.0,
    color = "black",
    show.legend = FALSE
  ) +

  # -------------------------
  # NEW: draw top/bottom boundary lines
  # -------------------------
  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  # Globe outline (draw last so it frames everything)
  geom_sf(
    data = globe_outline,
    fill = NA,
    color = "grey70",
    linewidth = 0.7
  ) +

  # Scales
  scale_color_manual(name = "Avulsion Type", values = colors) +
  scale_fill_manual(name = "Avulsion Type", values = colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +

  # Robinson projection
  coord_sf(crs = crs_robin, expand = FALSE) +

  # Styling
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    
    #legend.position = "none",
    legend.position = c(0.2, 0.25),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    legend.spacing.y = unit(-0.1, "cm"),
    legend.key.height = unit(0.3, "lines"),
    legend.background = element_rect(fill = "white", color = "white")
  ) +
  guides(
    fill  = guide_legend(byrow = TRUE),
    shape = guide_legend(byrow = TRUE)
  )

world_map_plot

ggsave(
  filename = here("plots_svg", "fig1_world_robinson_globe.svg"),
  plot = world_map_plot,
  width = 10,
  height = 7,
  device = svglite, 
  units = "in",
  dpi = 300
)

```
## Figure 2 Version 2 

```{r}

# -------------------------
# Paths
# -------------------------
av_year_updated_path <- here("data", "ssc_av_year_z.csv")
compiled_output_path <- here("data", "time_series.csv")
av_year_path_dis     <- here("data", "discharge_av_year_z.csv")
compiled_path_dis    <- here("data", "compiled_dis24.csv")
plots_dir <- here("plots_svg")

# -------------------------
# Load SSC data
# -------------------------
av_year_ssc  <- read_csv(av_year_updated_path, show_col_types = FALSE)
compiled_ssc <- read_csv(compiled_output_path, show_col_types = FALSE) %>%
  mutate(date = as.Date(date))

# -------------------------
# Load Discharge data
# -------------------------
av_year_dis  <- read_csv(av_year_path_dis, show_col_types = FALSE)
compiled_dis <- read_csv(compiled_path_dis, show_col_types = FALSE) %>%
  mutate(valid_time = as.POSIXct(valid_time, tz = "UTC"))

# -------------------------
# SELECT ONE SITE
# -------------------------
site <- "MALAWI1.2014"

site_ssc <- compiled_ssc %>% filter(site_name == site) %>% arrange(date)
site_dis <- compiled_dis %>% filter(site_name == site) %>% arrange(valid_time)

stopifnot(nrow(site_ssc) > 0, nrow(site_dis) > 0)

# Align discharge to Date for dual-axis plot
site_dis <- site_dis %>%
  mutate(date = as.Date(valid_time))

# -------------------------
# Avulsion year
# -------------------------
get_avulsion_year <- function(df) {
  if ("avulsion_start_date" %in% names(df)) {
    df %>% pull(avulsion_start_date) %>% as.numeric() %>% na.omit() %>% unique()
  } else {
    numeric(0)
  }
}

avs <- get_avulsion_year(site_ssc)

# -------------------------
# Shading windows
# -------------------------
shade_avulsion  <- NULL
shade_pre       <- NULL

if (length(avs) > 0) {
  av_yr <- avs[1]

  # Avulsion window: year-1 through year (slightly darker yellow)
  shade_avulsion <- tibble(
    xmin = as.Date(paste0(av_yr - 1, "-01-01")),
    xmax = as.Date(paste0(av_yr,     "-12-31")),
    ymin = -Inf, ymax = Inf
  )

  # Pre-avulsion window: 5 years before through 1 year before (lighter yellow)
  shade_pre <- tibble(
    xmin = as.Date(paste0(av_yr - 6, "-01-01")),
    xmax = as.Date(paste0(av_yr - 2, "-12-31")),
    ymin = -Inf, ymax = Inf
  )
}

# -------------------------
# p95 lines
# -------------------------
year_p95_ssc <- av_year_ssc %>% filter(site_name == site) %>% pull(p95_mean) %>% first()
year_p95_dis <- av_year_dis %>% filter(site_name == site) %>% pull(p95_mean) %>% first()

# -------------------------
# p95 points within avulsion window
# -------------------------
get_p95_pt_ssc <- function(df, shade) {
  if (is.null(shade)) return(NULL)
  w <- df %>% filter(date >= shade$xmin, date <= shade$xmax, !is.na(y_pred))
  if (nrow(w) == 0) return(NULL)
  p95_val <- as.numeric(quantile(w$y_pred, 0.95, na.rm = TRUE, type = 7))
  max_row <- w %>% filter(y_pred == max(y_pred, na.rm = TRUE)) %>% arrange(date) %>% slice(1)
  tibble(date = max_row$date, y_pred = p95_val)
}

get_p95_pt_dis <- function(df, shade) {
  if (is.null(shade)) return(NULL)
  w <- df %>% filter(date >= shade$xmin, date <= shade$xmax, !is.na(dis24))
  if (nrow(w) == 0) return(NULL)
  p95_val <- as.numeric(quantile(w$dis24, 0.95, na.rm = TRUE, type = 7))
  max_row <- w %>% filter(dis24 == max(dis24, na.rm = TRUE)) %>% arrange(date) %>% slice(1)
  tibble(date = max_row$date, dis24 = p95_val)
}

p95_pt_ssc <- get_p95_pt_ssc(site_ssc, shade_avulsion)
p95_pt_dis <- get_p95_pt_dis(site_dis, shade_avulsion)

# -------------------------
# Scaling factor for dual y-axis
# (maps discharge range onto SSC range)
# -------------------------
ssc_range <- range(site_ssc$y_pred, na.rm = TRUE)
dis_range <- range(site_dis$dis24,  na.rm = TRUE)

# linear transform: dis -> ssc space
# dis_scaled = (dis - dis_min) / (dis_max - dis_min) * (ssc_max - ssc_min) + ssc_min
scale_dis <- function(x) {
  (x - dis_range[1]) / diff(dis_range) * diff(ssc_range) + ssc_range[1]
}
unscale_dis <- function(x) {
  (x - ssc_range[1]) / diff(ssc_range) * diff(dis_range) + dis_range[1]
}

site_dis <- site_dis %>% mutate(dis_scaled = scale_dis(dis24))

# Scale p95 points and lines into SSC space
p95_pt_dis_scaled <- if (!is.null(p95_pt_dis)) {
  p95_pt_dis %>% mutate(dis_scaled = scale_dis(dis24))
} else NULL

year_p95_dis_scaled <- if (!is.na(year_p95_dis)) scale_dis(year_p95_dis) else NA

# -------------------------
# Shared shading helper
# -------------------------
shade_layers <- function() {
  list(
    if (!is.null(shade_pre))
      geom_rect(
        data = shade_pre,
        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
        inherit.aes = FALSE,
        fill  = "#fff7bc",   # back to original lighter yellow
        alpha = 0.6
      ),
    if (!is.null(shade_avulsion))
      geom_rect(
        data = shade_avulsion,
        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
        inherit.aes = FALSE,
        fill  = "#ffea4d",   # slightly lighter than before
        alpha = 0.45
      )
  )
}

# -------------------------
# Shared x scale + theme
# -------------------------
shared_x <- scale_x_date(
  limits = as.Date(c("1983-01-01", "2020-12-31")),
  breaks = seq(as.Date("1983-01-01"), as.Date("2020-01-01"), by = "5 years"),
  labels = date_format("%Y"),
  expand = c(0, 0)
)

shared_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.25, color = "grey90"),
    axis.title.y       = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x       = element_blank(),
    axis.line.x        = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y        = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x        = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y        = element_text(color = "grey30", size = 16, face = "bold"),
    axis.ticks.y       = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks.x       = element_line(color = "grey30", linewidth = 1.2),
    legend.position    = "none",
    plot.background    = element_rect(fill = "white", colour = NA)
  )

# -------------------------
# Combined SSC + Discharge plot (dual y-axis)
# -------------------------
p_combined <- ggplot(site_ssc, aes(x = date, y = y_pred)) +

  shade_layers() +
  
  # Discharge line + points (scaled to left axis)
  geom_line(data  = site_dis, aes(x = date, y = dis_scaled),
            inherit.aes = FALSE, color = "#2f6fa5", linewidth = 0.75) +
  geom_point(data = site_dis, aes(x = date, y = dis_scaled),
             inherit.aes = FALSE, color = "#2f6fa5", size = 0.75) +

  # SSC line + points (left axis)
  geom_line(color  = "#9bbf30", linewidth = 0.75) +
  geom_point(color = "#9bbf30", size = 0.75) +

  # Dual y-axis: left = SSC, right = discharge (back-transformed)
  scale_y_continuous(
    name   = "Discharge (m³/s)",
    expand = c(0, 0.05),
    labels = function(x) round(unscale_dis(x), 0),  # show discharge values on left
    sec.axis = sec_axis(
      transform = ~ .,  # SSC already in native units on right
      name      = "SSC (mg/L)"
    )
  ) +

  shared_x +

  labs(x = "Year") +

  shared_theme 

# -------------------------
# Save combined plot
# -------------------------
ggsave(
  filename = file.path(plots_dir, "fig2_combined_timeseries.png"),
  plot     = p_combined,
  width    = 9, height = 3, dpi = 300, bg = "white"
)

p_combined

```

```{r}

# =========================================================
# BOTTOM PLOT: Z-score time series (unchanged structure,
# updated shading windows)
# =========================================================

malawi_dis_zscores <- read_csv(here("data", "discharge_zscores_one_site.csv"))
malawi_ssc_zscores <- read_csv(here("data", "ssc_zscores_one_site.csv"))

prep_df <- function(df, label) {
  df %>%
    mutate(
      year       = as.integer(year),
      zlog_p95_q = as.numeric(zlog_p95_q),
      dataset    = label,
      date       = as.Date(paste0(year, "-01-01"))
    ) %>%
    filter(!is.na(year), !is.na(zlog_p95_q))
}

dis_red  <- prep_df(malawi_dis_zscores, "Discharge (m^3/s)")
ssc_blue <- prep_df(malawi_ssc_zscores, "SSC (mg/L)")
df_all   <- bind_rows(dis_red, ssc_blue)
stopifnot(nrow(df_all) > 0)

# ---- Shading windows for z-score plot ----
# Avulsion year assumed = 2014 (adjust if needed)
av_yr_z <- 2014L

shade_avulsion_z <- tibble(
  xmin = as.Date(paste0(av_yr_z - 1, "-01-01")),
  xmax = as.Date(paste0(av_yr_z,     "-12-31")),
  ymin = -Inf, ymax = Inf
)

shade_pre_z <- tibble(
  xmin = as.Date(paste0(av_yr_z - 6, "-01-01")),
  xmax = as.Date(paste0(av_yr_z - 2, "-12-31")),
  ymin = -Inf, ymax = Inf
)

# ---- Marker points ----
mark_year_dis <- 2014L
mark_year_ssc <- 2013L

mark_pts_dis <- df_all %>% filter(dataset == "Discharge (m^3/s)", year == mark_year_dis)
mark_pts_ssc <- df_all %>% filter(dataset == "SSC (mg/L)",        year == mark_year_ssc)

# ---- Plot ----
p_z <- ggplot(df_all, aes(x = date, y = zlog_p95_q, color = dataset)) +

  # Pre-avulsion window (lighter yellow)
  geom_rect(
    data = shade_pre_z,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill  = "#fff7bc",
    alpha = 0.55
  ) +

  # Avulsion window (slightly darker yellow)
  geom_rect(
    data = shade_avulsion_z,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill  = "#ffea4d",
    alpha = 0.45
  ) +

  geom_line(linewidth  = 0.9) +
  geom_point(size = 1.8) +

  { if (nrow(mark_pts_dis) > 0)
    geom_point(data = mark_pts_dis, aes(x = date, y = zlog_p95_q),
               inherit.aes = FALSE, shape = 21, size = 5, stroke = 2,
               fill = NA, color = "red", alpha = 0.8)
  } +

  { if (nrow(mark_pts_ssc) > 0)
    geom_point(data = mark_pts_ssc, aes(x = date, y = zlog_p95_q),
               inherit.aes = FALSE, shape = 21, size = 5, stroke = 2,
               fill = NA, color = "red", alpha = 0.8)
  } +

  scale_color_manual(values = c(
    "Discharge (m^3/s)" = "#2f6fa5",
    "SSC (mg/L)"        = "#9bbf30"
  )) +

  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +

  scale_x_date(
    limits = c(as.Date("1983-01-01"), as.Date("2020-12-31")),
    breaks = seq(as.Date("1983-01-01"), as.Date("2020-01-01"), by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(x = "Year", y = "Z-Score (-)") +

  shared_theme +
  theme(
    axis.title.x     = element_text(size = 20, face = "bold", color = "grey30"),
    legend.position  = "none",
    legend.direction = "horizontal",
    legend.title     = element_blank(),
    legend.text      = element_text(size = 14, face = "bold", color = "grey30"),
    legend.background = element_rect(fill = alpha("white", 0.7), colour = NA),
    legend.key        = element_rect(fill = NA, colour = NA)
  )

p_z <- p_z +
  theme(
    # Match font sizes to combined plot
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 20, face = "bold"),
    axis.text    = element_text(color = "grey30", size = 16, face = "bold"),

    # Add right-side margin to account for the second y-axis on the combined plot
    plot.margin = margin(t = 5.5, r = 50, b = 5.5, l = 5.5, unit = "pt")
  )

# ---- Save ----
ggsave(
  filename = file.path(plots_dir, "fig2_log_zscore.png"),
  plot     = p_z,
  width    = 9, height = 3.1,dpi = 300, bg = "white"
)

p_z

```
## Figure 2 - Time Series

### Sediment

```{r}

# -------------------------
# Paths
# -------------------------
av_year_updated_path <- here("data", "ssc_av_year_z.csv")
compiled_output_path <- here("data", "time_series.csv")
plots_dir <- "/Volumes/Crucial X6/paper_figures/rstudio/plots_svg"

# -------------------------
# Load data
# -------------------------
av_year_updated <- read_csv(av_year_updated_path, show_col_types = FALSE)
compiled <- read_csv(compiled_output_path, show_col_types = FALSE)

compiled <- compiled %>%
  mutate(date = as.Date(date))


# -------------------------
# SELECT ONE SITE HERE
# -------------------------
site <- "MALAWI1.2014"

site_df <- compiled %>%
  filter(site_name == site) %>%
  arrange(date)

# Check for NAs in your relevant columns
sum(is.na(site_df$y_pred))

# Check if any points fall outside your axis range
#df |> filter(y < your_lower_limit | y > your_upper_limit)

stopifnot(nrow(site_df) > 0)

# -------------------------
# Avulsion shading window
# -------------------------
avs <- if ("avulsion_start_date" %in% names(site_df)) {
  site_df %>%
    pull(avulsion_start_date) %>%
    as.numeric() %>%
    na.omit() %>%
    unique()
} else {
  numeric(0)
}

shade_df <- NULL
if (length(avs) > 0) {
  shade_df <- tibble(
    xmin = as.Date(paste0(avs[1] - 1, "-01-01")),
    xmax = as.Date(paste0(avs[1], "-12-31")),
    ymin = -Inf,
    ymax = Inf
  )
}

# -------------------------
# Mean 95th percentile line (from zscore table)
# -------------------------
year_p95 <- av_year_updated %>%
  filter(site_name == site) %>%
  pull(p95_mean) %>%
  first()

# -------------------------
# 95th percentile value within the shaded window,
# plotted at the exact date of the maximum within the window
# -------------------------
p95_pt <- NULL

if (!is.null(shade_df)) {
  window_df <- site_df %>%
    filter(date >= shade_df$xmin, date <= shade_df$xmax) %>%
    filter(!is.na(y_pred))

  if (nrow(window_df) > 0) {
    p95_val <- as.numeric(quantile(window_df$y_pred, probs = 0.95, na.rm = TRUE, type = 7))

    max_row <- window_df %>%
      filter(y_pred == max(y_pred, na.rm = TRUE)) %>%
      arrange(date) %>%   # if ties, use earliest max date
      slice(1)

    p95_pt <- tibble(
      date = max_row$date,
      y_pred = p95_val
    )
  }
}

# -------------------------
# Plot
# -------------------------
p_ssc <- ggplot(site_df, aes(x = date, y = y_pred)) +

  # Avulsion shading
  {if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
    # Mean p95 line
  {if (!is.na(year_p95))
    geom_hline(
      yintercept = year_p95,
      linetype = "dashed",
      color = "black",
      linewidth = 0.75,
      alpha = 0.75
    )
  } +

  # Time series
  geom_point(color = "#9bbf30", size = 0.75) +
  geom_line(color = "#9bbf30", linewidth = 0.75) +

# Red hollow circle: y = p95(window), x = date of max(window)
{ if (!is.null(p95_pt))
  geom_point(
    data = p95_pt,
    aes(x = date, y = y_pred),
    inherit.aes = FALSE,
    shape = 21,        # hollow circle
    size = 5,
    stroke = 2,      # outline thickness
    fill = NA,
    color = "red",
    alpha = 0.7
  )
} +

  # Axes (keep only scale_x_date to avoid scale replacement warnings)
  scale_x_date(
    limits = as.Date(c("1983-01-01", "2020-12-31")),
    breaks = seq(as.Date("1983-01-01"),
                 as.Date("2020-01-01"),
                 by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = c(0, 0)) +

  labs(
    x = "Year",
    y = "SSC (mg/L)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(
      linetype  = "solid",
      linewidth = 0.25,
      color = "grey90"),
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "white", colour = NA)
  )

# -------------------------
# Save
# -------------------------
safe_name <- site %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("/", "_")

ggsave(
  filename = file.path(plots_dir, "fig2_ssc_timeseries.svg"),
  plot = p_ssc,
  width = 9,
  device = svglite,
  height = 3,
  dpi = 300,
  bg = "white"
)

p_ssc


```

### Discharge

```{r}
# Paths
av_year_path  <- here("data","discharge_av_year_z.csv")
compiled_path <- here("data", "compiled_dis24.csv")

dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

# Read data
compiled        <- read_csv(compiled_path, show_col_types = FALSE)
av_year_updated <- read_csv(av_year_path, show_col_types = FALSE)

# Parse datetime
compiled <- compiled %>%
  mutate(valid_time = as.POSIXct(valid_time, tz = "UTC"))

xmin <- as.POSIXct("1983-01-01", tz = "UTC")
xmax <- as.POSIXct("2020-12-31", tz = "UTC")

# -------------------------
# CHOOSE ONE SITE HERE
# -------------------------
site <- "MALAWI1.2014"

site_df <- compiled %>%
  filter(site_name == site) %>%
  arrange(valid_time)

stopifnot(nrow(site_df) > 0)

# ---- Avulsion shading window (if column exists + has a value) ----
avs <- if ("avulsion_start_date" %in% names(site_df)) {
  site_df %>%
    pull(avulsion_start_date) %>%
    as.numeric() %>%
    na.omit() %>%
    unique()
} else {
  numeric(0)
}

shade_df <- NULL
if (length(avs) > 0) {
  shade_df <- tibble(
    xmin = as.POSIXct(paste0(avs[1] - 1, "-01-01"), tz = "UTC"),
    xmax = as.POSIXct(paste0(avs[1], "-12-31"), tz = "UTC"),
    by = "5 years",
    ymin = -Inf,
    ymax = Inf
  )
}

# ---- Mean 95th percentile line for this site (from zscore table) ----
year_p95 <- av_year_updated %>%
  filter(site_name == site) %>%
  pull(p95_mean) %>%
  first()

# ---- 95th percentile point within shaded window (x = date of max in window) ----
p95_pt <- NULL

if (!is.null(shade_df)) {
  window_df <- site_df %>%
    filter(valid_time >= shade_df$xmin, valid_time <= shade_df$xmax) %>%
    filter(!is.na(dis24))

  if (nrow(window_df) > 0) {
    # window-specific 95th percentile value
    p95_val <- as.numeric(quantile(window_df$dis24, probs = 0.95, na.rm = TRUE, type = 7))

    # exact date of maximum discharge within the window (tie-break: earliest)
    max_row <- window_df %>%
      filter(dis24 == max(dis24, na.rm = TRUE)) %>%
      arrange(valid_time) %>%
      slice(1)

    # plot p95 value at the max date
    p95_pt <- tibble(
      valid_time = max_row$valid_time,
      dis24      = p95_val
    )
  }
}

# ---- Plot ----
p_dis <- ggplot(site_df, aes(x = valid_time, y = dis24)) +

  # Avulsion shading
  {if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
    # Mean p95 line
  {if (!is.na(year_p95))
    geom_hline(
      yintercept = year_p95,
      color = "black",
      linetype = "dashed",
      linewidth = 0.75,
      alpha = 0.75
    )
  } +

  # Base line
  geom_point(color = "#2f6fa5", size = 0.75) +
  geom_line(color = "#2f6fa5", linewidth = 0.75) +

  # Blue point: y = p95(window), x = date of max(window)
  { if (!is.null(p95_pt))
  geom_point(
    data = p95_pt,
    aes(x = valid_time, y = dis24),
    inherit.aes = FALSE,
    shape = 21,        # hollow circle
    size = 5,
    stroke = 2,      # outline thickness
    fill = NA,
    color = "red",
    alpha = 0.7
  )
} +

  scale_y_continuous(expand = c(0, 0)) +

  # Date ticks + limits (add expand here; drop scale_x_continuous to avoid scale conflicts)
  scale_x_datetime(
    limits = c(xmin, xmax),
    breaks = seq(xmin, xmax, by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(
    x = "Year",
    y = "Discharge (m³/s)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.20, color = "grey90"),
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "white", colour = NA)
  )

# ---- Save (one file) ----
safe <- site %>%
  as.character() %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("/", "_")

ggsave(
  filename = file.path(plots_dir, "fig2_dis24_timeseries.svg"),
  plot = p_dis,
  width = 9,
  device = svglite,
  height = 3,
  dpi = 300,
  bg = "white"
)

# Print the plot in RStudio/Quarto
p_dis



```

### Creating Time Series of Discharge and Sediment Z-Scores for Each Year

```{r}
#Load One Site Discharge and Sediment Z-Score
malawi_dis_zscores <- read_csv(here("data", "discharge_zscores_one_site.csv"))
malawi_ssc_zscores <- read_csv(here("data", "ssc_zscores_one_site.csv"))
```

```{r}

# Optional: select one site if present
# site <- "MALAWI1.2014"
# df_red  <- df_red  %>% filter(site_name == site)
# df_blue <- df_blue %>% filter(site_name == site)

# -------------------------
# Clean + harmonize
# -------------------------
prep_df <- function(df, label) {
  df %>%
    mutate(
      year = as.integer(year),
      zlog_p95_q = as.numeric(zlog_p95_q),
      dataset = label,
      date = as.Date(paste0(year, "-01-01"))
    ) %>%
    filter(!is.na(year), !is.na(zlog_p95_q))
}

dis_red  <- prep_df(malawi_dis_zscores,  "Discharge (m^3/s)")
ssc_blue <- prep_df(malawi_ssc_zscores, "SSC (mg/L)")

df_all <- bind_rows(dis_red, ssc_blue)

stopifnot(nrow(df_all) > 0)

# ---- Shading window: 2013 through end of 2014 ----
shade_df <- tibble(
  xmin = as.Date("2013-01-01"),
  xmax = as.Date("2014-12-31"),
  ymin = -Inf,
  ymax = Inf
)
# -------------------------
# Marker year
# -------------------------
mark_year_dis <- 2014L
mark_year_ssc <- 2013L

mark_pts_dis <- df_all %>%
  filter(dataset == "Discharge (m^3/s)", year == mark_year_dis)

mark_pts_ssc <- df_all %>%
  filter(dataset == "SSC (mg/L)", year == mark_year_ssc)

# -------------------------
# X limits
# -------------------------
xmin <- as.Date(paste0(min(df_all$year), "-01-01"))
xmax <- as.Date(paste0(max(df_all$year), "-12-31"))

# -------------------------
# Plot
# -------------------------
p_z <- ggplot(df_all, aes(x = date, y = zlog_p95_q, color = dataset)) +
  
  { if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +

  { if (nrow(mark_pts_dis) > 0)
    geom_point(
      data = mark_pts_dis,
      aes(x = date, y = zlog_p95_q),
      inherit.aes = FALSE,
      shape = 21,
      size = 5,
      stroke = 2,
      fill = NA,
      color = "red",
      alpha = 0.8
    )
  } +
  
  { if (nrow(mark_pts_ssc) > 0)
    geom_point(
      data = mark_pts_ssc,
      aes(x = date, y = zlog_p95_q),
      inherit.aes = FALSE,
      shape = 21,
      size = 5,
      stroke = 2,
      fill = NA,
      color = "red",
      alpha = 0.8
    )
  } +

  scale_color_manual(
  values = c(
    "Discharge (m^3/s)" = "#2f6fa5",  # deep red
    "SSC (mg/L)"        = "#9bbf30"   # deep blue
  )
) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE)) +

  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +

  scale_x_date(
    limits = c(as.Date("1983-01-01"), as.Date("2020-12-31")),
    breaks = seq(as.Date("1983-01-01"), as.Date("2020-01-01"), by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(x = "Year", y = "Z-Score (-)") +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),

    axis.title.y = element_text(size = 20, face = "bold", color = "grey30"),
    axis.title.x = element_text(size = 20, face = "bold", color = "grey30"),

    axis.text = element_text(size = 16, face = "bold", color = "grey30"),
    axis.line = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks = element_line(color = "grey30", linewidth = 1.2),

    legend.position = "none",
    legend.justification = c(0.5, 1),
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 14, face = "bold", color = "grey30"),
    legend.background = element_rect(fill = alpha("white", 0.7), colour = NA),
    legend.key = element_rect(fill = NA, colour = NA),

    plot.background = element_rect(fill = "white", colour = NA)
  )
# -------------------------
# Save
# -------------------------
ggsave(
  filename = file.path(plots_dir, "fig2_log_zscore.svg"),
  plot = p_z,
  width = 9,
  height = 3.1,
  device = svglite,
  dpi = 300,
  bg = "white"
)

p_z

```

## Figure 3 - Boxplots of Discharge Z-Score

### Base Plot

```{r}
# ── COLOURS ───────────────────────────────────────────────────────────────
av_type_colors = 
combined_colors <- c("All Sites" = "grey40", av_type_colors)

# ── T-TESTS ───────────────────────────────────────────────────────────────
fan   <- join_sed |> filter(av_type == "fan")
delta <- join_sed |> filter(av_type == "delta")
il    <- join_sed |> filter(av_type == "intra-lobe")
retro <- join_sed |> filter(av_type == "retrogradational")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar <- function(p) {
  if (p < 0.001)     "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else               "ns"
}

# ── HELPER: BUILD PLOT DATA + SIG LABELS FOR ONE METRIC ──────────────────
make_panel <- function(data, metric_col, y_label, exclude_types = NULL) {

  # Optionally filter out avulsion types
  if (!is.null(exclude_types)) {
    data <- data |> filter(!av_type %in% exclude_types)
  }

  all_sites <- data |> mutate(av_type_plot = "All Sites")
  by_type   <- data |> mutate(av_type_plot = as.character(av_type))

  # Build factor levels dynamically based on remaining av_types
  remaining_types <- unique(as.character(data$av_type))
  all_levels <- c("All Sites", intersect(c("delta", "fan", "intra-lobe", "retrogradational"), remaining_types))

  plot_data <- bind_rows(all_sites, by_type) |>
    mutate(av_type_plot = factor(av_type_plot, levels = all_levels))

  # T-tests — only for types present in data
  t_all    <- t.test(data[[metric_col]], mu = 0)
  t_by_type <- lapply(remaining_types, function(typ) {
    t.test(data |> filter(av_type == typ) |> pull(metric_col), mu = 0)
  })
  names(t_by_type) <- remaining_types

  # Significance labels
  sig_labels <- data.frame(
    av_type_plot = factor(all_levels, levels = all_levels),
    label = c(
      pstar(t_all$p.value),
      sapply(remaining_types, function(typ) pstar(t_by_type[[typ]]$p.value))
    )
  )

  y_positions <- plot_data |>
    group_by(av_type_plot) |>
    summarise(y_pos = max(.data[[metric_col]], na.rm = TRUE) + 0.2)

  sig_labels <- sig_labels |> left_join(y_positions, by = "av_type_plot")

  # Colors — built dynamically
  base_colors <- c("delta" = plasma(4)[1], "fan" = plasma(4)[2],
                   "intra-lobe" = plasma(4)[3], "retrogradational" = plasma(4)[4])
  combined_colors <- c("All Sites" = "grey40", base_colors[remaining_types])

  ggplot(plot_data, aes(x = av_type_plot, y = .data[[metric_col]],
                         fill = av_type_plot, color = av_type_plot)) +

    geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
    geom_boxplot(show.legend = FALSE, alpha = 0.4,
                 outlier.shape = NA, color = "grey30") +
    stat_summary(fun = mean, geom = "crossbar", width = 0.5,
                 color = scales::alpha("red", 0.50), linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed",
               color = "grey30", linewidth = 1.2) +
    geom_vline(xintercept = 1.5, linetype = "dotted",
               color = "grey50", linewidth = 0.8) +
    geom_text(data = sig_labels,
              aes(x = av_type_plot, y = y_pos, label = label),
              inherit.aes = FALSE, color = "grey30", size = 6, fontface = "bold") +
    scale_fill_manual(values = combined_colors) +
    scale_color_manual(values = combined_colors) +
    scale_x_discrete(labels = c(
      "All Sites"        = "All Sites",
      "delta"            = "Delta",
      "fan"              = "Fan",
      "intra-lobe"       = "Intra-lobe",
      "retrogradational" = "Retro."
    )) +
    labs(y = y_label) +
    theme_classic(base_size = 18) +
    theme(
      axis.title.y    = element_text(color = "grey30", size = 20, face = "bold"),
      axis.title.x    = element_blank(),
      axis.line.x     = element_line(color = "grey30", linewidth = 1.2),
      axis.line.y     = element_line(color = "grey30", linewidth = 1.2),
      axis.text.x     = element_text(color = "grey30", size = 16, face = "bold"),
      axis.text.y     = element_text(color = "grey30", size = 16, face = "bold"),
      axis.ticks.y    = element_line(color = "grey30", linewidth = 1.2),
      axis.ticks.x    = element_line(color = "grey30", linewidth = 1.2),
      legend.position = "none",
      plot.background = element_rect(fill = "white", colour = NA)
    )
}

```

### Call Panels

```{r}

# ── BUILD ALL PANELS ──────────────────────────────────────────────────────
p_p95_ssc   <- make_panel(join_sed, "zlog_p95_ssc",   "P95 SSC Z (-)",        exclude_types = "retrogradational")
p_mean_ssc  <- make_panel(join_sed, "zlog_mean_ssc",  "Mean SSC Z (-)",       exclude_types = "retrogradational")
p_p95_q     <- make_panel(join_dis, "zlog_p95_q",     "P95 Discharge Z (-)")
p_mean_q    <- make_panel(join_dis, "zlog_mean_q",    "Mean Discharge Z (-)")
p_flood_dur <- make_panel(join_dis, "zlog_flood_dur", "Flood Duration Z (-)")

# ── SAVE SEPARATELY ───────────────────────────────────────────────────────
plots <- list(
  "fig_p95_ssc"   = p_p95_ssc,
  "fig_mean_ssc"  = p_mean_ssc,
  "fig_p95_q"     = p_p95_q,
  "fig_mean_q"    = p_mean_q,
  "fig_flood_dur" = p_flood_dur
)

#Save Panels
for (name in names(plots)) {
  ggsave(
    filename = here("plots_svg", paste0(name, ".svg")),
    plot     = plots[[name]],
    bg       = "white",
    device   = svglite,
    height   = 4, width = 6, units = "in"
  )
}

```

### Log SSC Data

```{r}

# ── RESHAPE DATA ──────────────────────────────────────────────────────────
plot_data <- join_sed |>
  select(zlog_p95_ssc, zlog_mean_ssc) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    variable = case_when(
      variable == "zlog_p95_ssc"  ~ "Peak",
      variable == "zlog_mean_ssc" ~ "Mean"
    ),
    variable = factor(variable, levels = c("Peak", "Mean"))
  )

# ── T-TESTS ───────────────────────────────────────────────────────────────
t_p95  <- t.test(join_sed$zlog_p95_ssc,  mu = 0)
t_mean <- t.test(join_sed$zlog_mean_ssc, mu = 0)

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar <- function(p) {
  if (p < 0.001)     "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else               "ns"
}

# ── SIGNIFICANCE LABEL DATAFRAME ─────────────────────────────────────────
y_positions <- plot_data |>
  group_by(variable) |>
  summarise(y_pos = max(value, na.rm = TRUE) + 0.2)

sig_labels <- data.frame(
  variable = factor(
    c("Peak", "Mean"),
    levels = c("Peak", "Mean")
  ),
  label = c(
    pstar(t_p95$p.value),
    pstar(t_mean$p.value)
  )
) |>
  left_join(y_positions, by = "variable")

# ── PLOT ──────────────────────────────────────────────────────────────────
ssc_overall <- plot_data |>
  ggplot(aes(x = variable, y = value, fill = variable, color = variable)) +
  # raw data
  geom_jitter(
    width = 0.15,
    alpha = 0.6,
    size = 2
  ) +
  # boxplot
  geom_boxplot(
    width = 0.35,
    alpha = 0.6,
    outlier.shape = NA,
    color = "grey30"  # box outline hardcoded, not from scale
  ) +
  # zero reference
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray30",
    linewidth = 1.2
  ) +
  # mean (solid red line)
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.35,
    color = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +
  # significance stars
  geom_text(
    data = sig_labels,
    aes(x = variable, y = y_pos, label = label),
    inherit.aes = FALSE,
    color = "grey30",
    size = 6,
    fontface = "bold"
  ) +
  scale_fill_manual(values  = c("Peak" = "grey70", "Mean" = "grey40")) +
  scale_color_manual(values = c("Peak" = "grey70", "Mean" = "grey40")) +
  labs(
    x = NULL,
    y = "SSC Z-Score (-)"
  ) +
  theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 20, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "white", colour = NA)
  )

ssc_overall

ggsave(
  filename = here("plots_svg", "fig3_ssc.svg"),
  bg = "white",
  plot = ssc_overall,
  device = svglite,
  height = 4, width = 6, units = "in"
)

```

## Figure 4 - Leadup to Avulsoins

#### Log P95 Annual Discharge

```{r}

plot_data <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_p95_q = mean(zlog_p95_q, na.rm = TRUE),
    se_zlog_p95_q   = sd(zlog_p95_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_p95_q))),
    n               = sum(!is.na(zlog_p95_q)),
    .groups = "drop"
  ) %>%
  mutate(
    # 95% CI using t-distribution (more accurate for small n than 1.96)
    ci_mult  = qt(0.975, df = pmax(n - 1, 1)),
    ci_lower = mean_zlog_p95_q - ci_mult * se_zlog_p95_q,
    ci_upper = mean_zlog_p95_q + ci_mult * se_zlog_p95_q
  )

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

p95_line <- ggplot(plot_data, aes(x = years_before, y = mean_zlog_p95_q,
                      color = av_type, fill = av_type, shape = av_type)) +

  # Confidence interval ribbon (very transparent, fill-matched)
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = av_type),
              color = NA, alpha = 0.12) +

  geom_line(linewidth = 2, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 1.2) +
  geom_errorbar(aes(ymin = mean_zlog_p95_q - se_zlog_p95_q,
                  ymax = mean_zlog_p95_q + se_zlog_p95_q),
              width = 0.15, linewidth = 0.6, alpha = 0.5) +
  geom_point(size = 5, stroke = 1, color = "white", alpha = 0.7) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Peak Discharge Z (-)",
    color = "Avulsion Type",
    fill  = "Avulsion Type",
    shape = "Avulsion Type"
  ) +
  theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 15, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 15, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA),
    plot.margin     = margin(5, 8, 5, 8)
  )

ggsave(
  filename = file.path(plots_dir, "fig4_p95_line.svg"),
  plot     = p95_line,
  width    = 3,
  height   = 5,
  dpi      = 300,
  device = svglite,
  bg       = "transparent"
)
p95_line

```

#### Log Mean Annual Discharge

```{r}

plot_data1 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_mean_q = mean(zlog_mean_q, na.rm = TRUE),
    se_zlog_mean_q   = sd(zlog_mean_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_mean_q))),
    .groups = "drop"
  )

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

mean_line <- ggplot(plot_data1, aes(x = years_before, y = mean_zlog_mean_q,
                      color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = 2, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 1.2) +
  geom_errorbar(aes(ymin = mean_zlog_mean_q - se_zlog_mean_q,
                  ymax = mean_zlog_mean_q + se_zlog_mean_q),
              width = 0.15, linewidth = 0.6, alpha = 0.5) +
  geom_point(size = 5, stroke = 1, color = "white", alpha = 0.7) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Mean Discharge Z (-)",
    color = "Avulsion Type",
    fill  = "Avulsion Type",
    shape = "Avulsion Type"
  ) +
  theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 15, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 15, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA),
    plot.margin     = margin(5, 8, 5, 8)
  )

mean_line

ggsave(
  filename = file.path(plots_dir, "fig4_mean_line.svg"),
  plot     = mean_line,
  width    = 3,
  height   = 5,
  dpi      = 300,
  device = svglite,
  bg       = "transparent"
)

```

#### Log Mean Annual Discharge

```{r}

plot_data2 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_flood_dur = mean(zlog_flood_dur, na.rm = TRUE),
    se_zlog_flood_dur   = sd(zlog_flood_dur, na.rm = TRUE) / sqrt(sum(!is.na(zlog_flood_dur))),
    .groups = "drop"
  )

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

flooddur_line <- ggplot(plot_data2, aes(x = years_before, y = mean_zlog_flood_dur,
                      color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = 2, alpha = 0.7                                          ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 1.2) +
  geom_errorbar(aes(ymin = mean_zlog_flood_dur - se_zlog_flood_dur,
                  ymax = mean_zlog_flood_dur + se_zlog_flood_dur),
              width = 0.15, linewidth = 0.6, alpha = 0.6) +
  geom_point(size = 5, stroke = 1, color = "white", alpha = 0.7) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Flood Duration Z (-)",
    color = "Avulsion Type",
    fill  = "Avulsion Type",
    shape = "Avulsion Type"
  ) +
  theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 15, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 15, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA),
    plot.margin     = margin(5, 8, 5, 8)
  )

flooddur_line

ggsave(
  filename = file.path(plots_dir, "fig4_flooddur.svg"),
  plot     = flooddur_line,
  width    = 3,
  height   = 5,
  device = svglite,
  dpi      = 300,
  bg       = "transparent"
)
```

### Combined Leadup

```{r}

plots_dir = here("plots_svg")

library(gridExtra)
library(grid)
library(svglite)

legend_source <- ggplot(plot_data, aes(x = years_before, y = mean_zlog_p95_q,
                                       color = av_type, fill = av_type, shape = av_type, alpha = 0.7)) +
  geom_point(size = 5, alpha = 0.7) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  guides(color = guide_legend(override.aes = list(size = 5), nrow = 1)) +
  labs(color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type") +
  theme_classic(base_size = 18) +
  theme(
    legend.position      = "bottom",
    legend.justification = "center",
    legend.text          = element_text(color = "grey30", size = 15, face = "bold"),
    legend.title         = element_text(color = "grey30", size = 15, face = "bold"),
    legend.key.size      = unit(1.5, "lines")
  )

g <- ggplotGrob(legend_source)
legend_grob <- g$grobs[[which(sapply(g$grobs, function(x) x$name) == "guide-box")]]

# Build the grob ONCE
combined_grob <- arrangeGrob(
  arrangeGrob(p95_line, mean_line, flooddur_line, ncol = 3),
  legend_grob,
  nrow = 2,
  heights = c(5, 1)
)

# Save to file
svglite(file.path(plots_dir, "fig_4_combined.svg"), width = 10, height = 6, bg = "transparent")
grid.newpage()
grid.draw(combined_grob)
dev.off()

# Display in RMarkdown
grid.newpage()
grid.draw(combined_grob)
```

## Figure 5 - Avulsion Count

## GRL Theme

```{r}

grl_theme <- theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30", margin = margin(r = 6)),
    axis.title.x  = element_text(face = "bold", color = "grey30", margin = margin(t = 6)),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 0.8),
    axis.ticks    = element_line(color = "grey30", linewidth = 0.8),
    legend.position  = "none",
    plot.background  = element_rect(fill = "transparent", colour = NA),
    plot.margin      = margin(t = 6, r = 12, b = 6, l = 8)
  )


```

### Avulsion Count

```{r}

# -------------------------
# Prep avulsion counts
# -------------------------
avulsion_counts <- join_dis %>%
  filter(!is.na(avulsion_year), !is.na(site_area)) %>%
  mutate(year = as.integer(avulsion_year)) %>%
  filter(site_area %in% c("Himalaya", "Pacific Islands", "South America")) %>%
  group_by(site_area, year) %>%
  summarise(n = n(), .groups = "drop")

all_sites <- join_dis %>%
  filter(!is.na(avulsion_year)) %>%
  mutate(year = as.integer(avulsion_year)) %>%
  group_by(year) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(site_area = "All Sites")

# -------------------------
# Poisson test per year
# -------------------------
# Total number of years observed
n_years  <- nrow(all_sites)
# Mean annual count = background Poisson rate
lambda   <- mean(all_sites$n)
mean_all <- lambda

# Run poisson.test() for each year's count against the background rate
# T = n_years so the expected value is lambda * 1 (per year)
poisson_results <- all_sites %>%
  rowwise() %>%
  mutate(
    p_value = poisson.test(n, T = 1, r = lambda)$p.value,
    sig     = p_value < 0.05,
    sig_label = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) %>%
  ungroup() %>%
  filter(sig)

# ---- Combine for plotting ----
df_plot <- bind_rows(avulsion_counts, all_sites)

site_areas <- c("South America", "Himalaya", "Pacific Islands")
all_levels <- c(site_areas, "All Sites")
df_plot    <- df_plot %>% mutate(site_area = factor(site_area, levels = all_levels))

n_sites <- length(site_areas)

ltype_vals <- setNames(rep("solid", length(all_levels)), all_levels)
lsize_vals <- setNames(c(rep(0.8, n_sites), 1.6), all_levels)
color_vals <- setNames(c("#5ec962", "#21918c", "#3b528b", "black"), all_levels)

# Some Scaling Things
# B/C were designed at width 4.5, base_size 16 — that's your reference "1x"
scale_factor <- 14 / 4.5   # ≈ 3.11

base_sz <- 16 * scale_factor          # ≈ 49.8 → text
line_sz <- 1.2 * scale_factor         # ≈ 3.73 → axis lines, ticks
grid_sz <- 0.2 * scale_factor         # ≈ 0.62 → gridlines
col_sz  <- 0.2 * scale_factor         # ≈ 0.62 → geom_col white borders
margin_r <- 8  * scale_factor         # ≈ 24.9
margin_t <- 4  * scale_factor         # ≈ 12.4
key_sz  <- 1.5 * scale_factor         # ≈ 4.67 lines

```

### Avulsion Count V2

```{r}

# -------------------------
# Plot prep
# -------------------------
df_stack <- df_plot %>%
  filter(site_area != "All Sites") %>%
  mutate(site_area = droplevels(factor(site_area)))

site_levels <- sort(unique(as.character(df_stack$site_area)))

grey_vals <- setNames(
  grey.colors(length(site_levels), start = 0.75, end = 0.15),
  site_levels
)

df_stack <- df_stack %>%
  mutate(site_area = factor(site_area, levels = site_levels))

# Asterisks above the stacked total again
df_totals <- df_stack %>%
  group_by(year) %>%
  summarise(n = sum(n), .groups = "drop")

poisson_results <- poisson_results %>%
  select(year, sig_label) %>%
  left_join(df_totals, by = "year")

# -------------------------
# Plot
# -------------------------
p_av <- ggplot(
  df_stack,
  aes(x = year, y = n, fill = site_area, group = site_area)
) +
  geom_col(width = 0.8, color = "white", linewidth = col_sz) +
  scale_fill_manual(values = grey_vals, name = NULL) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  scale_x_continuous(
    breaks = seq(1984, 2018, by = 4),
    limits = c(1983, 2019),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.0, 0.3))) +
  labs(x = "Year", y = "Avulsion\nCount") +
  theme_minimal(base_size = base_sz) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = grid_sz, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30", margin = margin(r = margin_r)),
    axis.title.x  = element_text(face = "bold", color = "grey30", margin = margin(t = margin_t)),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = line_sz),
    axis.ticks    = element_line(color = "grey30", linewidth = line_sz),
    plot.margin   = margin(t = 6 * scale_factor, r = 10 * scale_factor,
                            b = 6 * scale_factor, l = 12 * scale_factor),
    legend.position       = c(0.85, 0.75),
    legend.justification  = c(1, 1),
    legend.background     = element_rect(fill = alpha("white", 0.7), color = NA),
    legend.text           = element_text(color = "grey30", size = base_sz * (15/16), face = "bold"),
    legend.title          = element_text(color = "grey30", size = base_sz * (15/16), face = "bold"),
    legend.key.size       = unit(key_sz, "lines"),
    legend.margin         = margin(2 * scale_factor, 4 * scale_factor, 2 * scale_factor, 4 * scale_factor),
    plot.background       = element_rect(fill = "white", colour = NA)
  )

ggsave(
  filename = file.path(plots_dir, "fig5_avulsion_counts_by_site_v2.svg"),
  plot     = p_av,
  width    = 14,
  height   = 5,
  dpi      = 300,
  device = svglite,
  bg       = "white"
)
p_av

```

### Plotting Avulsion Count

```{r}
# -------------------------
# Plot
# -------------------------
p_av <- ggplot(
  df_plot,
  aes(
    x         = year,
    y         = n,
    color     = site_area,
    linetype  = site_area,
    linewidth = site_area,
    group     = site_area
  )
) +
  geom_line(linewidth = 0.8) +  # single linewidth for all regional lines
  
  geom_line(data = df_plot %>% filter(site_area == "All Sites"),
            aes(x = year, y = n, color = site_area, group = site_area),
            linewidth = 1.6, alpha = 0.5) +

  # Mean line
  geom_hline(
    yintercept = mean_all,
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.8,
    alpha      = 0.6
  ) +

  # Asterisks above significant peaks on the All Sites line
  geom_text(
    data        = poisson_results,
    aes(x = year, y = n, label = sig_label),
    inherit.aes = FALSE,
    vjust       = -0.5,
    size        = 10,
    fontface    = "bold",
    color       = "black"
  ) +

  scale_color_manual(values = color_vals, name = NULL) +
  scale_linetype_manual(values = ltype_vals, name = NULL) +
  guides(
    color    = guide_legend(nrow = 1, byrow = TRUE),
    linetype = guide_legend(nrow = 1, byrow = TRUE)
  ) +
  
  scale_x_continuous(
    breaks = seq(1984, 2020, by = 4),
    limits = c(1983, 2021),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.0, 0.3))) +

  labs(x = "Year", y = "Avulsion\nCount") +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30",
                                 margin = margin(r = 8)),   # pushes label away from edge
    axis.title.x  = element_text(face = "bold", color = "grey30",
                                 margin = margin(t = 4)),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks    = element_line(color = "grey30", linewidth = 1.2),
    plot.margin   = margin(t = 6, r = 10, b = 6, l = 12),
    legend.position   = "bottom",
    legend.justification = "center",
    legend.text          = element_text(color = "grey30", size = 15, face = "bold"),
    legend.title         = element_text(color = "grey30", size = 15, face = "bold"),
    legend.key.size      = unit(1.5, "lines"),
    plot.background   = element_rect(fill = "white", colour = NA)
  )

# -------------------------
# Save
# -------------------------
ggsave(
  filename = file.path(plots_dir, "fig5_avulsion_counts_by_site.svg"),
  plot     = p_av,
  width    = 9,
  height   = 5,
  dpi      = 300,
  device = svglite, 
  bg       = "white"
)
p_av
```

```{r}

# Check what's actually in df_plot
df_plot %>% 
  group_by(site_area) %>% 
  summarise(n_rows = n(), min_year = min(year), max_year = max(year))

# Check factor levels are correct
levels(df_plot$site_area)

# Check color/linetype vectors are named correctly
color_vals
ltype_vals


# Check the actual unique values in your data
join_dis %>% 
  distinct(site_area)
```

### Superelevation Plot

```{r}

# Loading in Alpha Beta Values from Nature Paper
alpha_beta <- read.csv(here("data", "nature_alpha_beta.csv")) |>
  clean_names() |>
  dplyr::select(avulsion_name, geomorphology, gamma, a, b, beta) |>
  rename(site_name = avulsion_name) |>
  mutate(site_name = str_replace_all(site_name, "_", "."))


# Joining Nature Data Paper to Avulsion Data
alpha_beta_av <- inner_join(join_dis, alpha_beta, by = c("site_name"))

corr_beta = cor.test(alpha_beta_av$zlog_mean_q, alpha_beta_av$beta)
print(corr_beta)
corr_gamma = cor.test(alpha_beta_av$zlog_mean_q, alpha_beta_av$gamma)
print(corr_gamma)

```

```{r}

beta_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = beta, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
    geom_point(size = 3, stroke = 0.1, color = "white") +
  labs(
    x = "Mean Discharge log Z-Score (-)",
    y = "Beta (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30"),
    axis.title.x  = element_text(face = "bold", color = "grey30"),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks    = element_line(color = "grey30", linewidth = 1.2),
    legend.position   = "none",
    legend.direction  = "horizontal",
    plot.background   = element_rect(fill = "white", colour = NA)
  )

ggsave(
  filename = file.path(plots_dir, "fig5_beta.svg"),
  plot     = beta_dis,
  width    = 4.5,
  height   = 3,
  dpi      = 300,
   
  bg       = "white"
)

beta_dis
```

### Slope Advantage Plot

```{r}

gamma_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = gamma, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  geom_point(size = 3, stroke = 0.1, color = "white") +
  labs(
    x = "Mean Discharge log Z-Score (-)",
    y = "Gamma (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30"),
    axis.title.x  = element_text( face = "bold", color = "grey30"),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks    = element_line(color = "grey30", linewidth = 1.2),
    legend.position   = "none",
    legend.direction  = "horizontal",
    plot.background   = element_rect(fill = "white", colour = NA)
  )

ggsave(
  filename = file.path(plots_dir, "fig5_gamma.svg"),
  plot     = gamma_dis,
  width    = 4.5,
  height   = 3,
  dpi      = 300,
  device   = svglite, 
  bg       = "white"
)

gamma_dis

```

```{r}

library(patchwork)

combined <- p_av / (beta_dis + gamma_dis)
combined

ggsave(
  filename = file.path(plots_dir, "fig5_combined.png"),
  plot     = combined,
  width    = 19,
  height   = 10,
  units    = "cm",
  dpi      = 300,
  bg       = "transparent"
)
```

## Figure 5 - All in One

```{r}

# =========================================================
# Figure 5 — GRL-compliant export
# Target: 170 mm wide, 6 pt text at final size (per AGU/GRL spec)
# Panel A: full width, 170 x 55 mm
# Panels B & C: side by side, 82 x 70 mm each (6 mm gutter)
# =========================================================

library(ggplot2)
library(dplyr)
library(svglite)

# ---- helper: mm -> inches for ggsave ----
mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL theme settings ----
# NOTE: GRL_BASE_SIZE must stay at 6pt to satisfy the AGU/GRL text-size
# requirement at final print size. Line widths, point sizes, and gridlines
# are NOT covered by that rule and can be kept close to original/bold
# values for visual punch without violating the spec.
GRL_BASE_SIZE   <- 12      # pt, per AGU/GRL spec -- do not increase
GRL_LINEWIDTH   <- 1.2    # axis/tick lines -- back to original weight
GRL_GRIDWIDTH   <- 0.3    # gridlines, slightly lighter than axis lines
GRL_POINT_SIZE  <- 3      # scatter point size -- back to original
GRL_SMOOTH_LW   <- 1.5    # trend line width -- back to original

grl_theme <- function(legend = TRUE) {
  theme_minimal(base_size = GRL_BASE_SIZE) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_line(linetype = "solid", linewidth = GRL_GRIDWIDTH, color = "grey90"),
      axis.title.y  = element_text(face = "bold", color = "grey30", size = GRL_BASE_SIZE,
                                    margin = margin(r = 2)),
      axis.title.x  = element_text(face = "bold", color = "grey30", size = GRL_BASE_SIZE,
                                    margin = margin(t = 2)),
      axis.text     = element_text(face = "bold", color = "grey30", size = GRL_BASE_SIZE),
      axis.line     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks    = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      plot.margin   = margin(t = 1, r = 2, b = 1, l = 2),
      legend.text   = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      legend.title  = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      legend.key.size = unit(0.6, "lines"),
      legend.margin   = margin(1, 1, 1, 1),
      legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
      plot.background   = element_rect(fill = "white", colour = NA)
    )
}

# =========================================================
# PANEL A — Avulsion counts by site/year (stacked bars)
# =========================================================

df_stack <- df_plot %>%
  filter(site_area != "All Sites") %>%
  mutate(site_area = droplevels(factor(site_area)))

site_levels <- sort(unique(as.character(df_stack$site_area)))
grey_vals <- setNames(
  grey.colors(length(site_levels), start = 0.75, end = 0.15),
  site_levels
)
df_stack <- df_stack %>%
  mutate(site_area = factor(site_area, levels = site_levels))

df_totals <- df_stack %>%
  group_by(year) %>%
  summarise(n = sum(n), .groups = "drop")

poisson_results <- poisson_results %>%
  select(year, sig_label) %>%
  left_join(df_totals, by = "year")

p_av <- ggplot(
  df_stack,
  aes(x = year, y = n, fill = site_area, group = site_area)
) +
  geom_col(width = 0.8, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = grey_vals, name = NULL) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  scale_x_continuous(
    breaks = seq(1984, 2018, by = 4),
    limits = c(1983, 2019),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.0, 0.3))) +
  labs(x = "Year", y = "Avulsion\nCount") +
  grl_theme() +
  theme(
    legend.position      = c(0.5, 0.85),
    legend.justification = c(0.5, 1)
  )

ggsave(
  filename = file.path(plots_dir, "fig5_avulsion_counts_by_site_v2.svg"),
  plot     = p_av,
  width    = mm_to_in(170),
  height   = mm_to_in(70),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

# =========================================================
# PANELS B & C — Beta / Gamma vs discharge Z-score
# =========================================================

alpha_beta <- read.csv(here("data", "nature_alpha_beta.csv")) |>
  clean_names() |>
  dplyr::select(avulsion_name, geomorphology, gamma, a, b, beta) |>
  rename(site_name = avulsion_name) |>
  mutate(site_name = str_replace_all(site_name, "_", "."))

alpha_beta_av <- inner_join(join_dis, alpha_beta, by = c("site_name"))

corr_beta  <- cor.test(alpha_beta_av$zlog_mean_q, alpha_beta_av$beta)
corr_gamma <- cor.test(alpha_beta_av$zlog_mean_q, alpha_beta_av$gamma)
print(corr_beta)
print(corr_gamma)

# ---- Panel B: Beta ----
beta_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = beta, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = GRL_SMOOTH_LW) +
  geom_point(size = GRL_POINT_SIZE, stroke = 0.1, color = "white") +
  labs(
    x = "Mean Discharge Z (-)",
    y = "Beta (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  grl_theme() +
  theme(
    legend.position  = "none",
    legend.direction = "horizontal"
  )

ggsave(
  filename = file.path(plots_dir, "fig5_beta.svg"),
  plot     = beta_dis,
  width    = mm_to_in(82),
  height   = mm_to_in(55),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

# ---- Panel C: Gamma ----
gamma_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = gamma, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = GRL_SMOOTH_LW) +
  geom_point(size = GRL_POINT_SIZE, stroke = 0.1, color = "white") +
  labs(
    x = "Mean Discharge Z (-)",
    y = "Gamma (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  grl_theme() +
  theme(
    legend.position  = "none",
    legend.direction = "horizontal"
  )

ggsave(
  filename = file.path(plots_dir, "fig5_gamma.svg"),
  plot     = gamma_dis,
  width    = mm_to_in(82),
  height   = mm_to_in(55),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

# =========================================================
# Notes for Affinity assembly:
# - Set Affinity document width to 170 mm (or your true final width).
# - Place each SVG and scale to EXACTLY its native mm size (100%) —
#   do not stretch/scale further, or the 6pt text will drift off-spec.
# - Panel A sits full-width on top (170 x 55 mm).
# - Panels B and C sit side by side below, each 82 x 70 mm,
#   with ~6 mm gutter between them (82 + 82 + 6 = 170).
# - For final submission, export the assembled figure from Affinity
#   as PDF or EPS with fonts outlined/converted to curves (AGU requirement).
# =========================================================


```

## Supplemental Figures:

### Scatter Plot of Discharge and SSC

```{r}

make_scatter <- function(data, x_col, y_col, x_label, y_label, exclude_types = NULL) {

  if (!is.null(exclude_types)) {
    data <- data |> filter(!av_type %in% exclude_types)
  }

  remaining_types <- intersect(c("delta", "fan", "intra-lobe", "retrogradational"),
                                unique(as.character(data$av_type)))
  base_colors <- c("delta"            = plasma(4)[1],
                   "fan"              = plasma(4)[2],
                   "intra-lobe"       = plasma(4)[3],
                   "retrogradational" = plasma(4)[4])
  type_colors <- base_colors[remaining_types]

  data <- data |> mutate(av_type = factor(av_type, levels = remaining_types))

  # ── LINEAR FIT STATS ──────────────────────────────────────────────────
  fit <- lm(data[[y_col]] ~ data[[x_col]], data = data)
  r2  <- summary(fit)$r.squared
  pval <- summary(fit)$coefficients[2, 4]

  # Format p-value
  p_label <- if (pval < 0.001) "p < 0.001" else paste0("p = ", round(pval, 3))
  fit_label <- paste0("R² = ", round(r2, 2), "\n", p_label)

  # Label position — top left of plot
  x_pos <- min(data[[x_col]], na.rm = TRUE)
  y_pos <- max(data[[y_col]], na.rm = TRUE)

  ggplot(data, aes(x = .data[[x_col]], y = .data[[y_col]],
                   color = av_type, fill = av_type)) +

    geom_point(alpha = 0.6, size = 2.5) +

    # Linear fit line with confidence interval
    geom_smooth(method = "lm", se = TRUE, inherit.aes = FALSE,
                aes(x = .data[[x_col]], y = .data[[y_col]]),
                color = "grey30", fill = "grey70",
                linewidth = 0.8, alpha = 0.2) +

    # R² and p-value annotation
    annotate("text", x = x_pos, y = y_pos,
             label = fit_label,
             hjust = 0, vjust = 1,
             color = "grey30", size = 5, fontface = "bold") +

    scale_color_manual(values = type_colors,
                       labels = c("delta"            = "Delta",
                                  "fan"              = "Fan",
                                  "intra-lobe"       = "Intra-lobe",
                                  "retrogradational" = "Retro.")) +
    scale_fill_manual(values = type_colors,
                      labels = c("delta"            = "Delta",
                                 "fan"              = "Fan",
                                 "intra-lobe"       = "Intra-lobe",
                                 "retrogradational" = "Retro.")) +

    labs(x = x_label, y = y_label, color = "Avulsion Type", fill = "Avulsion Type") +

    theme_classic(base_size = 18) +
    theme(
      axis.title.x    = element_text(color = "grey30", size = 20, face = "bold"),
      axis.title.y    = element_text(color = "grey30", size = 20, face = "bold"),
      axis.line.x     = element_line(color = "grey30", linewidth = 1.2),
      axis.line.y     = element_line(color = "grey30", linewidth = 1.2),
      axis.text.x     = element_text(color = "grey30", size = 16, face = "bold"),
      axis.text.y     = element_text(color = "grey30", size = 16, face = "bold"),
      axis.ticks.y    = element_line(color = "grey30", linewidth = 1.2),
      axis.ticks.x    = element_line(color = "grey30", linewidth = 1.2),
      legend.position = "none",
      legend.title    = element_text(color = "grey30", size = 14, face = "bold"),
      legend.text     = element_text(color = "grey30", size = 12),
      plot.background = element_rect(fill = "white", colour = NA)
    )
}


join_all <- join_dis |> 
  inner_join(join_sed, by = c("site_name", "av_type", "site_area", "avulsion_year"))

p_scatter_p95  <- make_scatter(join_all, "zlog_p95_q",  "zlog_p95_ssc",
                                "P95 Discharge Z (-)", "P95 SSC Z (-)")

p_scatter_mean <- make_scatter(join_all, "zlog_mean_q", "zlog_mean_ssc",
                                "Mean Discharge Z (-)", "Mean SSC Z (-)")

ggsave(here("plots_svg", "scatter_p95.svg"),  p_scatter_p95,
       bg = "white", device = svglite, height = 4, width = 6, units = "in")
ggsave(here("plots_svg", "scatter_mean.svg"), p_scatter_mean,
       bg = "white", device = svglite, height = 4, width = 6, units = "in")

```

### World Map of Discharge Z-Score

```{r}
# #Set Up Map Extent
# xmin <- -170
# xmax <- 170
# 
# #Find Standard Deviation for the Dataset
# #Purpose: use for the scale to condense and make a little easier to visualize 
# std_dev_dis <- sd(join_dis$p95_zscore, na.rm = TRUE)
# 
# world_map <- ne_countries(scale = "medium", returnclass = "sf") |>
#   filter(continent != "Antarctica")
# 
# world_map_plot <- ggplot() +
#   geom_sf(data = world_map, fill = "grey80", color = "grey80", linewidth = 0.3) +
#   geom_point(
#     data = join_dis,
#     aes(x = site_lon, y = site_lat, color = p95_zscore),
#     size = 3,
#     alpha = 0.85,
#     shape = 16
#   ) +
#   scale_color_distiller(
#     name = "Discharge Z-Score",
#     palette = "Spectral",
#     limits = c(-0.8, 0.8),
#     oob = scales::squish
#   ) +
#   coord_sf(xlim = c(xmin, xmax), ylim = c(-60, 85), expand = FALSE) +
# 
#   theme_void(base_size = 14) +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background  = element_rect(fill = "white", color = NA),
# 
#     # Legend placement and sizing
#     legend.position = "top",
#     legend.justification = "center",
#     legend.box = "vertical",
# 
#     # Text sizing
#     legend.title = element_text(face = "bold", size = 20),
#     legend.text  = element_text(size = 20),
# 
#     # Spacing
#     legend.margin = margin(t = 12, b = 12),
#     legend.spacing.x = unit(1.5, "cm"),
# 
#     # Give the plot a bit more breathing room
#     plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
#   ) +
#   guides(
#     color = guide_colorbar(
#       title.position = "top",
#       title.hjust = 0.5,
#       barwidth  = unit(0.4, "npc"),   # spans most of plot width
#       barheight = unit(0.4, "cm"),
#       direction = "horizontal"
#     )
#   )
# 
# world_map_plot
# 
# ggsave(
#   filename = here("plots_2.0", "supplemental_dis_map.png"),
#   plot = world_map_plot,
#   width = 16,
#   height = 8,
#   units = "in",
#   dpi = 300
# )
# 

```

### World Map of SSC

```{r}

# #Set Map Extents
# xmin <- -170
# xmax <- 170
# 
# #Find Standard Deviation for the Dataset
# #Purpose: use for the scale to condense and make a little easier to visualize 
# std_dev_sed <- sd(join_sed$p95_zscore, na.rm = TRUE)
# 
# world_map <- ne_countries(
#   scale = "medium", returnclass = "sf") |>
#   filter(continent != "Antarctica")
# 
# world_map_plot <- ggplot() +
#   geom_sf(data = world_map, fill = "grey80", color = "grey80", linewidth = 0.3) +
#   geom_point(
#     data = join_sed,
#     aes(x = site_lon, y = site_lat, color = p95_zscore),
#     size = 3,
#     alpha = 0.85,
#     shape = 16
#   ) +
#   scale_color_distiller(
#     name = "SSC Z-Score",
#     palette = "Spectral",
#     limits = c(-std_dev_sed, std_dev_sed),
#     oob = scales::squish
#   ) +
#   coord_sf(xlim = c(xmin, xmax), ylim = c(-60, 85), expand = FALSE) +
# 
#   theme_void(base_size = 14) +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background  = element_rect(fill = "white", color = NA),
# 
#     # Legend placement and sizing
#     legend.position = "top",
#     legend.justification = "center",
#     legend.box = "vertical",
# 
#     # Text sizing
#     legend.title = element_text(face = "bold", size = 20),
#     legend.text  = element_text(size = 20),
# 
#     # Spacing
#     legend.margin = margin(t = 12, b = 12),
#     legend.spacing.x = unit(1.5, "cm"),
# 
#     # Give the plot a bit more breathing room
#     plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
#   ) +
#   guides(
#     color = guide_colorbar(
#       title.position = "top",
#       title.hjust = 0.5,
#       barwidth  = unit(0.4, "npc"),   # spans most of plot width
#       barheight = unit(0.4, "cm"),
#       direction = "horizontal"
#     )
#   )
# 
# world_map_plot
# 
# ggsave(
#   filename = here("plots_2.0", "supplemental_ssc_map.png"),
#   plot = world_map_plot,
#   width = 16,
#   height = 8,
#   units = "in",
#   dpi = 300
# )
```
