---
title: "Templates for All Maps and Graphs + Descriptive Statistics"
description: "This document is meant to act as the location where all final version graphs are meant to be created, and all descriptive statistics put onto those graphs will be calcualted. This synthesizes material from other documents up to this point. "
author: "Sam Lance"
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
    theme: Lux
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r, include =FALSE}
library(dplyr)
library(tidyverse)
library(lubridate)
library(readr)
library(here)
library(ggplot2)
library(stringr)
library(purrr)
library(maps)
library(knitr)
library(kableExtra)
library(sf)
library(RColorBrewer)
library(patchwork)
library(tmap)
library(viridis)
library(sp)
library(rworldmap)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(devtools)
library(janitor)
library(svglite)
```

## Reading In + Cleaning Data

### Reading in Data

```{r, output=FALSE}}
#DISCHARGE
join_dis <- read_csv(here("data","vg_dis_z.csv")) |>
  inner_join(
    read_csv(here("data","base_sites.csv")) |>
      select(site_name, site_lat, site_lon),
    by = "site_name") |> #og data missing these parts, need to join
  mutate(
    zlog_p95_q  = coalesce(zlog_p95_q, z_p95_q),
    zlog_mean_q = coalesce(zlog_mean_q, z_mean_q)
  ) #coalesce falls back to the second option if the first is not available, doing because issue with intermittent flows at 10 sites 

#Sediment 
join_sed = read_csv(here("data","vg_ssc_z.csv")) |>
  drop_na()

```

### Cleaning Discharge Data

```{r}
#Convert DF to SF Object 
sf_dis <- st_as_sf(join_dis, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_dis, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_dis <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

```

### Cleaning Sediment Data

```{r}

#Convert DF to SF Object 
sf_sed <- st_as_sf(join_sed, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_sed, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_sed <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

```

### Stats for Paper

```{r}

#Discharge + SSC Overall
dis_stats <- join_dis |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE),
    mean_log_durz = mean(zlog_flood_dur, na.rm = TRUE)
  )

ssc_stats <- join_sed |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE))

#Summarize Same Way but for Avulsion Types
dis_stats_type <- join_dis |>
  group_by(av_type) |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE),
    mean_log_durz = mean(zlog_flood_dur, na.rm = TRUE)
  )

ssc_stats_type <- join_sed |>
  group_by(av_type) |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE))
```


### Reading in Full Time Series Data

```{r}

#Load Discharge Data 
discharge_ts <- read_csv(here("data", "vg_dis_z_all_years.csv"))

```

###  Pulling out 5 Years Prior to Avulsion

```{r}
pre_avulsion_5 <- discharge_ts %>%
  filter(year >= (avulsion_year - 5) & year < avulsion_year)

```

## Figure 1 - World Map with Avulsion Types

### Establishing Colors and Shapes for All Maps

```{r}
#Loading in World Map Data
world_map <- ne_countries(scale = "medium", returnclass = "sf")

colors <- c(
  "delta"             = "#0D0887FF",
  "fan"               = "#9C179EFF",
  "intra-lobe"        = "#F89441FF",
  "retrogradational"  = "#F0F921FF"
)

shapes <- c(
  "delta"             = 21,
  "fan"               = 22,
  "intra-lobe"        = 23,
  "retrogradational"  = 24
)

# #devtools::install_github("an-bui/calecopal")# filled shapes
# breaks <- c(-2, -1, 0, 1, 2, 3, 4, 5)
# colors <- brewer.pal(length(breaks) - 1, "RdYlBu")
# rev_colors <- rev(colors)  # Reverse the color order
# shapes <- c(16, 17, 15, 18)  # Different shapes for avulsion types

```

### Robinson Projection World Map

```{r}

library(tidyverse)
library(sf)
library(ggplot2)

# -------------------------
# Inputs assumed to exist:
# world_map (sf polygons, lon/lat)
# sf_dis (dataframe with site_lon, site_lat, site_name, av_type)
# rev_colors (vector of colors)
# shapes (named vector of shapes for av_type)
# -------------------------

# 1) Select sites to highlight
selected_sites <- sf_dis |>
  filter(site_name %in% c("B9", "R3.2", "JA.25.India.1990", "MANAMBOLO.2004"))

# 2) Ensure factor levels consistent
sf_dis <- sf_dis |>
  mutate(av_type = factor(av_type))

# 3) Named color vector for av_type
av_type_colors <- setNames(rev_colors, levels(sf_dis$av_type))

# 4) Robinson CRS
crs_robin <- "+proj=robin +lon_0=0 +datum=WGS84 +units=m +no_defs"

# 5) Make points sf (critical for correct projection)
sf_dis_sf <- st_as_sf(
  sf_dis,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

selected_sites_sf <- st_as_sf(
  selected_sites,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

# 6) Ensure world_map has CRS (if not already)
if (is.na(st_crs(world_map))) st_crs(world_map) <- 4326

# 7) Create a "globe" outline and optional graticule
globe_outline <- st_as_sfc(
  st_bbox(c(xmin = -180, ymin = -90, xmax = 180, ymax = 90), crs = st_crs(4326))
) |>
  st_transform(crs_robin)

# Optional graticule
grat <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = seq(-60, 60, by = 30),
  crs = st_crs(4326)
) |>
  st_transform(crs_robin)

# -------------------------
# NEW: Top + bottom boundary (latitude caps)
# -------------------------
lat_min <- -90
lat_max <- 90
lon_seq <- seq(-180, 180, by = 0.25)

make_parallel <- function(lat) {
  st_sfc(
    st_linestring(cbind(lon_seq, rep(lat, length(lon_seq)))),
    crs = 4326
  ) |>
    st_transform(crs_robin)
}

bottom_boundary <- make_parallel(lat_min)
top_boundary    <- make_parallel(lat_max)

# 8) Build plot (Robinson + globe boundary)
world_map_plot <- ggplot() +

  # Light graticule behind everything (optional; remove if you don’t want lines)
  geom_sf(data = grat, color = "grey70", linewidth = 0.25) +

  # Continents
  geom_sf(
    data = world_map,
    fill = "grey70",
    color = "grey60",
    linewidth = 0.25
  ) +

  # All sites
  geom_sf(
    data = sf_dis_sf,
    aes(fill = av_type, shape = av_type, color = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 0.5
  ) +

  # Selected sites: same fill/shape, red outline
  geom_sf(
    data = selected_sites_sf,
    aes(fill = av_type, shape = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 1.0,
    color = "black",
    show.legend = FALSE
  ) +

  # -------------------------
  # NEW: draw top/bottom boundary lines
  # -------------------------
  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  # Globe outline (draw last so it frames everything)
  geom_sf(
    data = globe_outline,
    fill = NA,
    color = "grey70",
    linewidth = 0.7
  ) +

  # Scales
  scale_color_manual(name = "Avulsion Type", values = colors) +
  scale_fill_manual(name = "Avulsion Type", values = colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +

  # Robinson projection
  coord_sf(crs = crs_robin, expand = FALSE) +

  # Styling
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    
    #legend.position = "none",
    legend.position = c(0.2, 0.25),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    legend.spacing.y = unit(-0.1, "cm"),
    legend.key.height = unit(0.3, "lines"),
    legend.background = element_rect(fill = "white", color = "white")
  ) +
  guides(
    fill  = guide_legend(byrow = TRUE),
    shape = guide_legend(byrow = TRUE)
  )

world_map_plot

ggsave(
  filename = here("plots_svg", "fig1_world_robinson_globe.svg"),
  plot = world_map_plot,
  width = 10,
  height = 7,
  device = svglite, 
  units = "in",
  dpi = 300
)

```

## Figure 2 - Time Series

### Sediment

```{r}

# -------------------------
# Paths
# -------------------------
av_year_updated_path <- here("data", "ssc_av_year_z.csv")
compiled_output_path <- here("data", "time_series.csv")
plots_dir <- "/Volumes/Crucial X6/paper_figures/rstudio/plots_svg"

# -------------------------
# Load data
# -------------------------
av_year_updated <- read_csv(av_year_updated_path, show_col_types = FALSE)
compiled <- read_csv(compiled_output_path, show_col_types = FALSE)

compiled <- compiled %>%
  mutate(date = as.Date(date))


# -------------------------
# SELECT ONE SITE HERE
# -------------------------
site <- "MALAWI1.2014"

site_df <- compiled %>%
  filter(site_name == site) %>%
  arrange(date)

# Check for NAs in your relevant columns
sum(is.na(site_df$y_pred))

# Check if any points fall outside your axis range
df |> filter(y < your_lower_limit | y > your_upper_limit)

stopifnot(nrow(site_df) > 0)

# -------------------------
# Avulsion shading window
# -------------------------
avs <- if ("avulsion_start_date" %in% names(site_df)) {
  site_df %>%
    pull(avulsion_start_date) %>%
    as.numeric() %>%
    na.omit() %>%
    unique()
} else {
  numeric(0)
}

shade_df <- NULL
if (length(avs) > 0) {
  shade_df <- tibble(
    xmin = as.Date(paste0(avs[1] - 1, "-01-01")),
    xmax = as.Date(paste0(avs[1], "-12-31")),
    ymin = -Inf,
    ymax = Inf
  )
}

# -------------------------
# Mean 95th percentile line (from zscore table)
# -------------------------
year_p95 <- av_year_updated %>%
  filter(site_name == site) %>%
  pull(p95_mean) %>%
  first()

# -------------------------
# 95th percentile value within the shaded window,
# plotted at the exact date of the maximum within the window
# -------------------------
p95_pt <- NULL

if (!is.null(shade_df)) {
  window_df <- site_df %>%
    filter(date >= shade_df$xmin, date <= shade_df$xmax) %>%
    filter(!is.na(y_pred))

  if (nrow(window_df) > 0) {
    p95_val <- as.numeric(quantile(window_df$y_pred, probs = 0.95, na.rm = TRUE, type = 7))

    max_row <- window_df %>%
      filter(y_pred == max(y_pred, na.rm = TRUE)) %>%
      arrange(date) %>%   # if ties, use earliest max date
      slice(1)

    p95_pt <- tibble(
      date = max_row$date,
      y_pred = p95_val
    )
  }
}

# -------------------------
# Plot
# -------------------------
p_ssc <- ggplot(site_df, aes(x = date, y = y_pred)) +

  # Avulsion shading
  {if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
    # Mean p95 line
  {if (!is.na(year_p95))
    geom_hline(
      yintercept = year_p95,
      linetype = "dashed",
      color = "black",
      linewidth = 0.75,
      alpha = 0.75
    )
  } +

  # Time series
  geom_point(color = "#9bbf30", size = 0.75) +
  geom_line(color = "#9bbf30", linewidth = 0.75) +

# Red hollow circle: y = p95(window), x = date of max(window)
{ if (!is.null(p95_pt))
  geom_point(
    data = p95_pt,
    aes(x = date, y = y_pred),
    inherit.aes = FALSE,
    shape = 21,        # hollow circle
    size = 5,
    stroke = 2,      # outline thickness
    fill = NA,
    color = "red",
    alpha = 0.7
  )
} +

  # Axes (keep only scale_x_date to avoid scale replacement warnings)
  scale_x_date(
    limits = as.Date(c("1984-01-01", "2020-12-31")),
    breaks = seq(as.Date("1984-01-01"),
                 as.Date("2020-01-01"),
                 by = "4 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = c(0, 0)) +

  labs(
    x = "Year",
    y = "SSC (mg/L)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(
      linetype  = "solid",
      linewidth = 0.25,
      color = "grey90"),
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA)
  )

# -------------------------
# Save
# -------------------------
safe_name <- site %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("/", "_")

ggsave(
  filename = file.path(plots_dir, "fig2_ssc_timeseries.svg"),
  plot = p_ssc,
  width = 9,
  device = svglite,
  height = 3,
  dpi = 300
)

p_ssc


```

### Discharge

```{r}
# Paths
av_year_path  <- here("data","discharge_av_year_z.csv")
compiled_path <- here("data", "compiled_dis24.csv")

dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

# Read data
compiled        <- read_csv(compiled_path, show_col_types = FALSE)
av_year_updated <- read_csv(av_year_path, show_col_types = FALSE)

# Parse datetime
compiled <- compiled %>%
  mutate(valid_time = as.POSIXct(valid_time, tz = "UTC"))

xmin <- as.POSIXct("1984-01-01", tz = "UTC")
xmax <- as.POSIXct("2020-12-31", tz = "UTC")

# -------------------------
# CHOOSE ONE SITE HERE
# -------------------------
site <- "MALAWI1.2014"

site_df <- compiled %>%
  filter(site_name == site) %>%
  arrange(valid_time)

stopifnot(nrow(site_df) > 0)

# ---- Avulsion shading window (if column exists + has a value) ----
avs <- if ("avulsion_start_date" %in% names(site_df)) {
  site_df %>%
    pull(avulsion_start_date) %>%
    as.numeric() %>%
    na.omit() %>%
    unique()
} else {
  numeric(0)
}

shade_df <- NULL
if (length(avs) > 0) {
  shade_df <- tibble(
    xmin = as.POSIXct(paste0(avs[1] - 1, "-01-01"), tz = "UTC"),
    xmax = as.POSIXct(paste0(avs[1], "-12-31"), tz = "UTC"),
    ymin = -Inf,
    ymax = Inf
  )
}

# ---- Mean 95th percentile line for this site (from zscore table) ----
year_p95 <- av_year_updated %>%
  filter(site_name == site) %>%
  pull(p95_mean) %>%
  first()

# ---- 95th percentile point within shaded window (x = date of max in window) ----
p95_pt <- NULL

if (!is.null(shade_df)) {
  window_df <- site_df %>%
    filter(valid_time >= shade_df$xmin, valid_time <= shade_df$xmax) %>%
    filter(!is.na(dis24))

  if (nrow(window_df) > 0) {
    # window-specific 95th percentile value
    p95_val <- as.numeric(quantile(window_df$dis24, probs = 0.95, na.rm = TRUE, type = 7))

    # exact date of maximum discharge within the window (tie-break: earliest)
    max_row <- window_df %>%
      filter(dis24 == max(dis24, na.rm = TRUE)) %>%
      arrange(valid_time) %>%
      slice(1)

    # plot p95 value at the max date
    p95_pt <- tibble(
      valid_time = max_row$valid_time,
      dis24      = p95_val
    )
  }
}

# ---- Plot ----
p_dis <- ggplot(site_df, aes(x = valid_time, y = dis24)) +

  # Avulsion shading
  {if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
    # Mean p95 line
  {if (!is.na(year_p95))
    geom_hline(
      yintercept = year_p95,
      color = "black",
      linetype = "dashed",
      linewidth = 0.75,
      alpha = 0.75
    )
  } +

  # Base line
  geom_point(color = "#2f6fa5", size = 0.75) +
  geom_line(color = "#2f6fa5", linewidth = 0.75) +

  # Blue point: y = p95(window), x = date of max(window)
  { if (!is.null(p95_pt))
  geom_point(
    data = p95_pt,
    aes(x = valid_time, y = dis24),
    inherit.aes = FALSE,
    shape = 21,        # hollow circle
    size = 5,
    stroke = 2,      # outline thickness
    fill = NA,
    color = "red",
    alpha = 0.7
  )
} +

  scale_y_continuous(expand = c(0, 0)) +

  # Date ticks + limits (add expand here; drop scale_x_continuous to avoid scale conflicts)
  scale_x_datetime(
    limits = c(xmin, xmax),
    breaks = seq(xmin, xmax, by = "4 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(
    x = "Year",
    y = "Discharge (m³/s)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.20, color = "grey90"),
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA)
  )

# ---- Save (one file) ----
safe <- site %>%
  as.character() %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("/", "_")

ggsave(
  filename = file.path(plots_dir, "fig2_dis24_timeseries.svg"),
  plot = p_dis,
  width = 9,
  device = svglite,
  height = 3,
  dpi = 300
)

# Print the plot in RStudio/Quarto
p_dis



```

### Creating Time Series of Discharge and Sediment Z-Scores for Each Year

```{r}
#Load One Site Discharge and Sediment Z-Score
malawi_dis_zscores <- read_csv(here("data", "discharge_zscores_one_site.csv"))
malawi_ssc_zscores <- read_csv(here("data", "ssc_zscores_one_site.csv"))
```

```{r}

# Optional: select one site if present
# site <- "MALAWI1.2014"
# df_red  <- df_red  %>% filter(site_name == site)
# df_blue <- df_blue %>% filter(site_name == site)

# -------------------------
# Clean + harmonize
# -------------------------
prep_df <- function(df, label) {
  df %>%
    mutate(
      year = as.integer(year),
      zlog_p95_q = as.numeric(zlog_p95_q),
      dataset = label,
      date = as.Date(paste0(year, "-01-01"))
    ) %>%
    filter(!is.na(year), !is.na(zlog_p95_q))
}

dis_red  <- prep_df(malawi_dis_zscores,  "Discharge (m^3/s)")
ssc_blue <- prep_df(malawi_ssc_zscores, "SSC (mg/L)")

df_all <- bind_rows(dis_red, ssc_blue)

stopifnot(nrow(df_all) > 0)

# ---- Shading window: 2013 through end of 2014 ----
shade_df <- tibble(
  xmin = as.Date("2013-01-01"),
  xmax = as.Date("2014-12-31"),
  ymin = -Inf,
  ymax = Inf
)
# -------------------------
# Marker year
# -------------------------
mark_year_dis <- 2014L
mark_year_ssc <- 2013L

mark_pts_dis <- df_all %>%
  filter(dataset == "Discharge (m^3/s)", year == mark_year_dis)

mark_pts_ssc <- df_all %>%
  filter(dataset == "SSC (mg/L)", year == mark_year_ssc)

# -------------------------
# X limits
# -------------------------
xmin <- as.Date(paste0(min(df_all$year), "-01-01"))
xmax <- as.Date(paste0(max(df_all$year), "-12-31"))

# -------------------------
# Plot
# -------------------------
p_z <- ggplot(df_all, aes(x = date, y = zlog_p95_q, color = dataset)) +
  
  { if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +

  { if (nrow(mark_pts_dis) > 0)
    geom_point(
      data = mark_pts_dis,
      aes(x = date, y = zlog_p95_q),
      inherit.aes = FALSE,
      shape = 21,
      size = 5,
      stroke = 2,
      fill = NA,
      color = "red",
      alpha = 0.8
    )
  } +
  
  { if (nrow(mark_pts_ssc) > 0)
    geom_point(
      data = mark_pts_ssc,
      aes(x = date, y = zlog_p95_q),
      inherit.aes = FALSE,
      shape = 21,
      size = 5,
      stroke = 2,
      fill = NA,
      color = "red",
      alpha = 0.8
    )
  } +

  scale_color_manual(
  values = c(
    "Discharge (m^3/s)" = "#2f6fa5",  # deep red
    "SSC (mg/L)"        = "#9bbf30"   # deep blue
  )
) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE)) +

  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +

  scale_x_date(
    limits = c(as.Date("1985-01-01"), as.Date("2020-12-31")),
    breaks = seq(as.Date("1985-01-01"), as.Date("2020-01-01"), by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(x = "Year", y = "log Z-Score (-)") +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),

    axis.title.y = element_text(size = 20, face = "bold", color = "grey30"),
    axis.title.x = element_text(size = 20, face = "bold", color = "grey30"),

    axis.text = element_text(size = 16, face = "bold", color = "grey30"),
    axis.line = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks = element_line(color = "grey30", linewidth = 1.2),

    legend.position = "none",
    legend.justification = c(0.5, 1),
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 14, face = "bold", color = "grey30"),
    legend.background = element_rect(fill = alpha("white", 0.7), colour = NA),
    legend.key = element_rect(fill = NA, colour = NA),

    plot.background = element_rect(fill = "transparent", colour = NA)
  )
# -------------------------
# Save
# -------------------------
ggsave(
  filename = file.path(plots_dir, "fig2_log_zscore.svg"),
  plot = p_z,
  width = 9,
  height = 3.1,
  device = svglite,
  dpi = 300,
  bg = "transparent"
)

p_z

```

## Figure 3 - Boxplots of Discharge Z-Score

### Log 95th Percentile Discharge

```{r}

# ── COLOURS ───────────────────────────────────────────────────────────────
colors <- plasma(4)
av_type_colors <- setNames(colors, c("delta", "fan", "intra-lobe", "retrogradational"))
print(av_type_colors)
combined_colors <- c("All Sites" = "grey50", av_type_colors)
print(names(combined_colors))
```

```{r}
# ── T-TESTS ───────────────────────────────────────────────────────────────
fan   <- join_dis |> filter(av_type == "fan")
delta <- join_dis |> filter(av_type == "delta")
il    <- join_dis |> filter(av_type == "intra-lobe")
retro <- join_dis |> filter(av_type == "retrogradational")

t_fan   <- t.test(fan$zlog_p95_q,      mu = 0)
t_delta <- t.test(delta$zlog_p95_q,    mu = 0)
t_il    <- t.test(il$zlog_p95_q,       mu = 0)
t_retro <- t.test(retro$zlog_p95_q,    mu = 0)
t_all   <- t.test(join_dis$zlog_p95_q, mu = 0)
print(t_all)

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar <- function(p) {
  if (p < 0.001)     "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else               "ns"
}

# ── ADD "ALL SITES" AS A FACTOR LEVEL ────────────────────────────────────
all_sites <- join_dis |>
  mutate(av_type_plot = "All Sites")

by_type <- join_dis |>
  mutate(av_type_plot = as.character(av_type))

plot_data <- bind_rows(all_sites, by_type) |>
  mutate(av_type_plot = factor(av_type_plot,
                                levels = c("All Sites", "delta", "fan",
                                           "intra-lobe", "retrogradational")))

# ── SIGNIFICANCE LABEL DATAFRAME ─────────────────────────────────────────
sig_labels <- data.frame(
  av_type_plot = factor(
    c("All Sites", "delta", "fan", "intra-lobe", "retrogradational"),
    levels = c("All Sites", "delta", "fan", "intra-lobe", "retrogradational")
  ),
  label = c(
    pstar(t_all$p.value),
    pstar(t_delta$p.value),
    pstar(t_fan$p.value),
    pstar(t_il$p.value),
    pstar(t_retro$p.value)
  )
) |>
  left_join(
    plot_data |>
      group_by(av_type_plot) |>
      summarise(y_pos = max(zlog_p95_q, na.rm = TRUE) + 0.2),
    by = "av_type_plot"
  )

# ── COLOURS INCLUDING ALL SITES ───────────────────────────────────────────
combined_colors <- c("All Sites" = "grey70", av_type_colors)

# ── SINGLE COMBINED PANEL ─────────────────────────────────────────────────
p_combined <- plot_data |>
  ggplot(aes(x = av_type_plot, y = zlog_p95_q, fill = av_type_plot, color = av_type_plot)) +

  geom_jitter(
    width = 0.2,
    alpha = 0.4,
    size  = 2
  ) +

  geom_boxplot(
    show.legend    = FALSE,
    alpha          = 0.4,
    outlier.shape  = NA,
    color          = "grey30"
  ) +

  stat_summary(
    fun      = mean,
    geom     = "crossbar",
    width    = 0.5,
    color    = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +

  geom_hline(
    yintercept = 0,
    linetype   = "dashed",
    color      = "grey30",
    linewidth  = 1.2
  ) +

  geom_vline(
    xintercept = 1.5,
    linetype   = "dotted",
    color      = "grey50",
    linewidth  = 0.8
  ) +

  # ── SIGNIFICANCE STARS ──────────────────────────────────────────────────
  geom_text(
    data        = sig_labels,
    aes(x = av_type_plot, y = y_pos, label = label),
    inherit.aes = FALSE,
    color       = "grey30",
    size        = 6,
    fontface    = "bold"
  ) +

  scale_fill_manual(values  = combined_colors) +
  scale_color_manual(values = combined_colors) +

  scale_x_discrete(labels = c(
    "All Sites"        = "All Sites",
    "delta"            = "Delta",
    "fan"              = "Fan",
    "intra-lobe"       = "Intra-lobe",
    "retrogradational" = "Retro."
  )) +

  labs(y = "log Discharge P95 Z (-)") +

  theme_classic(base_size = 18) +
  theme(
    axis.title.y    = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x    = element_blank(),
    axis.line.x     = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y     = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x     = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y     = element_text(color = "grey30", size = 16, face = "bold"),
    axis.ticks.y    = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks.x    = element_line(color = "grey30", linewidth = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA)
  )

print(p_combined)

ggsave(
  filename = here("plots_svg", "fig3_95.svg"),
  plot     = p_combined,
  bg       = "transparent",
  device   = svglite,
  height   = 4, width = 6, units = "in"
)

print(levels(plot_data$av_type_plot))
print(names(combined_colors))
```
### Log Mean Annual Discharge

```{r}

# ── COLOURS ───────────────────────────────────────────────────────────────
colors <- plasma(4)
av_type_colors <- setNames(colors, c("delta", "fan", "intra-lobe", "retrogradational"))
combined_colors <- c("All Sites" = "grey40", av_type_colors)
print(names(combined_colors))

# ── T-TESTS ───────────────────────────────────────────────────────────────
fan   <- join_dis |> filter(av_type == "fan")
delta <- join_dis |> filter(av_type == "delta")
il    <- join_dis |> filter(av_type == "intra-lobe")
retro <- join_dis |> filter(av_type == "retrogradational")

t_fan   <- t.test(fan$zlog_mean_q,      mu = 0)
t_delta <- t.test(delta$zlog_mean_q,    mu = 0)
t_il    <- t.test(il$zlog_mean_q,       mu = 0)
t_retro <- t.test(retro$zlog_mean_q,    mu = 0)
t_all   <- t.test(join_dis$zlog_mean_q, mu = 0)
print(t_all)

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar <- function(p) {
  if (p < 0.001)     "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else               "ns"
}

# ── SIGNIFICANCE LABEL DATAFRAME ─────────────────────────────────────────
# y position is set just above the max value for each group
sig_labels <- data.frame(
  av_type_plot = factor(
    c("All Sites", "delta", "fan", "intra-lobe", "retrogradational"),
    levels = c("All Sites", "delta", "fan", "intra-lobe", "retrogradational")
  ),
  label = c(
    pstar(t_all$p.value),
    pstar(t_delta$p.value),
    pstar(t_fan$p.value),
    pstar(t_il$p.value),
    pstar(t_retro$p.value)
  )
)

# calculate y position for each label (just above the max of each group)
y_positions <- plot_data |>
  group_by(av_type_plot) |>
  summarise(y_pos = max(zlog_mean_q, na.rm = TRUE) + 0.2)  # +0.2 padding above max

sig_labels <- sig_labels |>
  left_join(y_positions, by = "av_type_plot")

# ── ADD "ALL SITES" AS A FACTOR LEVEL ────────────────────────────────────
all_sites <- join_dis |>
  mutate(av_type_plot = "All Sites")

by_type <- join_dis |>
  mutate(av_type_plot = as.character(av_type))

plot_data <- bind_rows(all_sites, by_type) |>
  mutate(av_type_plot = factor(av_type_plot,
                                levels = c("All Sites", "delta", "fan",
                                           "intra-lobe", "retrogradational")))

# recalculate y positions now plot_data is defined
y_positions <- plot_data |>
  group_by(av_type_plot) |>
  summarise(y_pos = max(zlog_mean_q, na.rm = TRUE) + 0.2)

sig_labels <- sig_labels |>
  select(-y_pos) |>
  left_join(y_positions, by = "av_type_plot")

# ── SINGLE COMBINED PANEL ─────────────────────────────────────────────────
p_combined <- plot_data |>
  ggplot(aes(x = av_type_plot, y = zlog_mean_q, fill = av_type_plot, color = av_type_plot)) +

  geom_jitter(
    width = 0.2,
    alpha = 0.4,
    size  = 2
  ) +

  geom_boxplot(
    show.legend    = FALSE,
    alpha          = 0.4,
    outlier.shape  = NA,
    color          = "grey30"
  ) +

  stat_summary(
    fun      = mean,
    geom     = "crossbar",
    width    = 0.5,
    color    = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +

  geom_hline(
    yintercept = 0,
    linetype   = "dashed",
    color      = "grey30",
    linewidth  = 1.2
  ) +

  geom_vline(
    xintercept = 1.5,
    linetype   = "dotted",
    color      = "grey50",
    linewidth  = 0.8
  ) +

  # ── SIGNIFICANCE STARS ──────────────────────────────────────────────────
  geom_text(
    data    = sig_labels,
    aes(x = av_type_plot, y = y_pos, label = label),
    inherit.aes = FALSE,      # don't inherit fill/color from main aes
    color   = "grey30",
    size    = 6,
    fontface = "bold"
  ) +

  scale_fill_manual(values  = combined_colors) +
  scale_color_manual(values = combined_colors) +

  scale_x_discrete(labels = c(
    "All Sites"        = "All Sites",
    "delta"            = "Delta",
    "fan"              = "Fan",
    "intra-lobe"       = "Intra-lobe",
    "retrogradational" = "Retro."
  )) +

  labs(y = "log Discharge Mean Z (-)") +

  theme_classic(base_size = 18) +
  theme(
    axis.title.y    = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x    = element_blank(),
    axis.line.x     = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y     = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x     = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y     = element_text(color = "grey30", size = 16, face = "bold"),
    axis.ticks.y    = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks.x    = element_line(color = "grey30", linewidth = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA)
  )

print(p_combined)

ggsave(
  filename = here("plots_svg", "fig3_mean.svg"),
  plot     = p_combined,
  bg       = "transparent",
  device   = svglite,
  height   = 4, width = 6, units = "in"
)
```

### Log Flood Duration

```{r}

# ── COLOURS ───────────────────────────────────────────────────────────────
colors <- plasma(4)
av_type_colors <- setNames(colors, c("delta", "fan", "intra-lobe", "retrogradational"))
combined_colors <- c("All Sites" = "grey20", av_type_colors)
print(names(combined_colors))

# ── T-TESTS ───────────────────────────────────────────────────────────────
fan   <- join_dis |> filter(av_type == "fan")
delta <- join_dis |> filter(av_type == "delta")
il    <- join_dis |> filter(av_type == "intra-lobe")
retro <- join_dis |> filter(av_type == "retrogradational")

t_fan   <- t.test(fan$zlog_flood_dur,      mu = 0)
t_delta <- t.test(delta$zlog_flood_dur,    mu = 0)
t_il    <- t.test(il$zlog_flood_dur,       mu = 0)
t_retro <- t.test(retro$zlog_flood_dur,    mu = 0)
t_all   <- t.test(join_dis$zlog_flood_dur, mu = 0)
print(t_all)

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar <- function(p) {
  if (p < 0.001)     "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else               "ns"
}

# ── SIGNIFICANCE LABEL DATAFRAME ─────────────────────────────────────────
# y position is set just above the max value for each group
sig_labels <- data.frame(
  av_type_plot = factor(
    c("All Sites", "delta", "fan", "intra-lobe", "retrogradational"),
    levels = c("All Sites", "delta", "fan", "intra-lobe", "retrogradational")
  ),
  label = c(
    pstar(t_all$p.value),
    pstar(t_delta$p.value),
    pstar(t_fan$p.value),
    pstar(t_il$p.value),
    pstar(t_retro$p.value)
  )
)

# calculate y position for each label (just above the max of each group)
y_positions <- plot_data |>
  group_by(av_type_plot) |>
  summarise(y_pos = max(zlog_flood_dur, na.rm = TRUE) + 0.2)  # +0.2 padding above max

sig_labels <- sig_labels |>
  left_join(y_positions, by = "av_type_plot")

# ── ADD "ALL SITES" AS A FACTOR LEVEL ────────────────────────────────────
all_sites <- join_dis |>
  mutate(av_type_plot = "All Sites")

by_type <- join_dis |>
  mutate(av_type_plot = as.character(av_type))

plot_data <- bind_rows(all_sites, by_type) |>
  mutate(av_type_plot = factor(av_type_plot,
                                levels = c("All Sites", "delta", "fan",
                                           "intra-lobe", "retrogradational")))

# recalculate y positions now plot_data is defined
y_positions <- plot_data |>
  group_by(av_type_plot) |>
  summarise(y_pos = max(zlog_flood_dur, na.rm = TRUE) + 0.2)

sig_labels <- sig_labels |>
  select(-y_pos) |>
  left_join(y_positions, by = "av_type_plot")

# ── SINGLE COMBINED PANEL ─────────────────────────────────────────────────
p_combined <- plot_data |>
  ggplot(aes(x = av_type_plot, y = zlog_flood_dur, fill = av_type_plot, color = av_type_plot)) +

  geom_jitter(
    width = 0.2,
    alpha = 0.4,
    size  = 2
  ) +

  geom_boxplot(
    show.legend    = FALSE,
    alpha          = 0.4,
    outlier.shape  = NA,
    color          = "grey30"
  ) +

  stat_summary(
    fun      = mean,
    geom     = "crossbar",
    width    = 0.5,
    color    = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +

  geom_hline(
    yintercept = 0,
    linetype   = "dashed",
    color      = "grey30",
    linewidth  = 1.2
  ) +

  geom_vline(
    xintercept = 1.5,
    linetype   = "dotted",
    color      = "grey50",
    linewidth  = 0.8
  ) +

  # ── SIGNIFICANCE STARS ──────────────────────────────────────────────────
  geom_text(
    data    = sig_labels,
    aes(x = av_type_plot, y = y_pos, label = label),
    inherit.aes = FALSE,      # don't inherit fill/color from main aes
    color   = "grey30",
    size    = 6,
    fontface = "bold"
  ) +

  scale_fill_manual(values  = combined_colors) +
  scale_color_manual(values = combined_colors) +

  scale_x_discrete(labels = c(
    "All Sites"        = "All Sites",
    "delta"            = "Delta",
    "fan"              = "Fan",
    "intra-lobe"       = "Intra-lobe",
    "retrogradational" = "Retro."
  )) +

  labs(y = "log Flood Duration Z (-)") +

  theme_classic(base_size = 18) +
  theme(
    axis.title.y    = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x    = element_blank(),
    axis.line.x     = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y     = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x     = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y     = element_text(color = "grey30", size = 16, face = "bold"),
    axis.ticks.y    = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks.x    = element_line(color = "grey30", linewidth = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA)
  )

print(p_combined)

ggsave(
  filename = here("plots_svg", "fig3_dur.svg"),
  plot     = p_combined,
  bg       = "transparent",
  device   = svglite,
  height   = 4, width = 6, units = "in"
)

```
### Log SSC Data
```{r}

# ── RESHAPE DATA ──────────────────────────────────────────────────────────
plot_data <- join_sed |>
  select(zlog_p95_q, zlog_mean_q) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    variable = case_when(
      variable == "zlog_p95_q"  ~ "log P95 SSC Z (-)",
      variable == "zlog_mean_q" ~ "log Mean SSC Z (-)"
    ),
    variable = factor(variable, levels = c("log P95 SSC Z (-)", "log Mean SSC Z (-)"))
  )

# ── T-TESTS ───────────────────────────────────────────────────────────────
t_p95  <- t.test(join_sed$zlog_p95_q,  mu = 0)
t_mean <- t.test(join_sed$zlog_mean_q, mu = 0)

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar <- function(p) {
  if (p < 0.001)     "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else               "ns"
}

# ── SIGNIFICANCE LABEL DATAFRAME ─────────────────────────────────────────
y_positions <- plot_data |>
  group_by(variable) |>
  summarise(y_pos = max(value, na.rm = TRUE) + 0.2)

sig_labels <- data.frame(
  variable = factor(
    c("log P95 SSC Z (-)", "log Mean SSC Z (-)"),
    levels = c("log P95 SSC Z (-)", "log Mean SSC Z (-)")
  ),
  label = c(
    pstar(t_p95$p.value),
    pstar(t_mean$p.value)
  )
) |>
  left_join(y_positions, by = "variable")

# ── PLOT ──────────────────────────────────────────────────────────────────
ssc_overall <- plot_data |>
  ggplot(aes(x = variable, y = value, fill = variable, color = variable)) +
  # raw data
  geom_jitter(
    width = 0.15,
    alpha = 0.6,
    size = 2
  ) +
  # boxplot
  geom_boxplot(
    width = 0.35,
    alpha = 0.6,
    outlier.shape = NA,
    color = "grey30"  # box outline hardcoded, not from scale
  ) +
  # zero reference
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray30",
    linewidth = 1.2
  ) +
  # mean (solid red line)
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.35,
    color = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +
  # significance stars
  geom_text(
    data = sig_labels,
    aes(x = variable, y = y_pos, label = label),
    inherit.aes = FALSE,
    color = "grey30",
    size = 6,
    fontface = "bold"
  ) +
  scale_fill_manual(values  = c("log P95 SSC Z (-)" = "grey70", "log Mean SSC Z (-)" = "grey40")) +
  scale_color_manual(values = c("log P95 SSC Z (-)" = "grey70", "log Mean SSC Z (-)" = "grey40")) +
  labs(
    x = NULL,
    y = "SSC Z-Score (-)"
  ) +
  theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 20, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA)
  )

ssc_overall

ggsave(
  filename = here("plots_svg", "fig3_ssc.svg"),
  bg = "transparent",
  plot = ssc_overall,
  device = svglite,
  height = 4, width = 6, units = "in"
)

```

## Figure 4

### Leadup to Avulsions

#### Log P95 Annual Discharge

```{r}

plot_data <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%   # 1 = one year before, 5 = five years before
  group_by(av_type, years_before) %>%
  summarise(mean_zlog_p95_q = mean(zlog_p95_q, na.rm = TRUE), .groups = "drop")

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

p95_line <- ggplot(plot_data, aes(x = years_before, y = mean_zlog_p95_q,
                      color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 1.2) +
  geom_point(size = 5, stroke = 1, color = "white") +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +  # t-1 closest to avulsion on right
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.4, 0.3)) +
  labs(
    x = "Years Relative to Avulsion (years)",
    y = "P95 Discharge log Z-Score",
    color = "Avulsion Type",
    fill  = "Avulsion Type",
    shape = "Avulsion Type"
  ) +
    theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 15, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 15, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA),
    plot.margin     = margin(5, 8, 5, 8)
  )

ggsave(
  filename = file.path(plots_dir, "fig4_p95_line.svg"),
  plot     = p95_line,
  width    = 3,
  height   = 5,
  dpi      = 300,
  device = svglite, 
  bg       = "transparent"
)
p95_line

```

#### Log Mean Annual Discharge 

```{r}

plot_data1 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%   # 1 = one year before, 5 = five years before
  group_by(av_type, years_before) %>%
  summarise(mean_zlog_mean_q = mean(zlog_mean_q, na.rm = TRUE), .groups = "drop")

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

mean_line <- ggplot(plot_data1, aes(x = years_before, y = mean_zlog_mean_q,
                      color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 1.2) +
  geom_point(size = 5, stroke = 1, color = "white") +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +  # t-1 closest to avulsion on right
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.4, 0.3)) +
  labs(
    x = "Years Relative to Avulsion (years)",
    y = "Mean Discharge log Z-Score",
    color = "Avulsion Type",
    fill  = "Avulsion Type",
    shape = "Avulsion Type"
  ) +
    theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 15, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 15, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA),
    plot.margin     = margin(5, 8, 5, 8)
  )

mean_line

ggsave(
  filename = file.path(plots_dir, "fig4_mean_line.svg"),
  plot     = mean_line,
  width    = 3,
  height   = 5,
  dpi      = 300,
  device = svglite, 
  bg       = "transparent"
)

```

#### Log Mean Annual Discharge 

```{r}

plot_data2 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%   # 1 = one year before, 5 = five years before
  group_by(av_type, years_before) %>%
  summarise(mean_zlog_flood_dur = mean(zlog_flood_dur, na.rm = TRUE), .groups = "drop")

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

flooddur_line <- ggplot(plot_data2, aes(x = years_before, y = mean_zlog_flood_dur,
                      color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = 1.2) +
  geom_point(size = 5, stroke = 1, color = "white") +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +  # t-1 closest to avulsion on right
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.4, 0.3)) +
  labs(
    x = "Years Relative to Avulsion (years)",
    y = "Mean Flood Duration log Z-Score",
    color = "Avulsion Type",
    fill  = "Avulsion Type",
    shape = "Avulsion Type"
  ) +
    theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 15, face = "bold"),
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.x  = element_text(color = "grey30", size = 15, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 15, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    legend.position = "none",
    plot.background = element_rect(fill = "transparent", colour = NA),
    plot.margin     = margin(5, 8, 5, 8)
  )

flooddur_line

ggsave(
  filename = file.path(plots_dir, "fig4_flooddur.svg"),
  plot     = flooddur_line,
  width    = 3,
  height   = 5,
  device = svglite, 
  dpi      = 300,
  bg       = "transparent"
)

```
## Figure 5 

## GRL Theme

```{r}

grl_theme <- theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30", margin = margin(r = 6)),
    axis.title.x  = element_text(face = "bold", color = "grey30", margin = margin(t = 6)),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 0.8),
    axis.ticks    = element_line(color = "grey30", linewidth = 0.8),
    legend.position  = "none",
    plot.background  = element_rect(fill = "transparent", colour = NA),
    plot.margin      = margin(t = 6, r = 12, b = 6, l = 8)
  )


```

### Avulsion Count 

```{r}

# -------------------------
# Prep avulsion counts
# -------------------------
avulsion_counts <- join_dis %>%
  filter(!is.na(avulsion_year), !is.na(site_area)) %>%
  mutate(year = as.integer(avulsion_year)) %>%
  filter(site_area %in% c("Himalaya", "Pacific Islands", "South America")) %>%
  group_by(site_area, year) %>%
  summarise(n = n(), .groups = "drop")

all_sites <- join_dis %>%
  filter(!is.na(avulsion_year)) %>%
  mutate(year = as.integer(avulsion_year)) %>%
  group_by(year) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(site_area = "All Sites")

# -------------------------
# Poisson test per year
# -------------------------
# Total number of years observed
n_years  <- nrow(all_sites)
# Mean annual count = background Poisson rate
lambda   <- mean(all_sites$n)
mean_all <- lambda

# Run poisson.test() for each year's count against the background rate
# T = n_years so the expected value is lambda * 1 (per year)
poisson_results <- all_sites %>%
  rowwise() %>%
  mutate(
    p_value = poisson.test(n, T = 1, r = lambda)$p.value,
    sig     = p_value < 0.05,
    sig_label = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) %>%
  ungroup() %>%
  filter(sig)

# ---- Combine for plotting ----
df_plot <- bind_rows(avulsion_counts, all_sites)

site_areas <- c("South America", "Himalaya", "Pacific Islands")
all_levels <- c(site_areas, "All Sites")
df_plot    <- df_plot %>% mutate(site_area = factor(site_area, levels = all_levels))

n_sites <- length(site_areas)

ltype_vals <- setNames(rep("solid", length(all_levels)), all_levels)
lsize_vals <- setNames(c(rep(0.8, n_sites), 1.6), all_levels)
color_vals <- setNames(c("grey85", "grey45", "black", "#c45c8a"), all_levels)

```

### Plotting Avulsion Count 

```{r}
# -------------------------
# Plot
# -------------------------
p_av <- ggplot(
  df_plot,
  aes(
    x         = year,
    y         = n,
    color     = site_area,
    linetype  = site_area,
    linewidth = site_area,
    group     = site_area
  )
) +
  geom_line() +

  # Mean line
  geom_hline(
    yintercept = mean_all,
    linetype   = "dashed",
    color      = "black",
    linewidth  = 0.8,
    alpha      = 0.6
  ) +

  # Asterisks above significant peaks on the All Sites line
  geom_text(
    data        = poisson_results,
    aes(x = year, y = n, label = sig_label),
    inherit.aes = FALSE,
    vjust       = -0.5,
    size        = 10,
    fontface    = "bold",
    color       = "black"
  ) +

  scale_color_manual(values = color_vals, name = NULL) +
  scale_linetype_manual(values = ltype_vals, name = NULL) +
  scale_discrete_manual("linewidth", values = lsize_vals, name = NULL) +

  guides(
    color     = guide_legend(nrow = 1, byrow = TRUE),
    linetype  = guide_legend(nrow = 1, byrow = TRUE),
    linewidth = guide_legend(nrow = 1, byrow = TRUE)
  ) +

  scale_x_continuous(
    breaks = seq(1984, 2020, by = 4),
    limits = c(1983, 2021),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.0, 0.1))) +

  labs(x = "Year", y = "Avulsion\nCount") +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30",
                                 margin = margin(r = 8)),   # pushes label away from edge
    axis.title.x  = element_text(face = "bold", color = "grey30",
                                 margin = margin(t = 4)),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks    = element_line(color = "grey30", linewidth = 1.2),
    plot.margin   = margin(t = 6, r = 10, b = 6, l = 12),
    legend.position   = "none",
    plot.background   = element_rect(fill = "white", colour = NA)
  )

# -------------------------
# Save
# -------------------------
ggsave(
  filename = file.path(plots_dir, "fig5_avulsion_counts_by_site.svg"),
  plot     = p_av,
  width    = 9,
  height   = 3,
  dpi      = 300,
  device = svglite, 
  bg       = "white"
)
p_av
```

### Superelevation Plot

```{r}

# Loading in Alpha Beta Values from Nature Paper
alpha_beta <- read.csv(here("data", "nature_alpha_beta.csv")) |>
  clean_names() |>
  dplyr::select(avulsion_name, geomorphology, gamma, a, b, beta) |>
  rename(site_name = avulsion_name) |>
  mutate(site_name = str_replace_all(site_name, "_", "."))


# Joining Nature Data Paper to Avulsion Data
alpha_beta_av <- inner_join(join_dis, alpha_beta, by = c("site_name"))

```

```{r}

beta_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = beta, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
    geom_point(size = 3, stroke = 0.1, color = "white") +
  labs(
    x = "Mean Discharge log Z-Score (-)",
    y = "Beta (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme_minimal(base_size = 16) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30"),
    axis.title.x  = element_text(face = "bold", color = "grey30"),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks    = element_line(color = "grey30", linewidth = 1.2),
    legend.position   = "none",
    legend.direction  = "horizontal",
    plot.background   = element_rect(fill = "white", colour = NA)
  )

ggsave(
  filename = file.path(plots_dir, "fig5_beta.svg"),
  plot     = beta_dis,
  width    = 4.5,
  height   = 3,
  dpi      = 300,
  device = svglite, 
  bg       = "white"
)

beta_dis
```

### Slope Advantage Plot

```{r}

gamma_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = gamma, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  geom_point(size = 3, stroke = 0.1, color = "white") +
  labs(
    x = "Mean Discharge log Z-Score (-)",
    y = "Gamma (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),
    axis.title.y  = element_text(face = "bold", color = "grey30"),
    axis.title.x  = element_text( face = "bold", color = "grey30"),
    axis.text     = element_text(face = "bold", color = "grey30"),
    axis.line     = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks    = element_line(color = "grey30", linewidth = 1.2),
    legend.position   = "none",
    legend.direction  = "horizontal",
    plot.background   = element_rect(fill = "white", colour = NA)
  )

ggsave(
  filename = file.path(plots_dir, "fig5_gamma.svg"),
  plot     = gamma_dis,
  width    = 4.5,
  height   = 3,
  dpi      = 300,
  device   = svglite, 
  bg       = "white"
)

gamma_dis

```

```{r}

library(patchwork)

combined <- p_av / (beta_dis + gamma_dis)
combined

ggsave(
  filename = file.path(plots_dir, "fig5_combined.png"),
  plot     = combined,
  width    = 19,
  height   = 10,
  units    = "cm",
  dpi      = 300,
  bg       = "transparent"
)
```

## Supplemental Figures:

### Scatter Plot of Discharge and SSC

```{r}

# scatter_df <- join_dis %>%
#   inner_join(
#     join_sed %>%
#       select(site_name, p95_zscore_sed = p95_zscore),by = "site_name") |>
#   filter(!is.na(p95_zscore), !is.na(p95_zscore_sed))
# 
# p_scatter <- ggplot(scatter_df, aes(x = p95_zscore, y = p95_zscore_sed, color = av_type)) +
#   geom_point(
#     size = 3,
#     alpha = 0.7) +
#   geom_hline(
#     yintercept = 0,
#     linetype = "dashed",
#     color = "gray30",
#     linewidth = 1.2) +
#    geom_vline(
#     xintercept = 0,
#     linetype = "dashed",
#     color = "gray30",
#     linewidth = 1.2) +
#   theme_minimal() +
#   scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
#   labs(x = "Discharge Z-Score (-)", y = "SSC Z-Score (-)") +
#   theme(
#     axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
#     axis.title.x = element_text(color = "grey30", size = 20, face = "bold"),
#     
#     axis.line.x = element_line(color = "grey30", linewidth = 1.2),
#     axis.line.y = element_line(color = "grey30", linewidth = 1.2),
#     
#     axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
#     axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
#     
#     axis.ticks.y = element_line(color = "grey30", size = 1.2),
#     axis.ticks.x = element_line(color = "grey30", size = 1.2),
#     
#     legend.position = "none",
#     
#   plot.background = element_rect(fill = "transparent", colour = NA) # Transparent plot bg
#   )
# p_scatter
# 
# ggsave(
#   filename = here("plots_2.0", "supplemental_scatter.png"),
#   plot = p_scatter,
#   width = 4,
#   height = 4,
#   units = "in",
#   dpi = 300
# )

```

### World Map of Discharge Z-Score

```{r}
# #Set Up Map Extent
# xmin <- -170
# xmax <- 170
# 
# #Find Standard Deviation for the Dataset
# #Purpose: use for the scale to condense and make a little easier to visualize 
# std_dev_dis <- sd(join_dis$p95_zscore, na.rm = TRUE)
# 
# world_map <- ne_countries(scale = "medium", returnclass = "sf") |>
#   filter(continent != "Antarctica")
# 
# world_map_plot <- ggplot() +
#   geom_sf(data = world_map, fill = "grey80", color = "grey80", linewidth = 0.3) +
#   geom_point(
#     data = join_dis,
#     aes(x = site_lon, y = site_lat, color = p95_zscore),
#     size = 3,
#     alpha = 0.85,
#     shape = 16
#   ) +
#   scale_color_distiller(
#     name = "Discharge Z-Score",
#     palette = "Spectral",
#     limits = c(-0.8, 0.8),
#     oob = scales::squish
#   ) +
#   coord_sf(xlim = c(xmin, xmax), ylim = c(-60, 85), expand = FALSE) +
# 
#   theme_void(base_size = 14) +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background  = element_rect(fill = "white", color = NA),
# 
#     # Legend placement and sizing
#     legend.position = "top",
#     legend.justification = "center",
#     legend.box = "vertical",
# 
#     # Text sizing
#     legend.title = element_text(face = "bold", size = 20),
#     legend.text  = element_text(size = 20),
# 
#     # Spacing
#     legend.margin = margin(t = 12, b = 12),
#     legend.spacing.x = unit(1.5, "cm"),
# 
#     # Give the plot a bit more breathing room
#     plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
#   ) +
#   guides(
#     color = guide_colorbar(
#       title.position = "top",
#       title.hjust = 0.5,
#       barwidth  = unit(0.4, "npc"),   # spans most of plot width
#       barheight = unit(0.4, "cm"),
#       direction = "horizontal"
#     )
#   )
# 
# world_map_plot
# 
# ggsave(
#   filename = here("plots_2.0", "supplemental_dis_map.png"),
#   plot = world_map_plot,
#   width = 16,
#   height = 8,
#   units = "in",
#   dpi = 300
# )
# 

```

### World Map of SSC

```{r}

# #Set Map Extents
# xmin <- -170
# xmax <- 170
# 
# #Find Standard Deviation for the Dataset
# #Purpose: use for the scale to condense and make a little easier to visualize 
# std_dev_sed <- sd(join_sed$p95_zscore, na.rm = TRUE)
# 
# world_map <- ne_countries(
#   scale = "medium", returnclass = "sf") |>
#   filter(continent != "Antarctica")
# 
# world_map_plot <- ggplot() +
#   geom_sf(data = world_map, fill = "grey80", color = "grey80", linewidth = 0.3) +
#   geom_point(
#     data = join_sed,
#     aes(x = site_lon, y = site_lat, color = p95_zscore),
#     size = 3,
#     alpha = 0.85,
#     shape = 16
#   ) +
#   scale_color_distiller(
#     name = "SSC Z-Score",
#     palette = "Spectral",
#     limits = c(-std_dev_sed, std_dev_sed),
#     oob = scales::squish
#   ) +
#   coord_sf(xlim = c(xmin, xmax), ylim = c(-60, 85), expand = FALSE) +
# 
#   theme_void(base_size = 14) +
#   theme(
#     panel.background = element_rect(fill = "white", color = NA),
#     plot.background  = element_rect(fill = "white", color = NA),
# 
#     # Legend placement and sizing
#     legend.position = "top",
#     legend.justification = "center",
#     legend.box = "vertical",
# 
#     # Text sizing
#     legend.title = element_text(face = "bold", size = 20),
#     legend.text  = element_text(size = 20),
# 
#     # Spacing
#     legend.margin = margin(t = 12, b = 12),
#     legend.spacing.x = unit(1.5, "cm"),
# 
#     # Give the plot a bit more breathing room
#     plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
#   ) +
#   guides(
#     color = guide_colorbar(
#       title.position = "top",
#       title.hjust = 0.5,
#       barwidth  = unit(0.4, "npc"),   # spans most of plot width
#       barheight = unit(0.4, "cm"),
#       direction = "horizontal"
#     )
#   )
# 
# world_map_plot
# 
# ggsave(
#   filename = here("plots_2.0", "supplemental_ssc_map.png"),
#   plot = world_map_plot,
#   width = 16,
#   height = 8,
#   units = "in",
#   dpi = 300
# )
```


