---
title: "Templates for All Maps and Graphs + Descriptive Statistics"
description: "This document is meant to act as the location where all final version graphs are meant to be created, and all descriptive statistics put onto those graphs will be calcualted. This synthesizes material from other documents up to this point. "
author: "Sam Lance"
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
    theme: Lux
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r, include =FALSE}
library(dplyr)
library(tidyverse)
library(lubridate)
library(readr)
library(here)
library(ggplot2)
library(stringr)
library(purrr)
library(maps)
library(knitr)
library(kableExtra)
library(sf)
library(RColorBrewer)
library(patchwork)
library(tmap)
library(viridis)
library(sp)
library(rworldmap)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(devtools)
```

## Reading In + Cleaning Data

### Reading in Data

```{r, output=FALSE}}
#DISCHARGE
join_dis <- read_csv(here("data","discharge_av_year_z.csv")) |>
  inner_join(
    read_csv(here("data","base_sites.csv")) |>
      select(site_name, site_lat, site_lon),   # no av_type
    by = "site_name") |>
  drop_na()

#Sediment 
join_sed = read_csv(here("data","ssc_av_year_z.csv")) |>
  drop_na()

```

### Cleaning Discharge Data

```{r}
#Convert DF to SF Object 
sf_dis <- st_as_sf(join_dis, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_dis, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_dis <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

```

### Cleaning Sediment Data

```{r}

#Convert DF to SF Object 
sf_sed <- st_as_sf(join_sed, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_sed, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_sed <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

```

## Stats for Results

### Discharge Mean + Median

```{r}

mean_dis <- mean(join_dis$p95_zscore)
median_dis <- median(join_dis$p95_zscore)
range_dis <- range(join_dis$p95_zscore)

print(c('Mean Dis:', mean_dis))
print(c('Median Dis', median_dis))
print(c('Range Dis', range_dis))

# Create Histogram of this Data
hist(join_dis$p95_zscore, 
     main = "Histogram of Discharge Z-Scores", 
     xlab = "Discharge Z-Score", 
     col = "lightblue", 
     border = "black")

```

### SSC Mean + Median

```{r}

mean_sed <- mean(join_sed$p95_zscore)
median_sed <- median(join_sed$p95_zscore)
range_sed <- range(join_sed$p95_zscore)

print(c('Mean SSC:', mean_sed))
print(c('Median SSC', median_sed))
print(c('Range SSC', range_sed))

#Create Histogram of this Data
hist(join_sed$p95_zscore, 
     main = "Histogram of Sediment Z-Scores", 
     xlab = "Sediment Z-Score", 
     col = "lightgreen", 
     border = "black")

```

## Figure 1 - World Map with Avulsion Types

### Establishing Colors and Shapes for All Maps

```{r}
#Loading in World Map Data
world_map <- ne_countries(scale = "medium", returnclass = "sf")

#Establish Legend Breaks, Colors, and Desired Shapes
colors <- plasma(4)
rev_colors <- colors
shapes <- c(21, 22, 23, 24)

#devtools::install_github("an-bui/calecopal")# filled shapes
# breaks <- c(-2, -1, 0, 1, 2, 3, 4, 5)
# colors <- brewer.pal(length(breaks) - 1, "RdYlBu")
# rev_colors <- rev(colors)  # Reverse the color order
# shapes <- c(16, 17, 15, 18)  # Different shapes for avulsion types

```

### Robinson Projection World Map

```{r}

library(tidyverse)
library(sf)
library(ggplot2)

# -------------------------
# Inputs assumed to exist:
# world_map (sf polygons, lon/lat)
# sf_dis (dataframe with site_lon, site_lat, site_name, av_type)
# rev_colors (vector of colors)
# shapes (named vector of shapes for av_type)
# -------------------------

# 1) Select sites to highlight
selected_sites <- sf_dis |>
  filter(site_name %in% c("B9", "R3.2", "JA.25.India.1990", "MANAMBOLO.2004"))

# 2) Ensure factor levels consistent
sf_dis <- sf_dis |>
  mutate(av_type = factor(av_type))

# 3) Named color vector for av_type
av_type_colors <- setNames(rev_colors, levels(sf_dis$av_type))

# 4) Robinson CRS
crs_robin <- "+proj=robin +lon_0=0 +datum=WGS84 +units=m +no_defs"

# 5) Make points sf (critical for correct projection)
sf_dis_sf <- st_as_sf(
  sf_dis,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

selected_sites_sf <- st_as_sf(
  selected_sites,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

# 6) Ensure world_map has CRS (if not already)
if (is.na(st_crs(world_map))) st_crs(world_map) <- 4326

# 7) Create a "globe" outline and optional graticule
globe_outline <- st_as_sfc(
  st_bbox(c(xmin = -180, ymin = -90, xmax = 180, ymax = 90), crs = st_crs(4326))
) |>
  st_transform(crs_robin)

# Optional graticule
grat <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = seq(-60, 60, by = 30),
  crs = st_crs(4326)
) |>
  st_transform(crs_robin)

# -------------------------
# NEW: Top + bottom boundary (latitude caps)
# -------------------------
lat_min <- -90
lat_max <- 90
lon_seq <- seq(-180, 180, by = 0.25)

make_parallel <- function(lat) {
  st_sfc(
    st_linestring(cbind(lon_seq, rep(lat, length(lon_seq)))),
    crs = 4326
  ) |>
    st_transform(crs_robin)
}

bottom_boundary <- make_parallel(lat_min)
top_boundary    <- make_parallel(lat_max)

# 8) Build plot (Robinson + globe boundary)
world_map_plot <- ggplot() +

  # Light graticule behind everything (optional; remove if you don’t want lines)
  geom_sf(data = grat, color = "grey70", linewidth = 0.25) +

  # Continents
  geom_sf(
    data = world_map,
    fill = "grey70",
    color = "grey60",
    linewidth = 0.25
  ) +

  # All sites
  geom_sf(
    data = sf_dis_sf,
    aes(fill = av_type, shape = av_type, color = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 0.5
  ) +

  # Selected sites: same fill/shape, red outline
  geom_sf(
    data = selected_sites_sf,
    aes(fill = av_type, shape = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 1.0,
    color = "black",
    show.legend = FALSE
  ) +

  # -------------------------
  # NEW: draw top/bottom boundary lines
  # -------------------------
  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  # Globe outline (draw last so it frames everything)
  geom_sf(
    data = globe_outline,
    fill = NA,
    color = "grey70",
    linewidth = 0.7
  ) +

  # Scales
  scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +

  # Robinson projection
  coord_sf(crs = crs_robin, expand = FALSE) +

  # Styling
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    
    #legend.position = "none",
    legend.position = c(0.2, 0.25),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    legend.spacing.y = unit(-0.1, "cm"),
    legend.key.height = unit(0.3, "lines"),
    legend.background = element_rect(fill = "white", color = "white")
  ) +
  guides(
    fill  = guide_legend(byrow = TRUE),
    shape = guide_legend(byrow = TRUE)
  )

world_map_plot

ggsave(
  filename = here("plots_2.0", "fig1_world_robinson_globe.png"),
  plot = world_map_plot,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300
)

```

## Figure 2 - Time Series

### Sediment

```{r}

# -------------------------
# Paths
# -------------------------
av_year_updated_path <- here("data", "ssc_av_year_z.csv")
compiled_output_path <- here("data", "time_series.csv")
plots_dir <- "/Volumes/Crucial X6/paper_figures/rstudio/plots_2.0"

# -------------------------
# Load data
# -------------------------
av_year_updated <- read_csv(av_year_updated_path, show_col_types = FALSE)
compiled <- read_csv(compiled_output_path, show_col_types = FALSE)

compiled <- compiled %>%
  mutate(date = as.Date(date))

# -------------------------
# SELECT ONE SITE HERE
# -------------------------
site <- "MALAWI1.2014"

site_df <- compiled %>%
  filter(site_name == site) %>%
  arrange(date)

stopifnot(nrow(site_df) > 0)

# -------------------------
# Avulsion shading window
# -------------------------
avs <- if ("avulsion_start_date" %in% names(site_df)) {
  site_df %>%
    pull(avulsion_start_date) %>%
    as.numeric() %>%
    na.omit() %>%
    unique()
} else {
  numeric(0)
}

shade_df <- NULL
if (length(avs) > 0) {
  shade_df <- tibble(
    xmin = as.Date(paste0(avs[1] - 1, "-01-01")),
    xmax = as.Date(paste0(avs[1], "-12-31")),
    ymin = -Inf,
    ymax = Inf
  )
}

# -------------------------
# Mean 95th percentile line (from zscore table)
# -------------------------
year_p95 <- av_year_updated %>%
  filter(site_name == site) %>%
  pull(p95_mean) %>%
  first()

# -------------------------
# 95th percentile value within the shaded window,
# plotted at the exact date of the maximum within the window
# -------------------------
p95_pt <- NULL

if (!is.null(shade_df)) {
  window_df <- site_df %>%
    filter(date >= shade_df$xmin, date <= shade_df$xmax) %>%
    filter(!is.na(y_pred))

  if (nrow(window_df) > 0) {
    p95_val <- as.numeric(quantile(window_df$y_pred, probs = 0.95, na.rm = TRUE, type = 7))

    max_row <- window_df %>%
      filter(y_pred == max(y_pred, na.rm = TRUE)) %>%
      arrange(date) %>%   # if ties, use earliest max date
      slice(1)

    p95_pt <- tibble(
      date = max_row$date,
      y_pred = p95_val
    )
  }
}

# -------------------------
# Plot
# -------------------------
p_ssc <- ggplot(site_df, aes(x = date, y = y_pred)) +

  # Avulsion shading
  {if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
    # Mean p95 line
  {if (!is.na(year_p95))
    geom_hline(
      yintercept = year_p95,
      linetype = "dashed",
      color = "black",
      linewidth = 0.75,
      alpha = 0.75
    )
  } +

  # Time series
  geom_point(color = "#9bbf30", size = 0.75) +
  geom_line(color = "#9bbf30", linewidth = 0.75) +

# Red hollow circle: y = p95(window), x = date of max(window)
{ if (!is.null(p95_pt))
  geom_point(
    data = p95_pt,
    aes(x = date, y = y_pred),
    inherit.aes = FALSE,
    shape = 21,        # hollow circle
    size = 5,
    stroke = 2,      # outline thickness
    fill = NA,
    color = "red",
    alpha = 0.7
  )
} +

  # Axes (keep only scale_x_date to avoid scale replacement warnings)
  scale_x_date(
    limits = as.Date(c("1985-01-01", "2020-12-31")),
    breaks = seq(as.Date("1985-01-01"),
                 as.Date("2020-01-01"),
                 by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = c(0, 0)) +

  labs(
    x = "Year",
    y = "SSC (mg/L)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(
      linetype  = "solid",
      linewidth = 0.25,
      color = "grey90"),
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA)
  )

# -------------------------
# Save
# -------------------------
safe_name <- site %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("/", "_")

ggsave(
  filename = file.path(plots_dir, paste0(safe_name, "fig2_ssc_timeseries.png")),
  plot = p_ssc,
  width = 9,
  height = 3,
  dpi = 300
)

p_ssc


```

### Discharge

```{r}
# Paths
av_year_path  <- here("data","discharge_av_year_z.csv")
compiled_path <- here("data", "compiled_dis24.csv")
plots_dir     <- "/Volumes/Crucial X6/paper_figures/rstudio/plots_2.0"

dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)

# Read data
compiled        <- read_csv(compiled_path, show_col_types = FALSE)
av_year_updated <- read_csv(av_year_path, show_col_types = FALSE)

# Parse datetime
compiled <- compiled %>%
  mutate(valid_time = as.POSIXct(valid_time, tz = "UTC"))

xmin <- as.POSIXct("1985-01-01", tz = "UTC")
xmax <- as.POSIXct("2020-12-31", tz = "UTC")

# -------------------------
# CHOOSE ONE SITE HERE
# -------------------------
site <- "MALAWI1.2014"

site_df <- compiled %>%
  filter(site_name == site) %>%
  arrange(valid_time)

stopifnot(nrow(site_df) > 0)

# ---- Avulsion shading window (if column exists + has a value) ----
avs <- if ("avulsion_start_date" %in% names(site_df)) {
  site_df %>%
    pull(avulsion_start_date) %>%
    as.numeric() %>%
    na.omit() %>%
    unique()
} else {
  numeric(0)
}

shade_df <- NULL
if (length(avs) > 0) {
  shade_df <- tibble(
    xmin = as.POSIXct(paste0(avs[1] - 1, "-01-01"), tz = "UTC"),
    xmax = as.POSIXct(paste0(avs[1], "-12-31"), tz = "UTC"),
    ymin = -Inf,
    ymax = Inf
  )
}

# ---- Mean 95th percentile line for this site (from zscore table) ----
year_p95 <- av_year_updated %>%
  filter(site_name == site) %>%
  pull(p95_mean) %>%
  first()

# ---- 95th percentile point within shaded window (x = date of max in window) ----
p95_pt <- NULL

if (!is.null(shade_df)) {
  window_df <- site_df %>%
    filter(valid_time >= shade_df$xmin, valid_time <= shade_df$xmax) %>%
    filter(!is.na(dis24))

  if (nrow(window_df) > 0) {
    # window-specific 95th percentile value
    p95_val <- as.numeric(quantile(window_df$dis24, probs = 0.95, na.rm = TRUE, type = 7))

    # exact date of maximum discharge within the window (tie-break: earliest)
    max_row <- window_df %>%
      filter(dis24 == max(dis24, na.rm = TRUE)) %>%
      arrange(valid_time) %>%
      slice(1)

    # plot p95 value at the max date
    p95_pt <- tibble(
      valid_time = max_row$valid_time,
      dis24      = p95_val
    )
  }
}

# ---- Plot ----
p_dis <- ggplot(site_df, aes(x = valid_time, y = dis24)) +

  # Avulsion shading
  {if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
    # Mean p95 line
  {if (!is.na(year_p95))
    geom_hline(
      yintercept = year_p95,
      color = "black",
      linetype = "dashed",
      linewidth = 0.75,
      alpha = 0.75
    )
  } +

  # Base line
  geom_point(color = "#2f6fa5", size = 0.75) +
  geom_line(color = "#2f6fa5", linewidth = 0.75) +

  # Blue point: y = p95(window), x = date of max(window)
  { if (!is.null(p95_pt))
  geom_point(
    data = p95_pt,
    aes(x = valid_time, y = dis24),
    inherit.aes = FALSE,
    shape = 21,        # hollow circle
    size = 5,
    stroke = 2,      # outline thickness
    fill = NA,
    color = "red",
    alpha = 0.7
  )
} +

  scale_y_continuous(expand = c(0, 0)) +

  # Date ticks + limits (add expand here; drop scale_x_continuous to avoid scale conflicts)
  scale_x_datetime(
    limits = c(xmin, xmax),
    breaks = seq(xmin, xmax, by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(
    x = "Year",
    y = "Discharge (m³/s)"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.20, color = "grey90"),
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA)
  )

# ---- Save (one file) ----
safe <- site %>%
  as.character() %>%
  str_replace_all(" ", "_") %>%
  str_replace_all("/", "_")

ggsave(
  filename = file.path(plots_dir, paste0(safe, "fig2_dis24_timeseries.png")),
  plot = p_dis,
  width = 9,
  height = 3,
  dpi = 300
)

# Print the plot in RStudio/Quarto
p_dis



```
### Creating Time Series of Discharge and Sediment Z-Scores for Each Year 

```{r}
#Load One Site Discharge and Sediment Z-Score
malawi_dis_zscores <- read_csv(here("data", "discharge_zscores_one_site.csv"))
malawi_ssc_zscores <- read_csv(here("data", "ssc_zscores_one_site.csv"))
```

```{r}

# Optional: select one site if present
# site <- "MALAWI1.2014"
# df_red  <- df_red  %>% filter(site_name == site)
# df_blue <- df_blue %>% filter(site_name == site)

# -------------------------
# Clean + harmonize
# -------------------------
prep_df <- function(df, label) {
  df %>%
    mutate(
      year = as.integer(year),
      p95_zscore = as.numeric(p95_zscore),
      dataset = label,
      date = as.Date(paste0(year, "-01-01"))
    ) %>%
    filter(!is.na(year), !is.na(p95_zscore))
}

dis_red  <- prep_df(malawi_dis_zscores,  "Discharge (m^3/s)")
ssc_blue <- prep_df(malawi_ssc_zscores, "SSC (mg/L)")

df_all <- bind_rows(dis_red, ssc_blue)

stopifnot(nrow(df_all) > 0)

# ---- Shading window: 2013 through end of 2014 ----
shade_df <- tibble(
  xmin = as.Date("2013-01-01"),
  xmax = as.Date("2014-12-31"),
  ymin = -Inf,
  ymax = Inf
)
# -------------------------
# Marker year
# -------------------------
mark_year_dis <- 2014L
mark_year_ssc <- 2013L

mark_pts_dis <- df_all %>%
  filter(dataset == "Discharge (m^3/s)", year == mark_year_dis)

mark_pts_ssc <- df_all %>%
  filter(dataset == "SSC (mg/L)", year == mark_year_ssc)

# -------------------------
# X limits
# -------------------------
xmin <- as.Date(paste0(min(df_all$year), "-01-01"))
xmax <- as.Date(paste0(max(df_all$year), "-12-31"))

# -------------------------
# Plot
# -------------------------
p_z <- ggplot(df_all, aes(x = date, y = p95_zscore, color = dataset)) +
  
  { if (!is.null(shade_df))
    geom_rect(
      data = shade_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "#fff7bc",
      alpha = 0.6
    )
  } +
  
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +

  { if (nrow(mark_pts_dis) > 0)
    geom_point(
      data = mark_pts_dis,
      aes(x = date, y = p95_zscore),
      inherit.aes = FALSE,
      shape = 21,
      size = 5,
      stroke = 2,
      fill = NA,
      color = "red",
      alpha = 0.8
    )
  } +
  
  { if (nrow(mark_pts_ssc) > 0)
    geom_point(
      data = mark_pts_ssc,
      aes(x = date, y = p95_zscore),
      inherit.aes = FALSE,
      shape = 21,
      size = 5,
      stroke = 2,
      fill = NA,
      color = "red",
      alpha = 0.8
    )
  } +

  scale_color_manual(
  values = c(
    "Discharge (m^3/s)" = "#2f6fa5",  # deep red
    "SSC (mg/L)"        = "#9bbf30"   # deep blue
  )
) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE)) +

  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +

  scale_x_date(
    limits = c(as.Date("1985-01-01"), as.Date("2020-12-31")),
    breaks = seq(as.Date("1985-01-01"), as.Date("2020-01-01"), by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(x = "Year", y = "Z-Score (-)") +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.2, color = "grey90"),

    axis.title.y = element_text(size = 20, face = "bold", color = "grey30"),
    axis.title.x = element_text(size = 20, face = "bold", color = "grey30"),

    axis.text = element_text(size = 16, face = "bold", color = "grey30"),
    axis.line = element_line(color = "grey30", linewidth = 1.2),
    axis.ticks = element_line(color = "grey30", linewidth = 1.2),

    legend.position = "none",
    legend.justification = c(0.5, 1),
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.text = element_text(size = 14, face = "bold", color = "grey30"),
    legend.background = element_rect(fill = alpha("white", 0.7), colour = NA),
    legend.key = element_rect(fill = NA, colour = NA),

    plot.background = element_rect(fill = "transparent", colour = NA)
  )
# -------------------------
# Save
# -------------------------
ggsave(
  filename = file.path(plots_dir, "fig2_p95_zscore_two_datasets.png"),
  plot = p_z,
  width = 9,
  height = 3.1,
  dpi = 300,
  bg = "transparent"
)

p_z

```


## Figure 3 - Boxplots of Discharge Z-Score

### Boxplot of Discharge Z-Scores by Type

```{r}
#Setting Up Same Colors as Above 
colors <- plasma(4)
av_type_colors <- setNames(colors, levels(join_dis$av_type))


#Establishing Title
title <- "Z-Score of Peak Discharge for Avulsion Year and Year Prior by Avulsion Type"

#Computing T-Tests
fan = join_dis |>
  filter(av_type == "fan")

delta = join_dis |>
  filter(av_type == "delta")

il = join_dis |>
  filter(av_type == "intra-lobe")

retro = join_dis |>
  filter(av_type == "retrogradational")


t_fan = t.test(fan$p95_zscore, mu = 0)
t_delta = t.test(delta$p95_zscore, mu = 0)
t_il = t.test(il$p95_zscore, mu = 0)
t_retro = t.test(retro$p95_zscore, mu = 0)
t_all = t.test(join_dis$p95_zscore, mu = 0)

print(t_all)

#Setting Up Plot Base
gardner_type <- join_dis |>
  ggplot(aes(x = av_type, y = p95_zscore, fill = av_type)) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 2, color = "gray30") +
  geom_boxplot(show.legend = FALSE, alpha = 0.5) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, color = scales::alpha("red", 0.50)) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 1.2) +
  labs(
       x = "Avulsion Type", y = "Discharge Z-Score (-)") +
  theme_classic(base_size = 18) +
  #scale_fill_brewer(palette = "Greys") +  
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA) # Transparent plot bg
  )
print(gardner_type)

ggsave(
  filename = here("plots_2.0", "fig3_dis_zscore_type_box.jpg"),
  plot = gardner_type,
  device = "jpg",
  height = 4, width = 8, units = "in"
)

```

### Boxplot of Overall Discharge

```{r}

#Create boxplot of all discharge 

overall_dis <- join_dis |>
  ggplot(aes(x = "all sites", y = p95_zscore)) +

  # raw data
  geom_jitter(
    width = 0.15,
    alpha = 0.6,
    size = 2,
    color = "gray30"
  ) +

  # boxplot
  geom_boxplot(
    width = 0.35,
    fill = "grey70",
    alpha = 0.6,
    outlier.shape = NA
  ) +

  # zero reference
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey30",
    linewidth = 1.2
  ) +

  # mean (solid red line)
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.35,
    color = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +

  labs(
    x = NULL,
    y = "Discharge Z-Score (-)"
  ) +

  theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 20, face = "bold"),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA) # Transparent plot bg
  )

overall_dis

ggsave(
  filename = here("plots_2.0", "fig3_discharge_overall_box.png"), 
  bg = "transparent",
  plot = overall_dis, 
  device = "png",
  height = 4, width = 4, units = "in"
)


```

### Boxplot of SSC:

```{r}

#Establishing Title
#title <- "Z-Score of Peak SSC for Avulsion Year and Year Prior by Avulsion Type"

#Computing T-Tests
fan = join_sed |>
  filter(av_type == "fan")

delta = join_sed |>
  filter(av_type == "delta")

il = join_sed |>
  filter(av_type == "intra-lobe")

retro = join_sed |>
  filter(av_type == "retrogradational")


t_fan = t.test(fan$p95_zscore, mu = 0)
print( t_fan)
t_delta = t.test(delta$p95_zscore, mu = 0)
print(t_delta)
t_il = t.test(il$p95_zscore, mu = 0)
t_retro = t.test(retro$p95_zscore, mu = 0)
t_all = t.test(join_sed$p95_zscore, mu = 0)

print(t_all)

#Setting Up Plot Base
gardner_type <- join_sed |>
  ggplot(aes(x = av_type, y = p95_zscore, fill = av_type)) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 2, color = "gray30") +
  geom_boxplot(show.legend = FALSE, alpha = 0.5) +
  stat_summary(fun = mean, 
               geom = "crossbar", 
               width = 0.5, 
               color = scales::alpha("red", 0.50)) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 1.2) +
  labs(title = element_blank(), 
       x = "Avulsion Type", y = "SSC Z-Score (-)") +
  theme_classic(base_size = 18) +
  #scale_fill_brewer(palette = "Greys") +  
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_blank(),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA) # Transparent plot bg
  )
print(gardner_type)

ggsave(
  filename = here("plots_2.0", "fig3_ssc_zscore_type_box.jpg"), 
  plot = gardner_type, 
  device = "jpg",
  height = 8, width = 8, units = "in"
)

```

### Boxplot of Overall SSC:

```{r}

ssc_overall <- join_sed |>
  ggplot(aes(x = "all sites", y = p95_zscore)) +

  # raw data
  geom_jitter(
    width = 0.15,
    alpha = 0.6,
    size = 2,
    color = "gray70"
  ) +

  # boxplot
  geom_boxplot(
    width = 0.35,
    fill = "grey70",
    alpha = 0.6,
    outlier.shape = NA
  ) +

  # zero reference
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray30",
    linewidth = 1.2
  ) +

  # mean (solid red line)
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.35,
    color = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +

  labs(
    x = NULL,
    y = "SSC Z-Score (-)"
  ) +

  theme_classic(base_size = 18) +
  theme(
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 20, face = "bold"),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA) # Transparent plot bg
  )

ssc_overall

ggsave(
  filename = here("plots_2.0", "fig3_ssc_overall_box.png"), 
  bg = "transparent",
  plot = ssc_overall, 
  device = "png",
  height = 4, width = 4, units = "in"
)

```

## Supplemental Figures:

### World Map of Discharge Z-Score

```{r}
#Set Up Map Extent
xmin <- -170
xmax <- 170

#Find Standard Deviation for the Dataset
#Purpose: use for the scale to condense and make a little easier to visualize 
std_dev_dis <- sd(join_dis$p95_zscore, na.rm = TRUE)

world_map <- ne_countries(scale = "medium", returnclass = "sf") |>
  filter(continent != "Antarctica")

world_map_plot <- ggplot() +
  geom_sf(data = world_map, fill = "grey80", color = "grey80", linewidth = 0.3) +
  geom_point(
    data = join_dis,
    aes(x = site_lon, y = site_lat, color = p95_zscore),
    size = 3,
    alpha = 0.85,
    shape = 16
  ) +
  scale_color_distiller(
    name = "Discharge Z-Score",
    palette = "Spectral",
    limits = c(-0.8, 0.8),
    oob = scales::squish
  ) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(-60, 85), expand = FALSE) +

  theme_void(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),

    # Legend placement and sizing
    legend.position = "top",
    legend.justification = "center",
    legend.box = "vertical",

    # Text sizing
    legend.title = element_text(face = "bold", size = 20),
    legend.text  = element_text(size = 20),

    # Spacing
    legend.margin = margin(t = 12, b = 12),
    legend.spacing.x = unit(1.5, "cm"),

    # Give the plot a bit more breathing room
    plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
  ) +
  guides(
    color = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth  = unit(0.4, "npc"),   # spans most of plot width
      barheight = unit(0.4, "cm"),
      direction = "horizontal"
    )
  )

world_map_plot

ggsave(
  filename = here("plots_2.0", "supplemental_dis_map.png"),
  plot = world_map_plot,
  width = 16,
  height = 8,
  units = "in",
  dpi = 300
)


```

### World Map of SSC

```{r}

#Set Map Extents
xmin <- -170
xmax <- 170

#Find Standard Deviation for the Dataset
#Purpose: use for the scale to condense and make a little easier to visualize 
std_dev_sed <- sd(join_sed$p95_zscore, na.rm = TRUE)

world_map <- ne_countries(
  scale = "medium", returnclass = "sf") |>
  filter(continent != "Antarctica")

world_map_plot <- ggplot() +
  geom_sf(data = world_map, fill = "grey80", color = "grey80", linewidth = 0.3) +
  geom_point(
    data = join_sed,
    aes(x = site_lon, y = site_lat, color = p95_zscore),
    size = 3,
    alpha = 0.85,
    shape = 16
  ) +
  scale_color_distiller(
    name = "SSC Z-Score",
    palette = "Spectral",
    limits = c(-std_dev_sed, std_dev_sed),
    oob = scales::squish
  ) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(-60, 85), expand = FALSE) +

  theme_void(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),

    # Legend placement and sizing
    legend.position = "top",
    legend.justification = "center",
    legend.box = "vertical",

    # Text sizing
    legend.title = element_text(face = "bold", size = 20),
    legend.text  = element_text(size = 20),

    # Spacing
    legend.margin = margin(t = 12, b = 12),
    legend.spacing.x = unit(1.5, "cm"),

    # Give the plot a bit more breathing room
    plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
  ) +
  guides(
    color = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth  = unit(0.4, "npc"),   # spans most of plot width
      barheight = unit(0.4, "cm"),
      direction = "horizontal"
    )
  )

world_map_plot

ggsave(
  filename = here("plots_2.0", "supplemental_ssc_map.png"),
  plot = world_map_plot,
  width = 16,
  height = 8,
  units = "in",
  dpi = 300
)
```
### Scatter Plot of Discharge and SSC

```{r}

scatter_df <- join_dis %>%
  inner_join(
    join_sed %>%
      select(site_name, p95_zscore_sed = p95_zscore),by = "site_name") |>
  filter(!is.na(p95_zscore), !is.na(p95_zscore_sed))

p_scatter <- ggplot(scatter_df, aes(x = p95_zscore, y = p95_zscore_sed, color = av_type)) +
  geom_point(
    size = 3,
    alpha = 0.7) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "gray30",
    linewidth = 1.2) +
   geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "gray30",
    linewidth = 1.2) +
  theme_minimal() +
  scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
  labs(x = "Discharge Z-Score (-)", y = "SSC Z-Score (-)") +
  theme(
    axis.title.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.title.x = element_text(color = "grey30", size = 20, face = "bold"),
    
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    
    axis.text.x  = element_text(color = "grey30", size = 16, face = "bold"),
    axis.text.y = element_text(color = "grey30", size = 16, face = "bold"),
    
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
    axis.ticks.x = element_line(color = "grey30", size = 1.2),
    
    legend.position = "none",
    
  plot.background = element_rect(fill = "transparent", colour = NA) # Transparent plot bg
  )
p_scatter

ggsave(
  filename = here("plots_2.0", "supplemental_scatter.png"),
  plot = p_scatter,
  width = 4,
  height = 4,
  units = "in",
  dpi = 300
)

```