---
title: "Templates for All Maps and Graphs + Descriptive Statistics"
description: "This document is meant to act as the location where all final version graphs are meant to be created, and all descriptive statistics put onto those graphs will be calcualted. This synthesizes material from other documents up to this point. "
author: "Sam Lance"
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
    theme: Lux
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r, include =FALSE}
library(dplyr)
library(tidyverse)
library(lubridate)
library(readr)
library(here)
library(ggplot2)
library(stringr)
library(purrr)
library(maps)
library(knitr)
library(kableExtra)
library(sf)
library(RColorBrewer)
library(patchwork)
library(tmap)
library(viridis)
library(sp)
library(rworldmap)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(devtools)
library(calecopal)
```

## Reading In + Cleaning Data

### Reading in Data

```{r, output=FALSE}}

join_dis <- read_csv(here("data","discharge_av_year_z.csv")) |>
  inner_join(
    read_csv(here("data","all_site_data_102825.csv")) |>
      select(site_name, site_lat, site_lon),   # no av_type
    by = "site_name") |>
  drop_na()

```

### Cleaning Data

```{r}
#Convert DF to SF Object 
sf_dis <- st_as_sf(join_dis, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_dis, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_dis <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

```

## Figure 1 - Inset Maps for Selected Sites

### Establishing Colors and Shapes for All Maps

```{r}
#Loading in World Map Data
world_map <- ne_countries(scale = "medium", returnclass = "sf")

#Establish Legend Breaks, Colors, and Desired Shapes
colors <- plasma(4)
rev_colors <- colors
shapes <- c(21, 22, 23, 24)

#devtools::install_github("an-bui/calecopal")# filled shapes
# breaks <- c(-2, -1, 0, 1, 2, 3, 4, 5)
# colors <- brewer.pal(length(breaks) - 1, "RdYlBu")
# rev_colors <- rev(colors)  # Reverse the color order
# shapes <- c(16, 17, 15, 18)  # Different shapes for avulsion types

```

### World Map

```{r}

selected_sites <- sf_dis |>
  filter(site_name %in% c("B9", "R3.2", "JA.25.India.1990", "MANAMBOLO.2004"))

# Make sure av_type is a factor and levels are consistent
sf_dis$av_type <- factor(sf_dis$av_type)

# Create a named color vector for av_type
av_type_colors <- setNames(rev_colors, levels(sf_dis$av_type))

world_map_plot <- world_map |>
  ggplot() +
  geom_sf(fill = "grey80", color = "grey80") +
  
  # All sites: filled by av_type, NO outline (color = NA)
  geom_point(
    data = sf_dis,
    aes(
      x = site_lon,
      y = site_lat,
      fill = av_type,
      shape = av_type,
      color = av_type
    ),
    size = 3,
    alpha = 0.7         # no outline
  ) +
  
  # Selected sites: same fill, but yellow outline
  geom_point(
    data = selected_sites,
    aes(
      x = site_lon,
      y = site_lat,
      fill = av_type,
      shape = av_type
    ),
    size = 3,
    stroke = 2,
    alpha = 0.7,
    color = "black",     # yellow outline only for selected sites
    show.legend = FALSE   # <<< this removes yellow outlines from legend
  ) +
   scale_color_manual(
    name = "Avulsion Type",
    values = av_type_colors
  ) +
  scale_fill_manual(
    name = "Avulsion Type",
    values = av_type_colors
  ) +
  scale_shape_manual(
    name = "Avulsion Type",
    values = shapes
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    
    legend.position = c(0.2, 0.25),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    legend.spacing.y = unit(-0.1, "cm"),
    legend.key.height = unit(0.3, "lines"),
    legend.background = element_rect(fill = "white", color = "white")
  ) +
  guides(
    fill  = guide_legend(byrow = TRUE),
    shape = guide_legend(byrow = TRUE)
  )

world_map_plot

ggsave(
  filename = here("plots", "fig1_world.png"),
  plot = world_map_plot,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300
)

```

### Robinson Projection World Map

```{r}

library(tidyverse)
library(sf)
library(ggplot2)

# -------------------------
# Inputs assumed to exist:
# world_map (sf polygons, lon/lat)
# sf_dis (dataframe with site_lon, site_lat, site_name, av_type)
# rev_colors (vector of colors)
# shapes (named vector of shapes for av_type)
# -------------------------

# 1) Select sites to highlight
selected_sites <- sf_dis |>
  filter(site_name %in% c("B9", "R3.2", "JA.25.India.1990", "MANAMBOLO.2004"))

# 2) Ensure factor levels consistent
sf_dis <- sf_dis |>
  mutate(av_type = factor(av_type))

# 3) Named color vector for av_type
av_type_colors <- setNames(rev_colors, levels(sf_dis$av_type))

# 4) Robinson CRS
crs_robin <- "+proj=robin +lon_0=0 +datum=WGS84 +units=m +no_defs"

# 5) Make points sf (critical for correct projection)
sf_dis_sf <- st_as_sf(
  sf_dis,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

selected_sites_sf <- st_as_sf(
  selected_sites,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

# 6) Ensure world_map has CRS (if not already)
if (is.na(st_crs(world_map))) st_crs(world_map) <- 4326

# 7) Create a "globe" outline and optional graticule
globe_outline <- st_as_sfc(
  st_bbox(c(xmin = -180, ymin = -90, xmax = 180, ymax = 90), crs = st_crs(4326))
) |>
  st_transform(crs_robin)

# Optional graticule
grat <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = seq(-60, 60, by = 30),
  crs = st_crs(4326)
) |>
  st_transform(crs_robin)

# -------------------------
# NEW: Top + bottom boundary (latitude caps)
# -------------------------
lat_min <- -90
lat_max <- 90
lon_seq <- seq(-180, 180, by = 0.25)

make_parallel <- function(lat) {
  st_sfc(
    st_linestring(cbind(lon_seq, rep(lat, length(lon_seq)))),
    crs = 4326
  ) |>
    st_transform(crs_robin)
}

bottom_boundary <- make_parallel(lat_min)
top_boundary    <- make_parallel(lat_max)

# 8) Build plot (Robinson + globe boundary)
world_map_plot <- ggplot() +

  # Light graticule behind everything (optional; remove if you don’t want lines)
  geom_sf(data = grat, color = "grey70", linewidth = 0.25) +

  # Continents
  geom_sf(
    data = world_map,
    fill = "grey70",
    color = "grey60",
    linewidth = 0.25
  ) +

  # All sites
  geom_sf(
    data = sf_dis_sf,
    aes(fill = av_type, shape = av_type, color = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 0.5
  ) +

  # Selected sites: same fill/shape, red outline
  geom_sf(
    data = selected_sites_sf,
    aes(fill = av_type, shape = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 1.0,
    color = "black",
    show.legend = FALSE
  ) +

  # -------------------------
  # NEW: draw top/bottom boundary lines
  # -------------------------
  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  # Globe outline (draw last so it frames everything)
  geom_sf(
    data = globe_outline,
    fill = NA,
    color = "grey70",
    linewidth = 0.7
  ) +

  # Scales
  scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +

  # Robinson projection
  coord_sf(crs = crs_robin, expand = FALSE) +

  # Styling
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    
    #legend.position = "none",
    legend.position = c(0.2, 0.25),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    legend.spacing.y = unit(-0.1, "cm"),
    legend.key.height = unit(0.3, "lines"),
    legend.background = element_rect(fill = "white", color = "white")
  ) +
  guides(
    fill  = guide_legend(byrow = TRUE),
    shape = guide_legend(byrow = TRUE)
  )

world_map_plot

ggsave(
  filename = here("plots", "fig1_world_robinson_globe.png"),
  plot = world_map_plot,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300
)

```

### jPacific Islands

```{r}

pacific <- world_map %>%
  filter(name %in% c("Timor-Leste", "Indonesia", "Papua New Guinea"))

pacific_av <- sf_dis |>
  filter(country %in% c("Timor-Leste", "Indonesia", "Papua New Guinea"))

selected_sites <- sf_dis |>
  filter(site_name %in% c("R3.2"))

# Make sure av_type is a factor and levels are consistent

pacific_av$av_type <- factor(pacific_av$av_type)

# Create a named color vector for av_type
av_type_colors <- setNames(rev_colors, levels(pacific_av$av_type))

# Make sure both layers are in the same CRS (assumes world_map is lon/lat WGS84)
pacific_sf <- st_transform(pacific, 4326)

bb <- st_bbox(pacific_sf)

# padding in degrees (tweak as needed)
pad_x <- 2
pad_y <- 2

pacific_plot <- pacific_sf |>
  ggplot() +
  geom_sf(fill = "white", color = "white") +
  geom_point(
    data = pacific_av,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type, color = av_type),
    size = 5, alpha = 0.7, show.legend = FALSE 
  ) +
  geom_point(
    data = selected_sites,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type),
    size = 4, stroke = 1.5, color = "yellow", show.legend = FALSE
  ) +
  scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    axis.title = element_blank(),
    axis.text  = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA), # Transparent panel background
    plot.background = element_rect(fill = "transparent", colour = NA)  # Transparent plot background
    # legend.position = c(0.2, 0.25),
    # legend.direction = "vertical",
    # legend.box = "vertical",
    # legend.title = element_text(face = "bold"),
    # legend.title.align = 0.5,
    # legend.spacing.y = unit(-0.1, "cm"),
    # legend.key.height = unit(0.3, "lines"),
    # legend.background = element_rect(fill = "white", color = "white")
  ) +
  coord_sf(
    xlim = c(bb["xmin"] - pad_x, bb["xmax"] + pad_x),
    ylim = c(bb["ymin"] - pad_y, bb["ymax"] + pad_y),
    expand = FALSE
  ) 
  # guides(
  #   fill  = guide_legend(byrow = TRUE),
  #   shape = guide_legend(byrow = TRUE)
  # )

pacific_plot

ggsave(
  filename = here("plots", "pacific_fig1.png"),
  plot = pacific_plot,
  width = 10, height = 7, units = "in", dpi = 300
)

```

### South America

```{r}
south_america_countries <- c(
  "Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador",
  "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela",
  "French Guiana"
)

sa_map <- world_map %>%
  filter(name %in% south_america_countries)

sa_av <- sf_dis %>%
  filter(country %in% south_america_countries)

selected_sites <- sf_dis %>%
  filter(site_name %in% c("B9.a"))

# Make sure av_type is a factor and levels are consistent
sa_av$av_type <- factor(sa_av$av_type)

# Create a named color vector for av_type
av_type_colors <- setNames(rev_colors, levels(sa_av$av_type))

# Make sure both layers are in the same CRS (assumes world_map is lon/lat WGS84)
sa_sf <- st_transform(sa_map, 4326)

bb <- st_bbox(sa_sf)

# padding in degrees (tweak as needed)
pad_x <- 2
pad_y <- 2

sa_plot <- sa_sf |>
  ggplot() +
  geom_sf(fill = "white", color = "white") +
  geom_point(
    data = sa_av,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type, color = av_type),
    size = 5, alpha = 0.7, show.legend = FALSE
  ) +
  geom_point(
    data = selected_sites,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type),
    size = 4, stroke = 1.5, color = "yellow", show.legend = FALSE
  ) +
  scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    axis.title = element_blank(),
    axis.text  = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA), # Transparent panel background
    plot.background = element_rect(fill = "transparent", colour = NA)  # Transparent plot background
  ) +
  coord_sf(
    xlim = c(bb["xmin"] - pad_x, bb["xmax"] + pad_x),
    ylim = c(bb["ymin"] - pad_y, bb["ymax"] + pad_y),
    expand = FALSE
  )

sa_plot

ggsave(
  filename = here("plots", "south_america_fig1.png"),
  plot = sa_plot,
  width = 10, height = 7, units = "in", dpi = 300
)



```

### Madagascar

```{r}
madagascar_names <- c("Madagascar")

mg_map <- world_map %>%
  filter(name %in% madagascar_names)

mg_av <- sf_dis %>%
  filter(country %in% madagascar_names)

selected_sites <- sf_dis %>%
  filter(site_name %in% c("MANAMBOLO.2004"))  # change if your Madagascar site differs

# Ensure av_type factor + consistent levels
mg_av$av_type <- factor(mg_av$av_type)

av_type_colors <- setNames(rev_colors, levels(mg_av$av_type))

# CRS + bbox for zoom
mg_sf <- st_transform(mg_map, 4326)
bb <- st_bbox(mg_sf)

pad_x <- 1
pad_y <- 1

mg_plot <- mg_sf |>
  ggplot() +
  geom_sf(fill = "white", color = "white") +
  geom_point(
    data = mg_av,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type, color = av_type),
    size = 5, alpha = 0.7, show.legend = FALSE
  ) +
  geom_point(
    data = selected_sites,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type),
    size = 4, stroke = 1.5, color = "yellow", show.legend = FALSE
  ) +
  scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    axis.title = element_blank(),
    axis.text  = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA), # Transparent panel background
    plot.background = element_rect(fill = "transparent", colour = NA)  # Transparent plot background
  ) +
  coord_sf(
    xlim = c(bb["xmin"] - pad_x, bb["xmax"] + pad_x),
    ylim = c(bb["ymin"] - pad_y, bb["ymax"] + pad_y),
    expand = FALSE
  )

mg_plot

ggsave(
  filename = here("plots", "madagascar_fig1.png"),
  plot = mg_plot,
  width = 10, height = 7, units = "in", dpi = 300
)


```

### India Nepal

```{r}
india_nepal <- c("India", "Nepal")

in_np_map <- world_map %>%
  filter(name %in% india_nepal)

in_np_av <- sf_dis %>%
  filter(country %in% india_nepal)

selected_sites <- sf_dis %>%
  filter(site_name %in% c("JA.8.India.2004.2004"))  # change if your selected site differs

# Ensure av_type factor + consistent levels
in_np_av$av_type <- factor(in_np_av$av_type)
av_type_colors <- setNames(rev_colors, levels(in_np_av$av_type))

# CRS + bbox for zoom
in_np_sf <- st_transform(in_np_map, 4326)
bb <- st_bbox(in_np_sf)

pad_x <- 2
pad_y <- 2

in_np_plot <- in_np_sf |>
  ggplot() +
  geom_sf(fill = "white", color = "white") +
  geom_point(
    data = in_np_av,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type, color = av_type),
    size = 5, alpha = 0.7, show.legend = FALSE
  ) +
  geom_point(
    data = selected_sites,
    aes(x = site_lon, y = site_lat, fill = av_type, shape = av_type),
    size = 4, stroke = 1.5, color = "yellow", show.legend = FALSE
  ) +
  scale_color_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_fill_manual(name = "Avulsion Type", values = av_type_colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    axis.title = element_blank(),
    axis.text  = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA), # Transparent panel background
    plot.background = element_rect(fill = "transparent", colour = NA)  # Transparent plot background
  ) +
  coord_sf(
    xlim = c(bb["xmin"] - pad_x, bb["xmax"] + pad_x),
    ylim = c(bb["ymin"] - pad_y, bb["ymax"] + pad_y),
    expand = FALSE
  )

in_np_plot

ggsave(
  filename = here("plots", "india_nepal_fig1.png"),
  plot = in_np_plot,
  width = 10, height = 7, units = "in", dpi = 300
)


```

## Figure 3 - Map of Discharge Z-Score

### Establishing Baseline

```{r}
#Loading in World Map Data
world_map <- ne_countries(scale = "medium", returnclass = "sf")

#Establish Legend Breaks, Colors, and Desired Shapes
breaks <- c(-2, -1, 0, 1, 2, 3, 4, 5)
colors <- brewer.pal(length(breaks) - 1, "PiYG")
rev_colors <- rev(colors)  # Reverse the color order
shapes <- c(16, 17, 15, 18)  # Different shapes for avulsion types
sf_dis <- sf_dis |>
  mutate(zscore_category = cut(p95_zscore, breaks = breaks, include.lowest = TRUE))

```

### South America

```{r}
south_america <- world_map |>
  filter(continent == "South America")

south_america_av <- sf_dis|>
  filter(continent == "South America")   # only if av also has continent

south_america_map <- south_america %>%
  ggplot() +
  geom_sf(fill = "grey35", color = "white") +
  geom_point(
    data = south_america_av,
    aes(x = site_lon, y = site_lat, fill = zscore_category),
    shape = 21,
    color = "black",
    size = 5
  ) +
  scale_fill_manual(
    name = "Discharge Z-Score",
    values = rev_colors,
    drop = FALSE,
    guide = guide_legend(order = 1, title.position = "top")
  ) +
  theme_minimal(base_size = 14) +
  labs(title = "South America") +
  theme(
    plot.title = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    legend.justification = c(0, 0.5),
    legend.background = element_rect(fill = "white", color = "white"),
    legend.title.align = 0.5,
    legend.text.size = 10,
    legend.title.size = 12) +
  coord_sf(
    xlim = range(south_america_av$site_lon, na.rm = TRUE) + c(-10, 40),
    ylim = range(south_america_av$site_lat, na.rm = TRUE) + c(-40, 10),
    expand = FALSE
  )

south_america_map

ggsave(
  filename = here("plots", "dis_south_america.png"),
  plot = south_america_map,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300
)

```

### Oceania

```{r}

pacific <- world_map %>%
  filter(name %in% c("Papua New Guinea", "Timor Leste", " Philippines", "Indonesia"))

pacific_av <- av |>
  filter(country %in% c("Papua New Guinea", "Timor Leste", " Philippines", "Indonesia"))

pacific_map <- pacific |>
  ggplot() +
  geom_sf(fill = "grey35", color = "white") +
  geom_point(data = pacific_av, 
             aes(x = site_lon, 
                 y = site_lat, 
                 fill = zscore_category), 
             size = 7,
             shape = 21,
             color = "black") +
  scale_fill_manual(name = "Discharge Z-Score", 
                     values = rev_colors, 
                     drop = FALSE,
                     guide = guide_legend(order = 1, title.position = "top")) +
  theme_minimal(base_size = 14) +
  labs(title = "Oceania") +
  theme(
    plot.title = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none") +
  coord_sf(xlim = range(pacific_av$site_lon) + c(-10, 5),
           ylim = range(pacific_av$site_lat) + c(-10, 5), 
           expand = FALSE)
pacific_map

ggsave(
      filename = here("plots", "dis_oceania.png"),
      plot = pacific_map,
      width = 10,
      height = 7,
      units = "in", # or "cm", "mm", "px"
      dpi = 300 # Adjust resolution as needed
    )

```

### Madagascar

```{r}

madagascar <- world_map %>%
  filter(name == "Madagascar")

madagascar_av <- av |>
  filter(country %in% c("Madagascar"))

madagascar_map <- madagascar |>
  ggplot() +
  geom_sf(fill = "grey35", color = "white") +
  geom_point(data = madagascar_av, 
             aes(x = site_lon, 
                 y = site_lat, 
                 fill = zscore_category), 
             size = 7,
             color = "black",
             shape = 21) +
  scale_fill_manual(name = "Discharge Z-Score", 
                     values = rev_colors, 
                     drop = FALSE) +
  theme_minimal(base_size = 18) +
  labs(title = "Madagascar") +
  theme(
    plot.title = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    legend.background = element_rect(fill = "white", color = "white")) +
  coord_sf(xlim = range(madagascar_av$site_lon) + c(-5, 12),
           ylim = range(madagascar_av$site_lat) + c(-5, 5), 
           expand = FALSE)
madagascar_map

ggsave(
      filename = here("plots", "dis_madagascar.png"),
      plot = madagascar_map,
      width = 10,
      height = 7,
      units = "in", # or "cm", "mm", "px"
      dpi = 300 # Adjust resolution as needed
    )

```

### Himalaya

```{r}

himalaya <- world_map %>%
  filter(name %in% c("India", "Nepal", "Bhutan", "Bangladesh", "Myanmar"))

himalaya_av <- av |>
  filter(country %in% c("India", "Nepal", "Bhutan", "Bangladesh", "Myanmar"))

himalaya_map <- himalaya |>
  ggplot() +
  geom_sf(fill = "gray35", color = "white") +
  geom_point(data = himalaya_av, 
             aes(x = site_lon, 
                 y = site_lat, 
                 fill = zscore_category), 
             size = 7,
             shape = 21,
             color = "black") +
  scale_fill_manual(name = "Discharge Z-Score", 
                     values = rev_colors, drop = FALSE) +
  theme_minimal(base_size = 14) +
  labs(title = "Himalaya Region") +
  theme(
    plot.title = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.title.align = 0.5,
    legend.background = element_rect(fill = "white", color = "white")) +
  coord_sf(xlim = range(himalaya_av$site_lon) + c(-15, 15),
           ylim = range(himalaya_av$site_lat) + c(-20, 5), 
           expand = FALSE)
himalaya_map

ggsave(
      filename = here("plots", "dis_himalaya.png"),
      plot = himalaya_map,
      width = 10,
      height = 7,
      units = "in", # or "cm", "mm", "px"
      dpi = 300 # Adjust resolution as needed
    )
```

### World Map for Legend

```{r}
#Set Up Map Extent
xmin <- -170
xmax <- 170

#Find Standard Deviation for the Dataset
#Purpose: use for the scale to condense and make a little easier to visualize 
std_dev_dis <- sd(join_dis$p95_zscore, na.rm = TRUE)

world_map <- ne_countries(scale = "medium", returnclass = "sf") |>
  filter(continent != "Antarctica")

world_map_plot <- ggplot() +
  geom_sf(data = world_map, fill = "grey80", color = "grey80", linewidth = 0.3) +
  geom_point(
    data = join_dis,
    aes(x = site_lon, y = site_lat, color = p95_zscore),
    size = 3,
    alpha = 0.85,
    shape = 16
  ) +
  scale_color_distiller(
    name = "Discharge Z-Score",
    palette = "Spectral",
    limits = c(-0.8, 0.8),
    oob = scales::squish
  ) +
  coord_sf(xlim = c(xmin, xmax), ylim = c(-60, 85), expand = FALSE) +

  theme_void(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA),

    # Legend placement and sizing
    legend.position = "top",
    legend.justification = "center",
    legend.box = "vertical",

    # Text sizing
    legend.title = element_text(face = "bold", size = 20),
    legend.text  = element_text(size = 20),

    # Spacing
    legend.margin = margin(t = 12, b = 12),
    legend.spacing.x = unit(1.5, "cm"),

    # Give the plot a bit more breathing room
    plot.margin = margin(t = 15, r = 10, b = 10, l = 10)
  ) +
  guides(
    color = guide_colorbar(
      title.position = "top",
      title.hjust = 0.5,
      barwidth  = unit(0.4, "npc"),   # spans most of plot width
      barheight = unit(0.4, "cm"),
      direction = "horizontal"
    )
  )

world_map_plot

ggsave(
  filename = here("plots", "world_dis_fig3.png"),
  plot = world_map_plot,
  width = 16,
  height = 8,
  units = "in",
  dpi = 300
)


```

## Figure 4 - Boxplots of Discharge Z-Score

### Boxplots of Z-Scores by Type

```{r}

#Establishing Title
title <- "Z-Score of Peak Discharge for Avulsion Year and Year Prior by Avulsion Type"

#Computing T-Tests
fan = join_dis |>
  filter(av_type == "fan")

delta = join_dis |>
  filter(av_type == "delta")

il = join_dis |>
  filter(av_type == "intra-lobe")

retro = join_dis |>
  filter(av_type == "retrogradational")


t_fan = t.test(fan$p95_zscore, mu = 0)
t_delta = t.test(delta$p95_zscore, mu = 0)
t_il = t.test(il$p95_zscore, mu = 0)
t_retro = t.test(retro$p95_zscore, mu = 0)
t_all = t.test(join_dis$p95_zscore, mu = 0)

print(t_all)

#Setting Up Plot Base
gardner_type <- join_dis |>
  ggplot(aes(x = av_type, y = p95_zscore, fill = av_type)) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 2, color = "gray30") +
  geom_boxplot(show.legend = FALSE, alpha = 0.5) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, color = scales::alpha("red", 0.50)) + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", width = 0.5) +
  labs(
       x = "Avulsion Type", y = "Discharge Z-Score (-)") +
  theme_classic(base_size = 18) +
  scale_fill_brewer(palette = "Greys") +  
  theme(
    plot.title = element_text(size = 22, face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
print(gardner_type)

ggsave(
  filename = here("plots", "dis_zscore_type_box.jpg"), 
  plot = gardner_type, 
  device = "jpg",
  height = 8, width = 7, units = "in"
)

```

### Boxplots of Z-Scores by Region

```{r}

#Establishing Title
title <- "Z-Score of Peak SSC for Avulsion Year and Year Prior by Site Area"

#Setting Up Plot Base
gardner_area <- av |>
  ggplot(aes(x = continent, y = p95_zscore, fill = continent)) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 2, color = "gray30") +
  geom_boxplot(show.legend = FALSE, alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(title = str_wrap(title, width = 40), 
       x = "Site Area", y = "Z-Score (-)") +
  scale_x_discrete(labels = label_wrap(width = 10)) +
  theme_minimal(base_size = 18) +
  scale_fill_brewer(palette = "RdPu") +  
  theme(
    plot.title = element_text(size = 22, face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )
print(gardner_area)

ggsave(
  filename = here("plots", "dis_zscore_continent_box.jpg"), 
  plot = gardner_area, 
  device = "png",
  height = 10, width = 7, units = "in"
)

```

### Boxplot Overall

```{r}

#Create boxplot of all discharge 

overall_dis <- join_dis |>
  ggplot(aes(x = "", y = p95_zscore)) +

  # raw data
  geom_jitter(
    width = 0.15,
    alpha = 0.6,
    size = 2,
    color = "gray30"
  ) +

  # boxplot
  geom_boxplot(
    width = 0.35,
    fill = "grey70",
    alpha = 0.6,
    outlier.shape = NA
  ) +

  # zero reference
  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    color = "grey30",
    linewidth = 1.2
  ) +

  # mean (solid red line)
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.35,
    color = scales::alpha("red", 0.50),
    linewidth = 0.5
  ) +

  labs(
    x = NULL,
    y = "Discharge Z-Score (-)"
  ) +

  theme_classic(base_size = 18) +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    panel.background = element_rect(fill = "transparent", colour = NA), # Transparent panel bg
    axis.line.x = element_line(color = "grey30", linewidth = 1.2),
    axis.line.y = element_line(color = "grey30", linewidth = 1.2),
    axis.text.y = element_text(color = "grey30", size = 20, face = "bold"),
    axis.ticks.y = element_line(color = "grey30", size = 1.2),
  plot.background = element_rect(fill = "transparent", colour = NA) # Transparent plot bg
  )

overall_dis

ggsave(
  filename = here("plots", "discharge_overall_box.png"), 
  bg = "transparent",
  plot = overall_dis, 
  device = "png",
  height = 4, width = 4, units = "in"
)


```

## Making Mock Time Series

```{r}

set.seed(123)

# ---- Create daily time series ----
years <- 1985:2020
days_per_year <- 365

df <- expand.grid(
  year = years,
  day = 1:days_per_year
) %>%
  mutate(
    # decimal year (numeric x-axis)
    x_year = year + (day - 1) / days_per_year,

    # baseline low flow
    baseflow = rgamma(n(), shape = 1.2, scale = 120),

    # flood season probability (mid-year peak)
    flood_prob = dnorm(day, mean = 200, sd = 35),
    flood_prob = flood_prob / max(flood_prob),

    # random flood occurrence
    flood_event = rbinom(n(), 1, prob = 0.04 * flood_prob),

    # flood magnitude (heavy-tailed)
    flood_mag = flood_event * rgamma(n(), shape = 2.5, scale = 2200),

    discharge = baseflow + flood_mag
  )

avulsion_year <- 2004

mean_p95 <- quantile(df$discharge, probs = 0.95)


p <- ggplot(df, aes(x = x_year, y = discharge)) +

  # Avulsion year & prior shading
  annotate(
    "rect",
    xmin = avulsion_year - 1,
    xmax = avulsion_year,
    ymin = -Inf,
    ymax = Inf,
    fill = "yellow",
    alpha = 0.3
  ) +

  # Discharge time series
  geom_line(color = "#3b73a1", linewidth = 0.5) +

  # Mean 95th percentile line
  geom_hline(
    yintercept = mean_p95,
    linetype = "dashed",
    color = "grey40"
  ) +

  scale_x_continuous(
    limits = c(1985, 2020),
    breaks = seq(1985, 2020, by = 5)
  ) +

  labs(
    title = "C",
    x = "Year",
    y = expression("Discharge (m"^3*"/s)")
  ) +

  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(
      linetype = "dashed",
      linewidth = 0.4
    )
  )

print(p)


```
