---
title: "knn_cluster"
format: html
editor: visual
---
```{r}
library(sf)
library(dplyr)
library(spdep)
library(ggplot2)
library(janitor)
library(tidyverse)
library(here)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(RColorBrewer)
library(ggnewscale)
library(terra)
library(tidyterra)
library(spatialreg)
```

## Load and Join Coordinates

```{r}
#Load Data in OG Format No Modifications 
site_xy_og <- read_csv(here("data","base_sites.csv"), show_col_types = FALSE) 
dis_z <- read_csv(here("data","discharge_av_year_z.csv"), show_col_types = FALSE)
sed_z <- read_csv(here("data","ssc_av_year_z.csv"), show_col_types = FALSE)

#Load Avulsion Data
site_xy <- read_csv(here("data","base_sites.csv"), show_col_types = FALSE) |>
  select(site_name, site_lat, site_lon) |>
  distinct() |>
  drop_na()

#Load Discharge Z-Score Data + Convert to SF Object 
dis_sf <- read_csv(here("data","discharge_av_year_z.csv"), show_col_types = FALSE) |>
  inner_join(site_xy, by = "site_name") |> #need lat lon from xy data, so join 
  drop_na(p95_zscore, site_lat, site_lon) |> #get rid of NA values 
  st_as_sf(coords = c("site_lon","site_lat"), crs = 4326) |> #convert to sf object 
  st_transform(6933) #transform to equal area projection

#Load Sediment Z-Score Data + Convert to SF Object 
sed_sf <- read_csv(here("data","ssc_av_year_z.csv"), show_col_types = FALSE) |>
   st_as_sf(coords = c("site_lon","site_lat"), crs = 4326) |>  #convert to sf object
   st_transform(6933) #transform to equal area projection 
```

## Create Functions 

### Jitter Function 

Purpose: jitter all points by 0,01 if there is an overlap in coordinates to avoid issues with spatial weights and clustering

```{r}
# ============================================================================
# JITTER FUNCTION
# ============================================================================
jitter_sf <- function(sf_obj, amount = 0.01) {
  coords     <- st_coordinates(sf_obj)
  coords[,1] <- jitter(coords[,1], amount = amount) #jitter first coord by 0.01
  coords[,2] <- jitter(coords[,2], amount = amount) #jitter second coord by 0.01
  st_geometry(sf_obj) <- st_as_sfc(
    lapply(1:nrow(coords), function(i) st_point(coords[i,])), 
    crs = st_crs(sf_obj)
  )
  sf_obj
}

```

### Assign Continent

```{r}
# ============================================================================
# ASSIGN CONTINENTS FUNCTION 
# ============================================================================

assign_continents <- function(sf_df) {
  
  #Seeing if Already a Continent Columm 
  if ("continent" %in% names(sf_df)) {
    cat("Data already has continent column. Skipping assignment.\n")
    return(sf_df)
  }
  
  #Transform to Right Coordinate System for Spatial Join (WGS84)
  sf_wgs <- st_transform(sf_df, 4326)
  
  #Load in World with Ne Countries + Transform to WGS 84 
  world <- ne_countries(scale = "medium", returnclass = "sf") %>%
    dplyr::select(continent) %>%
    st_make_valid() %>%
    st_transform(4326)

  #Intersect Join to Get Continents 
  sf_joined <- st_join(sf_wgs, world, join = st_intersects, left = TRUE)

  #Fallback for any NA Values 
  miss <- which(is.na(sf_joined$continent))
  if (length(miss) > 0) {
    nearest_idx <- st_nearest_feature(sf_joined[miss, ], world)
    sf_joined$continent[miss] <- world$continent[nearest_idx]
  }
  
  #Label Unknowns 
  sf_joined$continent <- tidyr::replace_na(sf_joined$continent, "Unknown")

  #Transform Back to OG CRS 
  st_transform(sf_joined, st_crs(sf_df))
}
```

### Region Delineation  

Purpose: override original continent assignment with more specific regions 

```{r}
# ============================================================================
# ASSIGN REGIONS 
# ============================================================================

assign_regions <- function(sf_obj) {
  
  #Transform Coordinates 
  coords <- st_coordinates(st_transform(sf_obj, 4326)) 
  
  #Create Column of Latitudes and Colun of Longitudes
  sf_obj <- sf_obj %>%
    mutate(
      lon = coords[,1],
      lat = coords[,2],
      region = case_when(
        
        #Northern South America (Colombia, Venezuela, Ecuador, Peru, Brazil north)
        continent == "South America" & lat > -5 ~ "Northern South America",
        
        #Southern South America (Chile, Argentina, Bolivia, Brazil south)
        continent == "South America" & lat <= -5 ~ "Southern South America",
        
        #Northern India / Nepal / Bangladesh region
        continent == "Asia" & lat >= 20 & lat <= 32 & lon >= 68 & lon <= 108 ~ "India-Nepal-Bangladesh",
        
        #Southeast Asia
        continent == "Asia" & lat <= 20 & lon >= 95 & lon <= 180 ~ "Southeast Asia",
        continent == "Oceania" & lat <= 20 & lon >= 95 & lon <= 180 ~ "Southeast Asia",

        #Central/East Asia (Tibetan Plateau, China, Mongolia)
        continent == "Asia" & lat > 30 & lon >= 70 ~ "Central-East Asia",

        #Middle East / Western Asia
        continent == "Asia" & lon < 70 ~ "Middle East",

        #Madagascar separate from mainland Africa
        continent == "Africa" & lon >= 43 & lon <= 51 & lat >= -26 & lat <= -12 ~ "Madagascar",
        
        #Mainland Africa
        # continent == "Africa" ~ "Mainland Africa",
        
        # Everything else keeps continent name
        TRUE ~ continent
      )
    )
}
```

### Finding Optimal K 

Methodology:
-   Fit a spatial error model for each region across a range of k values (2 to max_k or n/2)
-   Calculate AIC for each model using the SEM and check if the spatial weights graph is fully connected (1 subgraph)
-   Select the lowest AIC value among the connected graphs as the optimal k for that region 

```{r}
# ============================================================================
# FIND OPTIMAL K VIA AIC (SPATIAL ERROR MODEL)
# ============================================================================

find_optimal_k_by_region <- function(sf_obj, min_obs = 5, max_k = 15) {
  
  #Calling Previous Functions
  sf_obj <- jitter_sf(sf_obj, amount = 0.01) #jitter
  sf_obj <- assign_continents(sf_obj) #assigning continents 
  sf_obj <- assign_regions(sf_obj) #modifying continents to specific regions 
  
  #Select Regions of Interest 
  active_regions <- c("Northern South America", "Southern South America",
                      "India-Nepal-Bangladesh", "Southeast Asia")
  
  #Transform to WGS84 for Coordinate Extraction
  sf_plot    <- st_transform(sf_obj, 4326)
  coords_all <- st_coordinates(sf_plot)
  
  #Create Dataframe 
  #Selecting coordinates, zscore, and region only 
  df <- data.frame(
    x      = coords_all[,1],
    y      = coords_all[,2],
    zscore = sf_plot$p95_zscore,
    region = sf_plot$region) |> filter(region %in% active_regions)
  
  #Count the Number of Datapoints Per Region
  region_counts <- df |>
    count(region) |>
    filter(n >= min_obs) #ensure above a specific threshold min_obs 
  
  #Create Lists and Vectors to Store Results
  all_aic_data <- list()
  plots        <- list()
  optimal_ks   <- c()
  
  #For Loop to Run Through Each Region and Test K Values
  for (reg in region_counts$region) {
    df_reg <- df |> filter(region == reg) #pick a specific region 
    
    n_obs  <- nrow(df_reg) #number of observations in that region
    k_max  <- min(max_k, floor(n_obs / 2)) #set maximum k value 
    
    #Printing Progress Message 
    cat("\n--- Region:", reg, "| n =", n_obs, "| testing k = 2 to", k_max, "---\n") 
    
    #Extracts Coords as Matrix, Pulls Z-score, Combines as DF
    coords_reg <- as.matrix(df_reg[, c("x", "y")])
    z          <- df_reg$zscore
    df_model   <- data.frame(z = z)  # intercept-only model
    
    #Creates Results Table to Store AIC and Connectivity Info for Each K
    aic_df <- data.frame(
      k         = 2:k_max,
      aic       = NA_real_,
      connected = NA
    )
    
    #Actual Loop
    #Going through each value of k from 2 to k_max 
    for (i in seq_along(2:k_max)) {
      k_try <- (2:k_max)[i]
      
      #Building KNN
      tryCatch({
        knn <- knearneigh(coords_reg, k = k_try) #finds knn for each points 
        nb  <- knn2nb(knn) #converts to list of neighbors
        lw  <- nb2listw(nb, style = "W", zero.policy = TRUE) #creates spatial weights object
        
        #Checking if Fully Connected Graph (1 subgraph) or Multiple Subgraphs
        n_sub <- n.comp.nb(nb)$nc
        aic_df$connected[i] <- (n_sub == 1)
        
        #Fitting Spatial Error Model and Extracting AIC
        #z = 1 because only interested in intercept 
        #data = df_model because has zscores
        #listw = spatial weights from current k being tested
        #zeropolicy = True can try ones with 0 
        mod <- errorsarlm(z ~ 1, data = df_model, listw = lw, zero.policy = TRUE)
        
        #Extracts AIC from Model for Each K and Stores in Results Table
        aic_df$aic[i] <- AIC(mod)
        
        #Error Handler to Catch Any Issues with Model Fitting (e.g. disconnected graph, singularities)
      }, error = function(e) {
        cat("  Error at k =", k_try, ":", e$message, "\n")
      })
    }
    
    #Saves in a Dataframe 
    all_aic_data[[reg]] <- aic_df
    
    #Select Lowest K Value with Connected Graph and Valid AIC
    optimal_k <- aic_df |>
      filter(connected == TRUE, !is.na(aic)) |>
      slice_min(aic, n = 1) |>  #minimize aic 
      pull(k) #pull value of k and put into optimal_k 
    
    #Fallback Method
    #If no connected k found, use minimum connected k regardless of AIC
    if (length(optimal_k) == 0) {
      optimal_k <- aic_df |>
        filter(connected == TRUE) |>
        slice(1) |>
        pull(k)
      cat("  Warning: AIC failed for all connected k — using min connected k\n")
    }
    
    optimal_ks[reg] <- optimal_k
    
    cat("  Optimal k (min AIC, connected):", optimal_k, "\n")
    cat("  AIC at optimal k:", aic_df$aic[aic_df$k == optimal_k], "\n")
    
    #Plotting AIC Curve 
    #Allows to see progression over time with different k values 
    p <- ggplot(aic_df, aes(x = k, y = aic)) +
      geom_line(color = "grey50", linewidth = 0.8) +
      geom_point(aes(color = connected, shape = connected), size = 3) +
      geom_vline(xintercept = optimal_k, linetype = "dashed",
                 color = "#d7191c", linewidth = 0.8) +
      annotate("text",
               x     = optimal_k + 0.2,
               y     = max(aic_df$aic, na.rm = TRUE),
               label = paste("optimal k =", optimal_k),
               hjust = 0, color = "#d7191c", size = 3.5) +
      scale_color_manual(values = c("TRUE" = "#2c7bb6", "FALSE" = "#fdae61"),
                         labels = c("TRUE" = "Connected", "FALSE" = "Sub-graphs"),
                         name = NULL) +
      scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 17),
                         labels = c("TRUE" = "Connected", "FALSE" = "Sub-graphs"),
                         name = NULL) +
      labs(title    = paste("AIC-Based k Selection (Spatial Error) -", reg),
           subtitle = "AIC vs k | dashed line = optimal k",
           x        = "k (number of neighbors)",
           y        = "AIC") +
      theme_minimal(base_size = 14) +
      theme(
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "bottom"
      )
    
    plots[[reg]] <- p
  }
  
  summary_df <- data.frame(
    region    = names(optimal_ks),
    optimal_k = as.integer(optimal_ks)
  )
  
  cat("\n============================================\n")
  cat("Optimal k by region (AIC, spatial error model):\n")
  print(summary_df)
  cat("============================================\n")
  
  list(plots = plots, optimal_ks = summary_df, aic_data = all_aic_data)
}
```
### K Nearest Neighbors 

Purpose: define neighbors based on k nearest points, which is more appropriate for global datasets with uneven spatial distribution than distance-based neighbors. This function runs the full spatial analysis suite (Moran's I, LISA, GI*) for a given sf object and specified k. It returns the sf object with results and the global Moran's I test result for summary.

```{r}

# ============================================================================
# SPATIAL SUITE WITH KNN AND REGIONS
# ============================================================================

run_spatial_suite_single_knn <- function(sf_df, k_neighbors = 8, nsim = 999, alpha = 0.05) {
  
  #Pulls Z-Scores 
  z <- sf_df |>
    st_drop_geometry() |>
    pull(p95_zscore) |>
    as.numeric()
  
  #Extracts Coordinates 
  coords <- st_coordinates(sf_df)
  
  #KNN neighbor Definition
  knn <- knearneigh(coords, k = min(k_neighbors, nrow(sf_df) - 1))
  nb  <- knn2nb(knn)
  lw  <- nb2listw(nb, style = "W", zero.policy = TRUE)
  
  #Conducts Global Morans I
  #Uses z-score, the weights we established 
  moran_global <- moran.test(z, lw, zero.policy = TRUE)
  
  #Conducts Getis-Ord GI* 
  gi     <- as.numeric(localG(z, lw, zero.policy = TRUE)) #testing, convert to numeric 
  gi_p   <- 2 * pnorm(-abs(gi)) #remove directionalty, convert to a p-value using normal distribution, and two-tailed 
  gi_class <- rep("Not significant", length(gi)) #default assumption of not significant 
  gi_class[gi_p <= alpha & gi > 0] <- "Hotspot (high)" #when statistically significant AND greater than 0 hotspot 
  gi_class[gi_p <= alpha & gi < 0] <- "Coldspot (low)" #when statistically significant AND below 0 coldspot 
  
  #LISA
  lisa_raw <- localmoran_perm(z, lw, nsim = nsim, zero.policy = TRUE)
  
  get_col <- function(x, candidates) {
    nm  <- colnames(x)
    hit <- candidates[candidates %in% nm][1]
    if (is.na(hit)) stop("Missing expected columns. Available: ", paste(nm, collapse = ", "))
    x[, hit]
  }
  
  Ii     <- get_col(lisa_raw, c("statistic", "Ii")) #pull i stat 
  p      <- get_col(lisa_raw, c("p.value", "p_sim", "p.value.sim", "p.value.perm", "Pr(z != E(Ii))")) #pull p-value 
  lag_z  <- lag.listw(lw, z, zero.policy = TRUE)
  
  sig        <- p <= alpha #set signficance threshold 
  lisa_class <- rep("Not significant", length(z))
  lisa_class[sig & z > 0 & lag_z > 0] <- "High-High" #when statistically significant AND above 0 AND neighbors above 0 high-high cluster
  lisa_class[sig & z < 0 & lag_z < 0] <- "Low-Low" #when statistically significant AND below 0 AND neighbors below 0 low-low cluster
  lisa_class[sig & z > 0 & lag_z < 0] <- "High-Low" #when statistically significant AND above 0 AND neighbors below 0 high-low cluster
  lisa_class[sig & z < 0 & lag_z > 0] <- "Low-High" #when statistically significant AND below 0 AND neighbors above 0 low-high cluster
  
  #Compile Output into One Big Dataframe 
  sf_out <- sf_df |>
    mutate(
      z_std        = z,
      moran_Ii     = Ii,
      moran_p      = p,
      lag_z        = lag_z,
      lisa_cluster = lisa_class,
      gi_star      = gi,
      gi_p         = gi_p,
      gi_cluster   = gi_class
    )
  
  list(sf = sf_out, lw = lw, global_moran = moran_global)
}

```

### Function to Run Analysis by Region

```{r}
# ============================================================================
# WRAPPER FUNCTION BY REGION
# ============================================================================

run_spatial_suite_by_region <- function(sf_obj, k_neighbors = 4,
                                         k_by_region = NULL,
                                         nsim = 999, alpha = 0.05,
                                         min_obs = 5) {
  
  active_regions <- c("Northern South America", "Southern South America",
                      "India-Nepal-Bangladesh", "Southeast Asia", "Madagascar")
  
  sf_obj <- jitter_sf(sf_obj, amount = 0.01)
  sf_obj <- assign_continents(sf_obj)
  sf_obj <- assign_regions(sf_obj)
  
  region_counts <- sf_obj %>%
    st_drop_geometry() %>%
    filter(region %in% active_regions) %>%
    count(region) %>%
    filter(n >= min_obs)
  
  if (nrow(region_counts) == 0) {
    stop("No regions have >= ", min_obs, " observations.")
  }
  
  cat("Analyzing", nrow(region_counts), "region(s):\n")
  print(region_counts)
  
  results <- region_counts$region %>%
    setNames(., .) %>%
    lapply(function(reg) {
      cat("\n--- Analyzing:", reg, "---\n")
      
      reg_data <- sf_obj %>% filter(region == reg)
      n_obs    <- nrow(reg_data)
      
      # Use region-specific k if provided, otherwise fall back to default
      if (!is.null(k_by_region) && reg %in% names(k_by_region)) {
        k_use <- k_by_region[[reg]]
        cat("  Using optimal k =", k_use, "(from elbow method)\n")
      } else {
        k_use <- min(k_neighbors, n_obs - 1)
        cat("  Using default k =", k_use, "\n")
      }
      
      cat("  Locations:", n_obs, "| k =", k_use, "\n")
      
      tryCatch({
        result        <- run_spatial_suite_single_knn(reg_data,
                                                       k_neighbors = k_use,
                                                       nsim = nsim,
                                                       alpha = alpha)
        result$region <- reg
        result
      }, error = function(e) {
        warning("Error in ", reg, ": ", e$message)
        NULL
      })
    })
  
  results <- results[!sapply(results, is.null)]
  
  combined_sf <- do.call(rbind, lapply(results, function(x) x$sf))
  
  summary_table <- do.call(rbind, lapply(results, function(x) {
    data.frame(
      region      = x$region,
      n_locations = nrow(x$sf),
      k_used      = x$lw$neighbours %>% sapply(length) %>% mean() %>% round(1),
      moran_I     = round(x$global_moran$estimate[1], 3),
      moran_p     = round(x$global_moran$p.value, 4),
      n_hotspots  = sum(x$sf$gi_cluster == "Hotspot (high)"),
      n_coldspots = sum(x$sf$gi_cluster == "Coldspot (low)")
    )
  }))
  
  return(list(
    results_by_region = results,
    combined_sf       = combined_sf,
    summary           = summary_table
  ))
}
```


## Run Analysis

```{r}

dis_elbow <- find_optimal_k_by_region(dis_sf, min_obs = 5, max_k = 15)
#sed_elbow <- find_optimal_k_by_region(sed_sf, min_obs = 5, max_k = 15)

dis_elbow$plots
#sed_elbow$plots

dis_optimal_ks <- setNames(dis_elbow$optimal_ks$optimal_k,
                            dis_elbow$optimal_ks$region)
#sed_optimal_ks <- setNames(sed_elbow$optimal_ks$optimal_k,
                            # sed_elbow$optimal_ks$region)



# Run regional analysis using the region-specific optimal k values
dis_results_region <- run_spatial_suite_by_region(
  dis_sf,
  k_neighbors  = 4,              # fallback if region not in k_by_region
  k_by_region  = dis_optimal_ks,  # your named vector from elbow
  nsim         = 999,
  alpha        = 0.05,
  min_obs      = 5
)

# sed_results_region <- run_spatial_suite_by_region(
#   sed_sf,
#   k_neighbors  = 4,
#   k_by_region  = sed_optimal_ks,
#   nsim         = 999,
#   alpha        = 0.05,
#   min_obs      = 5
# )

```

## Mapping

### Establish World Map

```{r}

world_map <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(st_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON"))

```

### Global Morans I

```{r}

# --- 3. Global Moran's I Summary Bar Chart ---
plot_moran_summary_regions <- function(results_obj, title_text) {
  
  summary_df <- results_obj$summary %>%
    mutate(
      significant = moran_p <= 0.05,
      label       = paste0("I=", moran_I, "\np=", moran_p)
    )
  
  ggplot(summary_df, aes(x = reorder(region, moran_I), y = moran_I, 
                          fill = significant)) +
    geom_col(alpha = 0.85) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_text(aes(label = paste0("p=", moran_p)),
              hjust = -0.1, size = 3.5) +
    scale_fill_manual(values = c("TRUE"  = "#d7191c", 
                                  "FALSE" = "grey60"),
                      labels = c("TRUE"  = "Significant (p≤0.05)",
                                  "FALSE" = "Not significant"),
                      name = NULL) +
    coord_flip(clip = "off") +
    expand_limits(y = max(abs(summary_df$moran_I)) * 1.4) +
    labs(title    = title_text,
         subtitle = "Global Moran's I by region | red = significant clustering",
         x        = NULL,
         y        = "Moran's I") +
    theme_minimal(base_size = 14) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

```

### Getis Ord GI* 

```{r}

# ============================================================================
# MAPPING FUNCTIONS FOR DISCHARGE REGIONAL RESULTS
# ============================================================================

# ============================================================================
# Hotspots and Coldspots 
# ============================================================================
plot_gi_world_regions <- function(results_obj, title_text) {
  
  sf_plot <- st_transform(results_obj$combined_sf, 4326)
  coords  <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x          = coords[,1],
    y          = coords[,2],
    gi_cluster = sf_plot$gi_cluster
  )
  
  df$gi_cluster <- factor(df$gi_cluster, 
                           levels = c("Hotspot (high)", "Coldspot (low)", "Not significant"))
  
  ggplot() +
    geom_sf(data = world_map, fill = "grey85", color = NA, linewidth = 0) +
    geom_point(data = df %>% filter(gi_cluster == "Not significant"),
               aes(x = x, y = y), color = "grey40", size = 1.8, alpha = 0.6) +
    geom_point(data = df %>% filter(gi_cluster != "Not significant"),
               aes(x = x, y = y, color = gi_cluster), size = 3, alpha = 0.9) +
    scale_color_manual(values = c("Hotspot (high)" = "#d7191c", 
                                   "Coldspot (low)"  = "#2c7bb6"),
                       name = NULL) +
    coord_sf(xlim = c(-175, 175), ylim = c(-85, 85), expand = FALSE, datum = NA) +
    labs(title    = title_text,
         subtitle = "GI* hotspots and coldspots by region") +
    theme_minimal(base_size = 14) +
    theme(
      axis.title      = element_blank(),
      axis.text       = element_blank(),
      axis.ticks      = element_blank(),
      panel.grid      = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      legend.position = c(0.12, 0.25)
    )
}
```

### LISA 
```{r}

# ============================================================================
# Hotspots and Coldspots 
# ============================================================================
plot_lisa_world_regions <- function(results_obj, title_text) {
  
  sf_plot <- st_transform(results_obj$combined_sf, 4326)
  coords  <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x            = coords[,1],
    y            = coords[,2],
    lisa_cluster = sf_plot$lisa_cluster
  )
  
  df$lisa_cluster <- factor(df$lisa_cluster,
                             levels = c("High-High", "Low-Low", 
                                        "High-Low", "Low-High", "Not significant"))
  
  ggplot() +
    geom_sf(data = world_map, fill = "grey85", color = NA, linewidth = 0) + 
    geom_point(data = df %>% filter(lisa_cluster == "Not significant"),
               aes(x = x, y = y), color = "grey40", size = 1.8, alpha = 0.6) +
    geom_point(data = df %>% filter(lisa_cluster != "Not significant"),
               aes(x = x, y = y, color = lisa_cluster), size = 3, alpha = 0.9) +
    scale_color_manual(values = c("High-High" = "#d7191c",
                                   "Low-Low"   = "#2c7bb6",
                                   "High-Low"  = "#fdae61",
                                   "Low-High"  = "#abd9e9"),
                       name = NULL) +
    coord_sf(xlim = c(-175, 175), ylim = c(-85, 85), expand = FALSE, datum = NA) +
    labs(title    = title_text,
         subtitle = "LISA cluster classifications by region") +
    theme_minimal(base_size = 14) +
    theme(
      axis.title      = element_blank(),
      axis.text       = element_blank(),
      axis.ticks      = element_blank(),
      panel.grid      = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      legend.position = c(0.12, 0.25)
    )
}

```

### Print Maps 

```{r}
# ============================================================================
# DISPLAY MAPS
# ============================================================================

p_dis_gi_regions    <- plot_gi_world_regions(dis_results_region, 
                                              "Discharge GI* Hotspots/Coldspots by Region")

p_dis_lisa_regions  <- plot_lisa_world_regions(dis_results_region,
                                                "Discharge LISA Clusters by Region")

p_dis_moran_regions <- plot_moran_summary_regions(dis_results_region,
                                                   "Discharge Global Moran's I by Region")

p_dis_gi_regions
p_dis_lisa_regions
p_dis_moran_regions

# ============================================================================
# Save Maps to Plots
# ============================================================================

ggsave(here("plots", "dis_gi_regions.png"), plot = p_dis_gi_regions, width = 6, height = 4, units = "in", dpi = 300)


```

### Regional Maps

```{r}

# ============================================================================
# INDIVIDUAL REGION MAPS - GI* AND LISA
# ============================================================================

plot_spatial_by_region <- function(results_obj, title_prefix, save_folder = NULL) {
  
  plots <- list()
  
  for (reg in names(results_obj$results_by_region)) {
    
    sf_reg  <- st_transform(results_obj$results_by_region[[reg]]$sf, 4326)
    coords  <- st_coordinates(sf_reg)
    
    df <- data.frame(
      x            = coords[,1],
      y            = coords[,2],
      gi_cluster   = sf_reg$gi_cluster,
      lisa_cluster = sf_reg$lisa_cluster
    )
    
    xlim <- range(df$x) + c(-2, 2)
    ylim <- range(df$y) + c(-2, 2)
    
    # --- GI* MAP ---
    df$gi_cluster <- factor(df$gi_cluster,
                             levels = c("Hotspot (high)", "Coldspot (low)", 
                                        "Not significant"))
    
    p_gi <- ggplot() +
      geom_sf(data = world_map, fill = "grey85", color = NA, linewidth = 0) +
      geom_point(data = df %>% filter(gi_cluster == "Not significant"),
                 aes(x = x, y = y), color = "grey40", size = 2, alpha = 0.6) +
      geom_point(data = df %>% filter(gi_cluster != "Not significant"),
                 aes(x = x, y = y, color = gi_cluster), size = 3.5, alpha = 0.9) +
      scale_color_manual(values = c("Hotspot (high)" = "#d7191c",
                                     "Coldspot (low)"  = "#2c7bb6"),
                         name = NULL) +
      coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, datum = NA) +
      labs(title    = paste(title_prefix, "-", reg),
           subtitle = "GI* hotspots and coldspots") +
      theme_minimal(base_size = 14) +
      theme(
        axis.title      = element_blank(),
        axis.text       = element_blank(),
        axis.ticks      = element_blank(),
        panel.grid      = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right"
      )
    
    plots[[paste0(reg, "_gi")]] <- p_gi
    
    if (!is.null(save_folder)) {
      fname <- paste0(save_folder, "/", tolower(gsub(" ", "_", title_prefix)),
                      "_gi_", tolower(gsub(" ", "_", reg)), ".png")
      ggsave(fname, p_gi, width = 10, height = 7)
      cat("Saved:", fname, "\n")
    }
    
    # --- LISA MAP ---
    df$lisa_cluster <- factor(df$lisa_cluster,
                               levels = c("High-High", "Low-Low",
                                          "High-Low", "Low-High",
                                          "Not significant"))
    
    p_lisa <- ggplot() +
      geom_sf(data = world_map, fill = "grey85", color = NA, linewidth = 0) +
      geom_point(data = df %>% filter(lisa_cluster == "Not significant"),
                 aes(x = x, y = y), color = "grey40", size = 2, alpha = 0.6) +
      geom_point(data = df %>% filter(lisa_cluster != "Not significant"),
                 aes(x = x, y = y, color = lisa_cluster), size = 3.5, alpha = 0.9) +
      scale_color_manual(values = c("High-High" = "#d7191c",
                                     "Low-Low"   = "#2c7bb6",
                                     "High-Low"  = "#fdae61",
                                     "Low-High"  = "#abd9e9"),
                         name = NULL) +
      coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, datum = NA) +
      labs(title    = paste(title_prefix, "-", reg),
           subtitle = "LISA cluster classifications") +
      theme_minimal(base_size = 14) +
      theme(
        axis.title      = element_blank(),
        axis.text       = element_blank(),
        axis.ticks      = element_blank(),
        panel.grid      = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right"
      )
    
    plots[[paste0(reg, "_lisa")]] <- p_lisa
    
    if (!is.null(save_folder)) {
      fname <- paste0(save_folder, "/", tolower(gsub(" ", "_", title_prefix)),
                      "_lisa_", tolower(gsub(" ", "_", reg)), ".png")
      ggsave(fname, p_lisa, width = 10, height = 7)
      cat("Saved:", fname, "\n")
    }
  }
  
  plots
}

# ============================================================================
# RUN IT
# ============================================================================
dis_region_maps <- plot_spatial_by_region(dis_results_region,
                                           title_prefix = "Discharge",
                                           save_folder  = here("lisa_plots"))

dis_region_maps

```

## Code Graveyard

### Elbow Method - NOT WORKING

### Finding the Ideal K

PROBLEM: this might not be working, elbows not actually elbows 

```{r}

# find_optimal_k_by_region <- function(sf_obj, min_obs = 5, max_k = 15) {
#   
#   # Jitter and assign regions
#   sf_obj <- jitter_sf(sf_obj, amount = 0.01)
#   sf_obj <- assign_continents(sf_obj) #call function above 
#   sf_obj <- assign_regions(sf_obj) #call function above 
#   
#   active_regions <- c("Northern South America", "Southern South America",
#                       "India-Nepal-Bangladesh", "Southeast Asia", "Madagascar") #pick ones of interest 
#   
#   sf_plot    <- st_transform(sf_obj, 4326) #transform to right projection 
#   coords_all <- st_coordinates(sf_plot)
#   
#   df <- data.frame(
#     x      = coords_all[,1],
#     y      = coords_all[,2],
#     zscore = sf_plot$p95_zscore,
#     region = sf_plot$region) |> filter(region %in% active_regions)
#   
#   region_counts <- df |>
#     count(region) |>
#     filter(n >= min_obs)
#   
#   all_elbow_data <- list()
#   plots          <- list()
#   optimal_ks     <- c()
#   
#   for (reg in region_counts$region) {
#     df_reg  <- df |>
#       filter(region == reg)
#     
#     n_obs   <- nrow(df_reg)
#     k_max   <- min(max_k, floor(n_obs / 2))
#     
#     cat("\n--- Region:", reg, "| n =", n_obs, "| testing k = 2 to", k_max, "---\n")
#     
#     coords_reg <- as.matrix(df_reg[, c("x", "y")])
#     z          <- df_reg$zscore
#     
#     elbow_df <- data.frame(
#       k           = 2:k_max,
#       moran_I     = NA,
#       n_subgraphs = NA,
#       connected   = NA
#     )
#     
#     for (i in seq_along(2:k_max)) {
#       k_try <- (2:k_max)[i]
#       
#       tryCatch({
#         knn <- knearneigh(coords_reg, k = k_try)
#         nb  <- knn2nb(knn)
#         
#         n_sub <- n.comp.nb(nb)$nc
#         elbow_df$n_subgraphs[i] <- n_sub
#         elbow_df$connected[i]   <- n_sub == 1
#         
#         lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
#         mt <- moran.test(z, lw, zero.policy = TRUE)
#         elbow_df$moran_I[i] <- mt$estimate[1]
#         
#       }, error = function(e) {
#         cat("  Error at k =", k_try, ":", e$message, "\n")
#       })
#     }
#     
#     all_elbow_data[[reg]] <- elbow_df
#     
#     # Use minimum connected k directly
#     min_connected_k <- elbow_df %>%
#       filter(connected == TRUE) %>%
#       slice(1) %>%
#       pull(k)
#     
#     optimal_k <- min_connected_k
#     optimal_ks[reg] <- optimal_k
#     
#     cat("  Minimum connected k:", min_connected_k, "\n")
#     cat("  Using k =", optimal_k, "\n")
#     
#     # Plot elbow curve
#     p <- ggplot(elbow_df, aes(x = k, y = moran_I)) +
#       geom_line(color = "grey50", linewidth = 0.8) +
#       geom_point(aes(color = connected, shape = connected), size = 3) +
#       geom_vline(xintercept = optimal_k, linetype = "dashed",
#                  color = "#d7191c", linewidth = 0.8) +
#       annotate("text", x = optimal_k + 0.2,
#                y = max(elbow_df$moran_I, na.rm = TRUE),
#                label = paste("min connected k =", optimal_k),
#                hjust = 0, color = "#d7191c", size = 3.5) +
#       scale_color_manual(values = c("TRUE" = "#2c7bb6", "FALSE" = "#fdae61"),
#                          labels = c("TRUE" = "Connected", "FALSE" = "Sub-graphs"),
#                          name = NULL) +
#       scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 17),
#                          labels = c("TRUE" = "Connected", "FALSE" = "Sub-graphs"),
#                          name = NULL) +
#       labs(title    = paste("Connectivity Check -", reg),
#            subtitle = "Moran's I vs k | dashed line = minimum connected k",
#            x        = "k (number of neighbors)",
#            y        = "Global Moran's I") +
#       theme_minimal(base_size = 14) +
#       theme(
#         plot.background = element_rect(fill = "white", color = NA),
#         legend.position = "bottom"
#       )
#     
#     plots[[reg]] <- p
#   }
#   
#   summary_df <- data.frame(
#     region    = names(optimal_ks),
#     optimal_k = as.integer(optimal_ks)
#   )
#   
#   cat("\n============================================\n")
#   cat("Minimum connected k by region:\n")
#   print(summary_df)
#   cat("============================================\n")
#   
#   list(plots = plots, optimal_ks = summary_df, elbow_data = all_elbow_data)
# }


```