---
title: "LISA Analysis for Analysis of Movement"
description: "Understanding clustering of the avulsion data"
author: "Sam Lance"
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
    theme: Lux
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r}
library(sf)
library(dplyr)
library(spdep)
library(ggplot2)
library(janitor)
library(tidyverse)
library(here)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(RColorBrewer)
library(ggnewscale)
library(terra)
library(tidyterra)

world_map <- ne_countries(scale = "medium", returnclass = "sf") |>
  st_make_valid() |>
  st_transform(4326)
```

## Load and Join Coordinates

```{r}

site_xy_og <- read_csv(here("data","base_sites.csv")) 
dis_z <- read_csv(here("data","discharge_av_year_z.csv"))
sed_z <- read_csv(here("data","ssc_av_year_z.csv"))

#Load Avulsion Data
site_xy <- read_csv(here("data","base_sites.csv")) |>
  select(site_name, site_lat, site_lon) |>
  distinct() |>
  drop_na()

#Load Discharge Z-Score Data + Convert to SF Object 
dis_sf <- read_csv(here("data","discharge_av_year_z.csv")) |>
  inner_join(site_xy, by = "site_name") |> #need lat lon from xy data, so join 
  drop_na(p95_zscore, site_lat, site_lon) |> #get rid of NA values 
  st_as_sf(coords = c("site_lon","site_lat"), crs = 4326) |> #convert to sf object 
  st_transform(6933) #transform to equal area projection

#Load Sediment Z-Score Data + Convert to SF Object 
sed_sf <- read_csv(here("data","ssc_av_year_z.csv")) |>
   st_as_sf(coords = c("site_lon","site_lat"), crs = 4326) |>  #convert to sf object
   st_transform(6933) #transform to equal area projection 

#Find Sites in Dis_sf with same lat lon values
duplicates <- site_xy_og %>%
  group_by(site_lat, site_lon) %>%
  filter(n() > 1) |>
  ungroup() |>
  inner_join(dis_sf %>% st_drop_geometry(), by = c("site_name")) 

```
## Spatial Analysis 

### Assigning Continents 

Need to assign continents to the data in order to group them by continent in our final analysis

```{r}
#Assigning Continents 
assign_continents <- function(sf_df) {
  
  #See if Already a Continent Column
  if ("continent" %in% colnames(sf_df)) {
    cat("Data already has continent column. Skipping assignment.\n")
    return(sf_df)
  }
  
  cat("Original CRS of data:", st_crs(sf_df)$input, "\n")
  
  #Convert Everything into WGS84
  sf_df_wgs84 <- st_transform(sf_df, 4326)  #my data
  cat("Transformed data to WGS84\n")
  
  world <- ne_countries(scale = "medium", returnclass = "sf") %>% #worldmap data from r 
    select(continent) %>%
    st_transform(4326)  
  
  cat("World map CRS: WGS 84\n")
  
  #Spatial Join to Add Continent Column 
  cat("Performing spatial join...\n")
  sf_df_joined <- st_join(sf_df_wgs84, world, join = st_nearest_feature, left = TRUE)
  
  #Confirm there is a Continent Column 
  if (!"continent" %in% colnames(sf_df_joined)) {
    warning("Spatial join did not create 'continent' column. Adding 'Unknown' for all.")
    sf_df_joined$continent <- "Unknown"
  } else {
    #Count How Many Got Assigned 
    n_assigned <- sum(!is.na(sf_df_joined$continent))
    cat("Successfully assigned continents to", n_assigned, "out of", nrow(sf_df_joined), "locations\n")
    
    #Handle NAs
    sf_df_joined <- sf_df_joined %>%
      mutate(continent = replace_na(continent, "Unknown"))
  }

    #Transform Back to OG CRS to Keep Consistent with Rest of Analysis
  sf_df_final <- st_transform(sf_df_joined, st_crs(sf_df))
  cat("Transformed back to original CRS\n\n")
  
  return(sf_df_final)
}

#Aggregate by Location
aggregate_locations <- function(sf_df) {
  sf_df %>%
    group_by(geometry, continent) %>%
    summarize(
      p95_zscore = mean(p95_zscore, na.rm = TRUE),
      n_events = n(),
      .groups = "drop"
    )
}

```

### Wrapper Function to Filter by Continent 
```{r}
#Create Function to Run Continent Analysis  
run_spatial_suite <- function(sf_df, k_clusters = 8, nsim = 999, alpha = 0.05, 
                              by_continent = TRUE, min_obs = 10) {
  
  #IF Analyze by Continent 
  if (by_continent) {
    
    #Only Select Continents with at Least 10 Observations 
    continent_counts <- sf_df %>%
      st_drop_geometry() %>%
      count(continent) %>%
      filter(n >= min_obs)
    
    if (nrow(continent_counts) == 0) {
      stop("No continents have >= ", min_obs, " observations. Set by_continent = FALSE or lower min_obs.")
    }
    
    cat("Analyzing", nrow(continent_counts), "continent(s):\n")
    print(continent_counts) #print result of continent analysis 
    
    #Run Analysis by Continent 
    results <- continent_counts$continent %>% #pull continent 
      setNames(., .) %>%
      lapply(function(cont) {
        cat("\n--- Analyzing:", cont, "---\n")
        
        cont_data <- sf_df %>% filter(continent == cont)
        n_obs <- nrow(cont_data) #number of observations in that continent
        k_use <- min(k_clusters, floor(n_obs / 3)) #ensure at least ~3 points per cluster on average
        k_use <- max(k_use, 2) #ensure at least 2 clusters
        
        cat("  Locations:", n_obs, "| k =", k_use, "\n")
        
        #Run the Single Continent Analysis
        #Specified below 
        tryCatch({
          result <- run_spatial_suite_single(cont_data, k_clusters = k_use,
                                   nsim = nsim, alpha = alpha)
          result$continent <- cont
          result
        }, error = function(e) {
          warning("Error in ", cont, ": ", e$message)
          NULL
        })
      })
    
    #Remove Anything that Did Not Work 
    results <- results[!sapply(results, is.null)]
    
    #Combine Results Together 
    combined_sf <- do.call(rbind, lapply(results, function(x) x$sf))
    
    #Create SUmmary Table of All Results 
    summary_table <- do.call(rbind, lapply(results, function(x) {
      data.frame(
        continent = x$continent,
        n_locations = nrow(x$sf),
        moran_I = round(x$global_moran$estimate[1], 3),
        moran_p = round(x$global_moran$p.value, 4),
        n_hotspots = sum(x$sf$gi_cluster == "Hotspot (high)"),
        n_coldspots = sum(x$sf$gi_cluster == "Coldspot (low)")
      )
    }))
    
    return(list(
      results_by_continent = results,
      combined_sf = combined_sf,
      summary = summary_table
    ))
    
  } else {
    # Run global analysis (original behavior)
    return(run_spatial_suite_single(sf_df, k_clusters = k_clusters, nsim = nsim, alpha = alpha))
  }
}


```


### Core Function
```{r}
run_spatial_suite_single <- function(sf_df, k_clusters = 8, nsim = 999, alpha = 0.05) {
  
  #Data Standardization 
  z <- sf_df |>
    st_drop_geometry() |> #remove geometry column 
    pull(p95_zscore) |> #just get the p95 zscore 
    as.numeric() #turn into a numeric 
  
  #Setup
  coords <- st_coordinates(sf_df) #pulling coordinates into variable coords
  
  #K-means Clustering: Groups points into k_clusters clusters based on coordinate proximity
  #Unlike KNN which finds k nearest neighbors per point, kmeans assigns all points to a cluster
  #then neighbors are defined as all other points within the same cluster
  set.seed(42) #set seed for reproducibility since kmeans uses random initialization
  km <- kmeans(coords, centers = k_clusters, nstart = 25) #nstart = 25 runs kmeans 25 times with different starting points and keeps the best result
  cluster_ids <- km$cluster #extract which cluster each point belongs to
  
  #Build Neighbor List from Cluster Membership
  #Each point's neighbors are all other points assigned to the same cluster
  
  #Build Neighbor List from Cluster Membership
#For any singleton points, assign them to their nearest neighbor as a fallback
  nb <- lapply(seq_len(nrow(sf_df)), function(i) {
    neighbors <- which(cluster_ids == cluster_ids[i])
    neighbors <- neighbors[neighbors != i]
  
    if (length(neighbors) == 0) {
    #If singleton, find the single nearest point by Euclidean distance
      dists <- sqrt(rowSums((coords - coords[rep(i, nrow(coords)), ])^2))
      dists[i] <- Inf #exclude self
      as.integer(which.min(dists)) #assign nearest point as neighbor
    } else {
      as.integer(neighbors)
    }
})
class(nb) <- "nb"
attr(nb, "region.id") <- as.character(seq_len(nrow(sf_df)))

#Convert Neighbors to Spatial Weights List with Row Standardized Weights
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
  
  #Calculating Global Morans I 
  # z = standardized variable created above 
  # lw = spatial weights list
  moran_global <- moran.test(z, lw, zero.policy = TRUE)
  
  #Calculating Getis Ord GI* 
  gi <- as.numeric(localG(z, lw, zero.policy = TRUE)) #calculates GI stat for each observation, compares global sum to local sum
  gi_p <- 2 * pnorm(-abs(gi)) #so we can get values that have significant clustering of high values (hotspots) and low values (coldspots)
  gi_class <- rep("Not significant", length(gi))
  gi_class[gi_p <= alpha & gi > 0] <- "Hotspot (high)" #positive GI values indicate clustering of high values (hotspots)
  gi_class[gi_p <= alpha & gi < 0] <- "Coldspot (low)" #negative GI values indicate clustering of low values (coldspots)
  
  #Calculating LISA 
  lisa_raw <- localmoran_perm(z, lw, nsim = nsim, zero.policy = TRUE) #finds local morans i for each observation using permutations 
  
  #Helps Extract Columns 
  get_col <- function(x, candidates) {
    nm <- colnames(x) #gets all column names from the lisa_raw object 
    hit <- candidates[candidates %in% nm][1] #finds the first candidate column name that matches one of the column names in lisa_raw
    if (is.na(hit)) stop("Missing expected columns. Available: ", paste(nm, collapse = ", "))
    x[, hit]
  }
  
  #What We Should Extract
  Ii <- get_col(lisa_raw, c("statistic", "Ii")) #look for names that are either one of these 
  p  <- get_col(lisa_raw, c("p.value", "p_sim", "p.value.sim", "p.value.perm", "Pr(z != E(Ii))")) #look for names that are either one of these
  
  #Calculating the Spatial Lag 
  #Finds the weighted average of neighbors values 
  lag_z <- lag.listw(lw, z, zero.policy = TRUE)
  
  #Classification 
  sig <- p <= alpha #sets significance threshhold 
  lisa_class <- rep("Not significant", length(z))
  lisa_class[sig & z > 0 & lag_z > 0] <- "High-High"
  lisa_class[sig & z < 0 & lag_z < 0] <- "Low-Low"
  lisa_class[sig & z > 0 & lag_z < 0] <- "High-Low"
  lisa_class[sig & z < 0 & lag_z > 0] <- "Low-High"
  
  #Creating Output Dataframe with All Results
  sf_out <- sf_df |>
    mutate(
      z_std        = z, #zscore 
      moran_Ii     = Ii, #local morans i i value  
      moran_p      = p, #local morans i p-value 
      lag_z        = lag_z, #finds the spatial lag 
      lisa_cluster = lisa_class, #cluster class of lisa 
      gi_star      = gi, #gi* statistic
      gi_p         = gi_p, #gi* p-value
      gi_cluster   = gi_class #gi* class assigned 
    )
  
  list(sf = sf_out, lw = lw, global_moran = moran_global)
}


# Step 1: Assign continents
dis_sf <- assign_continents(dis_sf)
sed_sf <- assign_continents(sed_sf)

# Step 2: Aggregate by location
dis_sf_agg <- aggregate_locations(dis_sf)
sed_sf_agg <- aggregate_locations(sed_sf)

# Step 3: Run analysis by continent (default)
dis_results <- run_spatial_suite(dis_sf_agg, k_clusters = 8, by_continent = TRUE, min_obs = 10)
sed_results <- run_spatial_suite(sed_sf_agg, k_clusters = 8, by_continent = TRUE, min_obs = 10)

# View summary
print(dis_results$summary)
print(sed_results$summary)

dis_results$combined_sf %>%
  st_drop_geometry() %>%
  count(continent, lisa_cluster)
print(dis_results$combined_sf %>%
  st_drop_geometry() %>%
  count(continent, lisa_cluster))

sed_results$combined_sf %>%
  st_drop_geometry() %>%
  count(continent, lisa_cluster)
print(sed_results$combined_sf %>%
  st_drop_geometry() %>%
  count(continent, lisa_cluster))

```
```{r}

# # For discharge results, picking a continent
# nb_dis <- dis_results$results_by_continent$`Africa`$lw$neighbours
# sum(sapply(nb_dis, length) == 1)
# 
# #For SSC Results
# nb_sed <- sed_results$results_by_continent$``$lw$neighbours
# sum(sapply(nb_sed, length) == 1)

```
### Kernel Density Estimation

#### Global KDE 

```{r}

# ============================================================================
# KDE FUNCTION
# ============================================================================
plot_kde_world <- function(sf_obj, title_text, save_path = NULL) {
  
  # Extract coordinates and zscore
  sf_plot <- st_transform(sf_obj, 4326)
  coords <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x = coords[, 1],
    y = coords[, 2],
    zscore = sf_plot$p95_zscore
  )
  
  # Global map
  p <- ggplot() +
    geom_sf(data = world_map, fill = "grey85", color = "grey65", linewidth = 0.25) +
    stat_density_2d(
      data = df,
      aes(x = x, y = y, fill = after_stat(level), weight = zscore),
      geom = "polygon",
      contour = TRUE,
      alpha = 0.6,
      bins = 12
    ) +
    scale_fill_distiller(palette = "RdYlBu", direction = -1, name = "Density\n(weighted by\np95_zscore)") +
    geom_point(data = df, aes(x = x, y = y), size = 0.8, alpha = 0.4, color = "black") +
    coord_sf(xlim = c(-175, 175), ylim = c(-85, 85), expand = FALSE, clip = "on") +
    labs(title = title_text) +
    theme_minimal(base_size = 14) +
    theme(
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      legend.position = "right"
    )
  
  if (!is.null(save_path)) {
    ggsave(save_path, p, width = 12, height = 7)
    cat("Saved:", save_path, "\n")
  }
  
  p
}
```

#### Continent KDE 

```{r}
plot_kde_by_continent <- function(sf_obj, title_prefix, save_folder = NULL) {
  
  sf_plot <- st_transform(sf_obj, 4326)
  coords  <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x         = coords[, 1],
    y         = coords[, 2],
    zscore    = sf_plot$p95_zscore,
    continent = sf_plot$continent
  )
  
  continents <- unique(df$continent)
  plots <- list()
  
  # Define explicit color palettes for hot and cold
  hot_colors  <- c("1" = "#FFFFB2", "2" = "#FECC5C", "3" = "#FD8D3C", "4" = "#E31A1C")
  cold_colors <- c("1" = "#EFF3FF", "2" = "#BDD7E7", "3" = "#6BAED6", "4" = "#2171B5")
  
  for (cont in continents) {
    df_cont <- df %>% filter(continent == cont)
    
    if (nrow(df_cont) < 5) {
      cat("Skipping", cont, "- too few points\n")
      next
    }
    
    df_hot  <- df_cont %>% filter(zscore >  0)
    df_cold <- df_cont %>% filter(zscore <  0)
    
    has_hot  <- nrow(df_hot)  >= 3
    has_cold <- nrow(df_cold) >= 3
    
    p <- ggplot() +
      geom_sf(data = world_map, fill = "grey85", color = "grey65", linewidth = 0.25)
    
    # --- Hotspot layer ---
  if (has_hot) {
    p <- p +
      geom_density_2d_filled(
        data  = df_hot,
        aes(x = x, y = y, weight = zscore),
        alpha = 0.55, bins = 4
      ) +
      scale_fill_manual(
        values = hot_colors,
        name   = "Hotspot\ndensity",
        labels = c("Low", "", "", "High"),
        drop   = FALSE
      )
  }

# --- Coldspot layer ---
  if (has_cold) {
    p <- p +
      ggnewscale::new_scale_fill() +   # reset AFTER first scale
      geom_density_2d_filled(
        data  = df_cold,
        aes(x = x, y = y, weight = abs(zscore)),
        alpha = 0.55, bins = 4
      ) +
      scale_fill_manual(
        values = cold_colors,
        name   = "Coldspot\ndensity",
        labels = c("Low", "", "", "High"),
        drop   = FALSE
      )
  }
    
    p <- p +
      geom_point(
        data  = df_cont,
        aes(x = x, y = y),
        size  = 1.2, alpha = 0.5, color = "black"
      ) +
      coord_sf(
        xlim   = range(df_cont$x) + c(-2, 2),
        ylim   = range(df_cont$y) + c(-2, 2),
        expand = FALSE, clip = "on"
      ) +
      labs(title = paste(title_prefix, "-", cont)) +
      theme_minimal(base_size = 14) +
      theme(
        axis.title      = element_blank(),
        axis.text       = element_blank(),
        axis.ticks      = element_blank(),
        panel.grid      = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right"
      )
    
    plots[[cont]] <- p
    
    if (!is.null(save_folder)) {
      fname <- paste0(
        save_folder, "/",
        tolower(gsub(" ", "_", title_prefix)), "_kde_",
        tolower(gsub(" ", "_", cont)), ".png"
      )
      ggsave(fname, p, width = 10, height = 7)
      cat("Saved:", fname, "\n")
    }
  }
  
  plots
}
```

#### Raster KDE 

```{r}

plot_zscore_raster_by_continent <- function(sf_obj, title_prefix, 
                                             sigma = 1, res = 0.25,
                                             save_folder = NULL) {
  
  sf_plot <- st_transform(sf_obj, 4326)
  coords  <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x         = coords[, 1],
    y         = coords[, 2],
    zscore    = sf_plot$p95_zscore,
    continent = sf_plot$continent
  )
  
  continents <- unique(df$continent)
  plots <- list()
  
  for (cont in continents) {
    df_cont <- df %>% filter(continent == cont)
    
    if (nrow(df_cont) < 5) {
      cat("Skipping", cont, "- too few points\n")
      next
    }
    
    # Define raster extent from continent points
    xlim <- range(df_cont$x) + c(-2, 2)
    ylim <- range(df_cont$y) + c(-2, 2)
    
    # Create empty raster at specified resolution
    r_template <- rast(xmin = xlim[1], xmax = xlim[2],
                       ymin = ylim[1], ymax = ylim[2],
                       resolution = res, crs = "EPSG:4326")
    
    # Convert points to SpatVector
    pts <- vect(df_cont, geom = c("x", "y"), crs = "EPSG:4326")
    
    # Rasterize - average zscore per cell
    r_mean <- rasterize(pts, r_template, field = "zscore", fun = "mean")
    
    # Smooth using focal mean (kernel smoothing)
    # sigma controls neighborhood size - larger = smoother
    fw <- focalMat(r_mean, sigma, type = "Gauss")
    r_smooth <- focal(r_mean, w = fw, fun = "sum", na.policy = "omit", na.rm = TRUE)
    
    # Convert to dataframe for ggplot
    r_df <- as.data.frame(r_smooth, xy = TRUE) %>%
      rename(zscore = 3) %>%
      filter(!is.na(zscore))
    
    # Symmetric color limits centered at zero
    max_val <- max(abs(r_df$zscore), na.rm = TRUE)
    
    p <- ggplot() +
      geom_sf(data = world_map, fill = "grey85", color = "grey65", linewidth = 0.25) +
      geom_tile(data = r_df, aes(x = x, y = y, fill = zscore), alpha = 0.75) +
      scale_fill_distiller(palette = "RdBu", direction = -1,
                           name = "p95_zscore",
                           limits = c(-max_val, max_val)) +
      geom_point(data = df_cont, aes(x = x, y = y), 
                 size = 1.2, alpha = 0.6, color = "black") +
      coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, clip = "on") +
      labs(title = paste(title_prefix, "-", cont),
           subtitle = paste("Smoothed p95_zscore | red = high | blue = low")) +
      theme_minimal(base_size = 14) +
      theme(
        axis.title = element_blank(),
        axis.text  = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right"
      )
    
    plots[[cont]] <- p
    
    if (!is.null(save_folder)) {
      fname <- paste0(save_folder, "/", tolower(gsub(" ", "_", title_prefix)), 
                      "_raster_", tolower(gsub(" ", "_", cont)), ".png")
      ggsave(fname, p, width = 10, height = 7)
      cat("Saved:", fname, "\n")
    }
  }
  
  plots
}

```

#### Summary Table 

```{r}
# ============================================================================
# SUMMARY FUNCTION
# ============================================================================
kde_summary <- function(sf_obj, label) {
  sf_plot <- st_transform(sf_obj, 4326)
  df <- data.frame(
    zscore    = sf_plot$p95_zscore,
    continent = sf_plot$continent
  )
  
  cat("\n===", label, "KDE Summary ===\n")
  summary_tbl <- df %>%
    group_by(continent) %>%
    summarise(
      n          = n(),
      mean_zscore = round(mean(zscore, na.rm = TRUE), 3),
      sd_zscore   = round(sd(zscore, na.rm = TRUE), 3),
      min_zscore  = round(min(zscore, na.rm = TRUE), 3),
      max_zscore  = round(max(zscore, na.rm = TRUE), 3)
    )
  print(summary_tbl)
  summary_tbl
}
```

#### Run Everything 

```{r}
# ============================================================================
# RUN EVERYTHING
# ============================================================================
# dir.create(here("kde_plots"), showWarnings = FALSE)

# # Global KDE maps
# p_dis_kde_global <- plot_kde_world(dis_results$combined_sf, 
#                                     "Discharge KDE (weighted by p95_zscore)",
#                                     here("kde_plots", "discharge_kde_global.png"))
# 
# p_sed_kde_global <- plot_kde_world(sed_results$combined_sf, 
#                                     "SSC KDE (weighted by p95_zscore)",
#                                     here("kde_plots", "ssc_kde_global.png"))

# # By continent KDE maps
# dis_kde_continent <- plot_kde_by_continent(dis_results$combined_sf, "Discharge", here("kde_plots"))
# sed_kde_continent <- plot_kde_by_continent(sed_results$combined_sf, "SSC", here("kde_plots"))

# Summaries
dis_summary <- kde_summary(dis_results$combined_sf, "Discharge")
sed_summary <- kde_summary(sed_results$combined_sf, "SSC")

# # Display global maps
# p_dis_kde_global
# p_sed_kde_global
# 
# # Display continent maps
# dis_kde_continent
# sed_kde_continent

# Run it
dis_raster_plots <- plot_zscore_raster_by_continent(dis_results$combined_sf, 
                                                     title_prefix = "Discharge",
                                                     sigma = 2, res = 0.5,
                                                     save_folder = here("kde_plots"))

sed_raster_plots <- plot_zscore_raster_by_continent(sed_results$combined_sf, 
                                                     title_prefix = "SSC",
                                                     sigma = 2, res = 0.5,
                                                     save_folder = here("kde_plots"))

# Display
dis_raster_plots
sed_raster_plots

```
### CURRENT CHOICE - Regional Analysis 

#### Filtering Out
Filter out more specific regions and do some testing 

```{r}

assign_regions <- function(sf_obj) {
  coords <- st_coordinates(st_transform(sf_obj, 4326))
  
  sf_obj <- sf_obj %>%
    mutate(
      lon = coords[,1],
      lat = coords[,2],
      region = case_when(
        
        # Northern South America (Colombia, Venezuela, Ecuador, Peru, Brazil north)
        continent == "South America" & lat > -5 ~ "Northern South America",
        
        # Southern South America (Chile, Argentina, Bolivia, Brazil south)
        continent == "South America" & lat <= -5 ~ "Southern South America",
        
        # Northern India / Nepal / Bangladesh region
        continent == "Asia" & lat >= 20 & lat <= 32 & lon >= 68 & lon <= 108 ~ "India-Nepal-Bangladesh",
        
        # Southeast Asia
        continent == "Asia" & lat <= 20 & lon >= 95 & lon <= 190 ~ "Southeast Asia",
        
        # # Southeast Asia
        # continent == "Oceania" & lat <= 20 & lon >= 95 & lon <= 180 ~ "Southeast Asia",
        
        # # Central/East Asia (Tibetan Plateau, China, Mongolia)
        # continent == "Asia" & lat > 30 & lon >= 70 ~ "Central-East Asia",
        
        # # Middle East / Western Asia
        # continent == "Asia" & lon < 70 ~ "Middle East",
        
        # Madagascar separate from mainland Africa
        continent == "Africa" & lon >= 43 & lon <= 51 & lat >= -26 & lat <= -12 ~ "Madagascar",
        
        # # Mainland Africa
        # continent == "Africa" ~ "Mainland Africa",
        
        # Everything else keeps continent name
        TRUE ~ continent
      )
    )
}

dis_sf_agg %>%
  assign_regions() %>%
  st_drop_geometry() %>%
  count(region)

library(RColorBrewer)

# First assign regions to your data
dis_sf_regions <- assign_regions(dis_sf_agg)
sed_sf_regions <- assign_regions(sed_sf_agg)

# Plot function
plot_regions_map <- function(sf_obj, title_text) {
  
  sf_plot <- st_transform(sf_obj, 4326)
  coords  <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x      = coords[,1],
    y      = coords[,2],
    region = sf_plot$region
  )
  
  # Get number of regions for palette
  n_regions <- length(unique(df$region))
  region_colors <- setNames(
    colorRampPalette(brewer.pal(8, "Set1"))(n_regions),
    unique(df$region)
  )
  
  ggplot() +
    geom_sf(data = world_map, fill = "grey90", color = "grey65", linewidth = 0.25) +
    geom_point(data = df, 
               aes(x = x, y = y, color = region),
               size = 3, alpha = 0.85) +
    scale_color_manual(values = region_colors, name = "Region") +
    coord_sf(xlim = c(-175, 175), ylim = c(-85, 85), expand = FALSE) +
    labs(title = title_text) +
    theme_minimal(base_size = 14) +
    theme(
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      legend.position = "right"
    )
}

# Run and save
p_dis_regions <- plot_regions_map(dis_sf_regions, "Discharge - Region Assignments")
p_sed_regions <- plot_regions_map(sed_sf_regions, "SSC - Region Assignments")

p_dis_regions
p_sed_regions

dis_sf_regions %>%
  st_drop_geometry() %>%
  filter(region %in% c("India-Nepal-Bangladesh", "Southeast Asia", "Central-East Asia")) %>%
  select(lon, lat, region) %>%
  arrange(lon)

```
#### Mapping 

##### KDE Only

```{r}

plot_kde_by_region <- function(sf_obj, title_prefix, 
                                min_obs = 5,
                                save_folder = NULL) {
  
  sf_plot <- st_transform(sf_obj, 4326)
  coords  <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x      = coords[, 1],
    y      = coords[, 2],
    zscore = sf_plot$p95_zscore,
    region = sf_plot$region
  )
  
  # Filter to regions with sufficient observations
  region_counts <- df %>% count(region) %>% filter(n >= min_obs)
  cat("Analyzing", nrow(region_counts), "region(s):\n")
  print(region_counts)
  
  plots <- list()
  
  for (reg in region_counts$region) {
    df_reg <- df %>% filter(region == reg)
    cat("\n--- Analyzing:", reg, "| n =", nrow(df_reg), "---\n")
    
    p <- ggplot() +
      geom_sf(data = world_map, fill = "grey85", color = "grey65", linewidth = 0.25) +
      stat_density_2d(
        data = df_reg,
        aes(x = x, y = y, fill = after_stat(level), weight = zscore),
        geom = "polygon",
        contour = TRUE,
        alpha = 0.6,
        bins = 8
      ) +
      scale_fill_distiller(palette = "RdYlBu", direction = -1, 
                           name = "Density\n(weighted by\np95_zscore)") +
      geom_point(data = df_reg, aes(x = x, y = y), 
                 size = 1.2, alpha = 0.5, color = "black") +
      coord_sf(
        xlim   = range(df_reg$x) + c(-2, 2),
        ylim   = range(df_reg$y) + c(-2, 2),
        expand = FALSE, clip = "on"
      ) +
      labs(title    = paste(title_prefix, "-", reg),
           subtitle = "KDE weighted by p95_zscore") +
      theme_minimal(base_size = 14) +
      theme(
        axis.title      = element_blank(),
        axis.text       = element_blank(),
        axis.ticks      = element_blank(),
        panel.grid      = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right"
      )
    
    plots[[reg]] <- p
    
    if (!is.null(save_folder)) {
      fname <- paste0(save_folder, "/", tolower(gsub(" ", "_", title_prefix)),
                      "_kde_", tolower(gsub(" ", "_", reg)), ".png")
      ggsave(fname, p, width = 10, height = 7)
      cat("Saved:", fname, "\n")
    }
  }
  
  plots
}

# Run it
dis_kde_region <- plot_kde_by_region(dis_sf_regions,
                                      title_prefix = "Discharge",
                                      save_folder  = here("kde_plots"))

sed_kde_region <- plot_kde_by_region(sed_sf_regions,
                                      title_prefix = "SSC",
                                      save_folder  = here("kde_plots"))

dis_kde_region
sed_kde_region
```

##### Hotcold KDE

```{r}



plot_kde_hotcold_by_region <- function(sf_obj, title_prefix,
                                        min_obs = 5,
                                        save_folder = NULL) {
  
  sf_plot <- st_transform(sf_obj, 4326)
  coords  <- st_coordinates(sf_plot)
  
  df <- data.frame(
    x      = coords[, 1],
    y      = coords[, 2],
    zscore = sf_plot$p95_zscore,
    region = sf_plot$region
  )
  
  active_regions <- c("Northern South America", "Southern South America",
                      "India-Nepal-Bangladesh", "Southeast Asia", "Madagascar")
  
  df <- df %>% filter(region %in% active_regions)
  
  region_counts <- df %>% count(region) %>% filter(n >= min_obs)
  
  cat("Analyzing", nrow(region_counts), "region(s):\n")
  print(region_counts)
  
  # ---- LOOP 1: Calculate global density ranges ----
  cat("\nCalculating global density ranges for consistent scaling...\n")
  all_hot_levels  <- c()
  all_cold_levels <- c()
  
  for (reg in region_counts$region) {
    df_reg  <- df %>% filter(region == reg)
    df_hot  <- df_reg %>% filter(zscore > 0)
    df_cold <- df_reg %>% filter(zscore < 0)
    
    if (nrow(df_hot) >= 3) {
      dens <- MASS::kde2d(df_hot$x, df_hot$y, n = 100,
                          lims = c(range(df_hot$x) + c(-2, 2),
                                   range(df_hot$y) + c(-2, 2)))
      all_hot_levels <- c(all_hot_levels, range(dens$z))
    }
    
    if (nrow(df_cold) >= 3) {
      dens <- MASS::kde2d(df_cold$x, df_cold$y, n = 100,
                          lims = c(range(df_cold$x) + c(-2, 2),
                                   range(df_cold$y) + c(-2, 2)))
      all_cold_levels <- c(all_cold_levels, range(dens$z))
    }
  }
  
  global_hot_max  <- max(all_hot_levels,  na.rm = TRUE)
  global_cold_max <- max(all_cold_levels, na.rm = TRUE)
  
  cat("Global hotspot density range:  0 -", round(global_hot_max, 5), "\n")
  cat("Global coldspot density range: 0 -", round(global_cold_max, 5), "\n")
  
  # ---- LOOP 2: Build plots ----
  plots <- list()
  
  for (reg in region_counts$region) {
    df_reg  <- df %>% filter(region == reg)
    
    # Weighted resampling for hotspots
    df_hot <- df_reg %>% filter(zscore > 0)
    if (nrow(df_hot) >= 3) {
      weight_times <- pmax(round(df_hot$zscore / min(df_hot$zscore) * 10), 1)
      df_hot <- df_hot[rep(seq_len(nrow(df_hot)), times = weight_times), ]
    }
    
    # Weighted resampling for coldspots
    df_cold <- df_reg %>% filter(zscore < 0)
    if (nrow(df_cold) >= 3) {
      weight_times <- pmax(round(abs(df_cold$zscore) / min(abs(df_cold$zscore)) * 10), 1)
      df_cold <- df_cold[rep(seq_len(nrow(df_cold)), times = weight_times), ]
    }
    
    xlim <- range(df_reg$x) + c(-2, 2)
    ylim <- range(df_reg$y) + c(-2, 2)
    
    cat("\n--- Plotting:", reg, "| n =", nrow(df_reg), "---\n")
    
    # --- HOTSPOT PLOT ---
    if (nrow(df_hot) >= 3) {
      p_hot <- ggplot() +
        geom_sf(data = world_map, fill = "grey85", color = "grey65", linewidth = 0.25) +
        stat_density_2d(
          data = df_hot,
          aes(x = x, y = y, fill = after_stat(level)),
          geom = "polygon", contour = TRUE, alpha = 0.65, bins = 6
        ) +
        scale_fill_distiller(palette = "YlOrRd", direction = 1,
                             name   = "Hotspot\ndensity",
                             limits = c(0, global_hot_max),
                             oob    = scales::squish) +
        geom_point(data = df_reg, aes(x = x, y = y),
                   size = 1.2, alpha = 0.5, color = "black") +
        coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, clip = "on") +
        labs(title    = paste(title_prefix, "-", reg, "| Hotspots"),
             subtitle = "Areas of clustered high p95_zscore (positive)") +
        theme_minimal(base_size = 14) +
        theme(
          axis.title      = element_blank(),
          axis.text       = element_blank(),
          axis.ticks      = element_blank(),
          panel.grid      = element_blank(),
          plot.background = element_rect(fill = "white", color = NA),
          legend.position = "right"
        )
      
      plots[[paste0(reg, "_hot")]] <- p_hot
      
      if (!is.null(save_folder)) {
        fname <- paste0(save_folder, "/", tolower(gsub(" ", "_", title_prefix)),
                        "_hot_", tolower(gsub(" ", "_", reg)), ".png")
        ggsave(fname, p_hot, width = 10, height = 7)
        cat("  Saved:", fname, "\n")
      }
    } else {
      cat("  Skipping hotspot plot - too few positive points\n")
    }
    
    # --- COLDSPOT PLOT ---
    if (nrow(df_cold) >= 3) {
      p_cold <- ggplot() +
        geom_sf(data = world_map, fill = "grey85", color = "grey65", linewidth = 0.25) +
        stat_density_2d(
          data = df_cold,
          aes(x = x, y = y, fill = after_stat(level)),
          geom = "polygon", contour = TRUE, alpha = 0.65, bins = 6
        ) +
        scale_fill_distiller(palette = "YlGnBu", direction = 1,
                             name   = "Coldspot\ndensity",
                             limits = c(0, global_cold_max),
                             oob    = scales::squish) +
        geom_point(data = df_reg, aes(x = x, y = y),
                   size = 1.2, alpha = 0.5, color = "black") +
        coord_sf(xlim = xlim, ylim = ylim, expand = FALSE, clip = "on") +
        labs(title    = paste(title_prefix, "-", reg, "| Coldspots"),
             subtitle = "Areas of clustered low p95_zscore (negative)") +
        theme_minimal(base_size = 14) +
        theme(
          axis.title      = element_blank(),
          axis.text       = element_blank(),
          axis.ticks      = element_blank(),
          panel.grid      = element_blank(),
          plot.background = element_rect(fill = "white", color = NA),
          legend.position = "right"
        )
      
      plots[[paste0(reg, "_cold")]] <- p_cold
      
      if (!is.null(save_folder)) {
        fname <- paste0(save_folder, "/", tolower(gsub(" ", "_", title_prefix)),
                        "_cold_", tolower(gsub(" ", "_", reg)), ".png")
        ggsave(fname, p_cold, width = 10, height = 7)
        cat("  Saved:", fname, "\n")
      }
    } else {
      cat("  Skipping coldspot plot - too few negative points\n")
    }
  }
  
  plots
}

# Run it
dis_kde_hotcold <- plot_kde_hotcold_by_region(dis_sf_regions,
                                               title_prefix = "Discharge",
                                               save_folder  = here("kde_plots"))

sed_kde_hotcold <- plot_kde_hotcold_by_region(sed_sf_regions,
                                               title_prefix = "SSC",
                                               save_folder  = here("kde_plots"))

dis_kde_hotcold
sed_kde_hotcold
```

## Creating Maps

```{r}


sf_use_s2(FALSE)  # helps prevent some odd wrap/line artifacts in lon/lat plots

# --- World map in lon/lat (no Robinson) ---
world_map <- ne_countries(scale = "medium", returnclass = "sf") |>
  st_make_valid() |>
  st_transform(4326)

# Graticule + globe outline in lon/lat
globe_outline <- st_as_sfc(
  st_bbox(c(xmin = -180, ymin = -90, xmax = 180, ymax = 90), crs = st_crs(4326))
)

grat <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = seq(-60,  60,  by = 30),
  crs = st_crs(4326)
)

# Latitude caps (avoid ±90 which often draws weird)
lon_seq <- seq(-180, 180, by = 0.25)
make_parallel <- function(lat) {
  st_sfc(st_linestring(cbind(lon_seq, rep(lat, length(lon_seq)))) , crs = 4326)
}
lat_min <- -89.5   # <- change to -60 if you want to clip Antarctica etc.
lat_max <-  89.5
bottom_boundary <- make_parallel(lat_min)
top_boundary    <- make_parallel(lat_max)

# Palette helper (for significant categories only)
make_palette <- function(keep_levels) {
  pal_sig <- setNames(hue_pal()(length(keep_levels)), keep_levels)
  pal_sig
}

# Plot helper: non-sig UNDER, sig OVER (no projection)
plot_world_clusters <- function(sf_obj, class_col, keep_levels, title_text) {

  sf_plot <- st_transform(sf_obj, 4326)

  cls <- sf_plot[[class_col]]
  cls2 <- ifelse(cls %in% keep_levels, as.character(cls), "Not significant")
  sf_plot[[class_col]] <- factor(cls2, levels = c("Not significant", keep_levels))

  nonsig <- sf_plot |> filter(.data[[class_col]] == "Not significant")
  sig    <- sf_plot |> filter(.data[[class_col]] != "Not significant")

  pal_sig <- make_palette(keep_levels)

  ggplot() +
    geom_sf(data = grat, color = "grey80", linewidth = 0.25) +
    geom_sf(data = world_map, fill = "grey85", color = "grey65", linewidth = 0.25) +

    # Not significant: make them visible on land/water using shape 21 (fill + outline)
    geom_sf(
      data = nonsig,
      shape = 21,
      fill  = "grey35",
      color = "grey35",
      size  = 1.8,
      alpha = 0.7,
      stroke = 0.3
    ) +

    # Significant: on top, colored by class
    geom_sf(
      data = sig,
      aes(color = .data[[class_col]]),
      size = 2.7,
      alpha = 0.95,
      stroke = 0.7
    ) +

    geom_sf(data = bottom_boundary, color = "grey75", linewidth = 0.5) +
    geom_sf(data = top_boundary,    color = "grey75", linewidth = 0.5) +
    geom_sf(data = globe_outline, fill = NA, color = "grey75", linewidth = 0.7) +

    scale_color_manual(values = pal_sig, name = NULL) +
    coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = FALSE) +
    labs(title = title_text) +
    theme_minimal(base_size = 16) +
    theme(
      axis.title = element_blank(),
      axis.text  = element_blank(),
      axis.ticks = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      legend.position = c(0.2, 0.22),
      legend.direction = "vertical",
      legend.box = "vertical",
      legend.title = element_text(face = "bold"),
      legend.title.align = 0.5,
      legend.background = element_rect(fill = "white", color = "white")
    )
}


p_dis_gi_world <- plot_world_clusters(dis_results$combined_sf, "gi_cluster",
                                     c("Hotspot (high)", "Coldspot (low)"),
                                     "Discharge Gi* hotspots/coldspots (p95_zscore)")

p_sed_gi_world <- plot_world_clusters(sed_results$combined_sf, "gi_cluster",
                                     c("Hotspot (high)", "Coldspot (low)"),
                                     "SSC Gi* hotspots/coldspots (p95_zscore)")

p_dis_lisa_world <- plot_world_clusters(dis_results$combined_sf, "lisa_cluster",
                                       c("High-High", "Low-Low", "High-Low", "Low-High"),
                                       "Discharge LISA (High-High & Low-Low) (p95_zscore)")

p_sed_lisa_world <- plot_world_clusters(sed_results$combined_sf, "lisa_cluster",
                                       c("High-High", "Low-Low", "High-Low", "Low-High"),
                                       "SSC LISA (High-High & Low-Low) (p95_zscore)")

p_dis_gi_world
p_dis_lisa_world
p_sed_gi_world
p_sed_lisa_world

#save maps to a folder called lisa_plots
ggsave(here("lisa_plots", "discharge_gi_world.png"), p_dis_gi_world, width = 10, height = 6)
ggsave(here("lisa_plots", "discharge_lisa_world.png"), p_dis_lisa_world, width = 10, height = 6)
ggsave(here("lisa_plots", "ssc_gi_world.png"), p_sed_gi_world, width = 10, height = 6)
ggsave(here("lisa_plots", "ssc_lisa_world.png"), p_sed_lisa_world, width = 10, height =6)


```