---
title: "Avulsion Project Code"
description: "Analysis of Movement Winter 2026"
author: "Sam Lance"
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
    theme: Lux
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r}
library(sf)
library(dplyr)
library(spdep)
library(ggplot2)
library(janitor)
library(tidyverse)
library(here)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(RColorBrewer)
library(ggnewscale)
library(terra)
library(tidyterra)
library(spatialreg)
library(ggrepel)
```

## Setup 
### Load Data 
```{r}

#Load Data in OG Format No Modifications 
site_xy_og <- read_csv(here("data","base_sites.csv"), show_col_types = FALSE) 
dis_z <- read_csv(here("data","discharge_av_year_z.csv"), show_col_types = FALSE)
sed_z <- read_csv(here("data","ssc_av_year_z.csv"), show_col_types = FALSE)

#Load Avulsion Data
site_xy <- read_csv(here("data","base_sites.csv"), show_col_types = FALSE) |>
  select(site_name, site_lat, site_lon) |>
  distinct() |>
  drop_na()

#Load Discharge Z-Score Data + Convert to SF Object 
avulsion_sf <- read_csv(here("data","discharge_av_year_z.csv"), show_col_types = FALSE) |>
  inner_join(site_xy, by = "site_name") |> #need lat lon from xy data, so join 
  drop_na(p95_zscore, site_lat, site_lon) |> #get rid of NA values 
  st_as_sf(coords = c("site_lon","site_lat"), crs = 4326)

```

### Create Region List 

```{r}

#Select Specific Regions 
regions <- list(
  "Northern South America" = c(xmin = -82, ymin =  -5, xmax = -55, ymax =  15),
  "Southern South America" = c(xmin = -82, ymin = -60, xmax = -55, ymax =  -5),
  "Himalaya"               = c(xmin =  72, ymin =  20, xmax = 108, ymax =  32),
  "PNG North"              = c(xmin =  130, ymin =  -15, xmax = 150, ymax =  5))

region_names <- names(regions)  # c("Northern South America", "Southern South America", ...)
 
#Convert into Polygon 
region_polys <- imap(regions, function(bb, name) {
  st_sf(
    region = name,
    geometry = st_sfc(st_polygon(list(
      matrix(c(bb["xmin"], bb["ymin"],
               bb["xmax"], bb["ymin"],
               bb["xmax"], bb["ymax"],
               bb["xmin"], bb["ymax"],
               bb["xmin"], bb["ymin"]),
             ncol = 2, byrow = TRUE))), crs = 4326)
  )
}) |> bind_rows()

```


### Reproject to World Mollweide 

Reason: want equal area CRS
```{r}
# Load world boundaries and project to match your points
world_wgs84 <- ne_countries(scale = "medium", returnclass = "sf")

target_crs <- "+proj=moll"   # World Mollweide (metres)
 
avulsion_proj  <- st_transform(avulsion_sf,   crs = target_crs)
region_proj    <- st_transform(region_polys,  crs = target_crs)
world_proj     <- st_transform(world_wgs84,   crs = target_crs)

```

### Info on Neighbor Distances

Commented out because only need to run once, meant to show what the distances look like between neighbors to select the value for the next chunk
```{r}

# for (region_name in names(regions)) {
# 
#   region_boundary <- region_proj |> filter(region == region_name)
#   pts_diag <- st_intersection(avulsion_proj, region_boundary)
# 
#   if (nrow(pts_diag) < 2) {
#     cat(sprintf("[%s] Too few points for distance diagnostic\n", region_name))
#     next
#   }
# 
#   coords <- st_coordinates(pts_diag)
#   dist_matrix <- as.matrix(dist(coords))
#   diag(dist_matrix) <- NA
# 
#   nn_dist_km <- apply(dist_matrix, 1, min, na.rm = TRUE) / 1000
# 
#   cat(sprintf("\n[%s] — %d points\n", region_name, nrow(pts_diag)))
#   cat(sprintf("  Min NN distance:    %.0f km\n", min(nn_dist_km)))
#   cat(sprintf("  Median NN distance: %.0f km\n", median(nn_dist_km)))
#   cat(sprintf("  Mean NN distance:   %.0f km\n", mean(nn_dist_km)))
#   cat(sprintf("  Max NN distance:    %.0f km\n", max(nn_dist_km)))
#   cat(sprintf("  Suggested cell size (median × 1.5): %.0f km\n",
#               median(nn_dist_km) * 1.5))
# }
# 
# stop("Diagnostic complete — review suggested cell sizes above,
#       then set cell_size_km and run the full script.")

```

## Building Spatial Weights Matrix

### Build Spatial Weights by Region 
```{r}
#Defines the Grid Cell
cell_size_km <- 25  #value is taken from the chunk above      
cell_size_m  <- cell_size_km * 1000 #put into meters 
 
build_region_weights <- function(region_name, points_proj, regions_proj, cell_km) {
 
  cell_m <- cell_km * 1000 #convert to meters 
 
  #Pick out Region Points 
  region_boundary <- regions_proj |> filter(region == region_name) #just pick out one region as running 
  pts <- st_intersection(points_proj, region_boundary) #clips only points within the region 
 
  #Error Statement: print out if there are no points found in a region, and print out how many points were found in each region 
  if (nrow(pts) == 0) {
    warning(paste("No points found in region:", region_name))
    return(NULL)
  }
  cat(sprintf("\n[%s] %d points\n", region_name, nrow(pts)))
 
  #Creates Bounding Box 
  bb <- st_bbox(region_boundary)
  x_breaks <- seq(floor(bb["xmin"] / cell_m) * cell_m,
                  ceiling(bb["xmax"] / cell_m) * cell_m,
                  by = cell_m)
  y_breaks <- seq(floor(bb["ymin"] / cell_m) * cell_m,
                  ceiling(bb["ymax"] / cell_m) * cell_m,
                  by = cell_m)
 
  cat(sprintf("  Grid: %d x-cells × %d y-cells at %d km\n",
              length(x_breaks) - 1, length(y_breaks) - 1, cell_km))
 
  #Assign Set ID 
  coords <- st_coordinates(pts)
  pts <- pts |>
    mutate(
      
      #Pull X and Y Values 
      x_m     = coords[, 1],
      y_m     = coords[, 2],
      
      #Finds What Grid Column and Row Each Point is In 
      col_idx = findInterval(x_m, x_breaks, rightmost.closed = TRUE),
      row_idx = findInterval(y_m, y_breaks, rightmost.closed = TRUE),
      cell_id = paste0("r", row_idx, "_c", col_idx)
    )
 
  cat(sprintf("  Occupied cells: %d\n", n_distinct(pts$cell_id)))
 
  #Count Number of Neighbros for Each 
  cell_ids       <- pts$cell_id #pull cell id values 
  n              <- nrow(pts)
  neighbor_counts <- table(table(cell_ids) - 1)  #number of neighbors each point has 
  print(neighbor_counts) #print these values 
 
  #INITIAL Build Neighbor List 
  #For each point finds all other points int he same cell, binary 0 or 1 neighbors 
  nb_list <- lapply(seq_len(n), function(i) { 
    same_cell <- which(cell_ids == cell_ids[i]) #identifies which one is itself 
    as.integer(same_cell[same_cell != i])
  })
  class(nb_list) <- "nb" #tells sbdep this is a neighbors list 
  attr(nb_list, "region.id") <- as.character(seq_len(n)) #labels each pointt by index AKA 1, 2, etc. 
  attr(nb_list, "call")      <- call("build_region_weights") #
  attr(nb_list, "sym")       <- FALSE #just in case neighbors not symmetric 
 
  #Finds Isolated Points 
  n_isolated <- sum(sapply(nb_list, length) == 0) #finds neighbor count for each point, finds all zeroes to see which points are isolated 
  cat(sprintf("  Isolated points (no neighbour): %d / %d\n", n_isolated, n))
 
 #CLEAN Fix for No Neighbors
  nb_list_clean <- lapply(nb_list, function(x) {
    if (length(x) == 0) 0L else x #put in OL as the length instead of 0 so spdep does not freak out
  })
  
  class(nb_list_clean) <- "nb"
  attr(nb_list_clean, "region.id") <- as.character(seq_len(n))
  attr(nb_list_clean, "call")      <- call("build_region_weights")
  attr(nb_list_clean, "sym")       <- FALSE

  #Build Spatial Weights 
  #W = row standardized, weights are divided by number of neighbors 
  lw <- tryCatch({
    nb2listw(nb_list_clean, style = "W", zero.policy = TRUE) #zero policy to allow for points with no neighbors to pass through 
  }, error = function(e) {
    cat(sprintf("  ERROR building weights for %s: %s\n", region_name, e$message))
    NULL
  })
 
  #Creats a Named List to Access Everything at the End 
  list(
    region      = region_name,
    points      = pts,
    nb          = nb_list_clean,
    lw          = lw,
    x_breaks    = x_breaks,
    y_breaks    = y_breaks
  )
}
```

### Run for All Regions 

```{r}

#Running the Above Code 
results <- imap(
  setNames(names(regions), names(regions)), #extracts names from regions, creates named vector, .x is region name string  
  ~ build_region_weights(.x, avulsion_proj, region_proj, cell_size_km) #running function above, .x is region name string
) |> compact()   #drop empty regions 
 
#Printing Completion Statment 
cat("\nSuccessfully built weights for:", paste(names(results), collapse = ", "), "\n")

```

### Save All Weights

```{r}
#Save all Weights as a RDS File 
#RDS File: R specific object
for (reg in names(results)) { #loops over each region that came out of the last section 
  if (!is.null(results[[reg]]$lw)) { #if not null, access the lw object for each region 
    filename <- paste0("weights_", gsub(" ", "_", tolower(reg)), ".rds") #filename setup 
    saveRDS(results[[reg]]$lw, filename) #save as an rds file 
    cat("Saved:", filename, "\n") #progress printout 
  }
}

```

## Spatial Statistics 
### Function for Spatial Stats  

```{r}

run_spatial_stats <- function(res) {
 
  if (is.null(res) || is.null(res[["lw"]])) {
    cat(sprintf("[%s] Skipping — no weights object\n", res$region))
    return(NULL)
  }
 
  pts          <- res$points
  lw           <- res$lw
  analysis_var <- pts$p95_zscore
  cat(sprintf("\n========== %s ==========\n", res$region))
 
  #Global Moran's I
  moran_result <- moran.test(analysis_var, lw, zero.policy = TRUE)
  cat("Global Moran's I:\n")
  print(moran_result)
 
  #Gi*
  nb_self  <- include.self(res$nb)
  lw_self  <- nb2listw(nb_self, style = "B", zero.policy = TRUE)
  gi_star  <- localG(analysis_var, lw_self, zero.policy = TRUE)
 
  pts$gi_star     <- as.numeric(gi_star)
  pts$gi_star_sig <- abs(pts$gi_star) > 1.96
 
  #LISA
  lisa_result  <- localmoran(analysis_var, lw, zero.policy = TRUE, na.action = na.omit)
  pts$lisa_Ii  <- lisa_result[, "Ii"]
  pts$lisa_p   <- lisa_result[, "Pr(z != E(Ii))"]
  pts$lisa_sig <- pts$lisa_p < 0.05
 
  z_var <- scale(analysis_var)[, 1]
  z_lag <- lag.listw(lw, z_var, zero.policy = TRUE)
 
  pts$lisa_quad <- case_when(
    z_var > 0 & z_lag > 0 & pts$lisa_sig ~ "HH",
    z_var > 0 & z_lag < 0 & pts$lisa_sig ~ "HL",
    z_var < 0 & z_lag > 0 & pts$lisa_sig ~ "LH",
    z_var < 0 & z_lag < 0 & pts$lisa_sig ~ "LL",
    TRUE ~ "Not significant"
  )
 
  cat("LISA quadrant counts:\n")
  print(table(pts$lisa_quad))

  #Print GI Star
  cat("\nSignificant Gi* sites (|Z| > 1.96):\n")
  gi_sig_sites <- pts |>
    st_drop_geometry() |>
    filter(gi_star_sig == TRUE) |>
    arrange(desc(gi_star)) |>
    select(site_name, gi_star) |>
    mutate(type = ifelse(gi_star > 0, "Hot spot", "Cold spot"))
  
  if (nrow(gi_sig_sites) == 0) {
    cat("  No significant Gi* sites\n")
  } else {
    print(gi_sig_sites)
  }

  #Print LISA Clusters
  cat("\nSignificant LISA sites (p < 0.05):\n")
  lisa_sig_sites <- pts |>
    st_drop_geometry() |>
    filter(lisa_sig == TRUE) |>
    arrange(lisa_quad, desc(abs(lisa_Ii))) |>
    select(site_name, lisa_quad, lisa_Ii, lisa_p)
  
  if (nrow(lisa_sig_sites) == 0) {
    cat("  No significant LISA sites\n")
  } else {
    print(lisa_sig_sites)
  }
 
  list(region = res$region, points = pts, moran = moran_result)
}
 
cat("\nResults structure check:\n")
for (reg in names(results)) {
  cat(sprintf("  %s — lw present: %s\n", reg, !is.null(results[[reg]]$lw)))
}

```
### Run Spatial Stats 
```{r}

#Run the Function
stats_results <- map(results, function(res) {
  if (is.null(res)) return(NULL)
  run_spatial_stats(res)
}) |> compact()
```

### Create Maps 

```{r}
#Setting up Colors 
quad_colours <- c("HH" = "#d7191c", "HL" = "#fdae61",
                  "LH" = "#abd9e9", "LL" = "#2c7bb6",
                  "Not significant" = "grey80")

#Looping Through Each Region and Each Process 
for (res in stats_results) {

  pts         <- res$points
  region_name <- res$region

  # Gi* map — significant points only
  pts_sig <- pts |> filter(abs(gi_star) > 1.96)

  p_gi <- ggplot() +
    geom_sf(data = world_proj, fill = "grey85", colour = "white", linewidth = 0.2) +
    geom_sf(data = pts_sig, aes(colour = gi_star), size = 2.5, alpha = 0.9) +
    scale_colour_gradient2(low = "blue", mid = "white", high = "red",
                           midpoint = 0, name = "Gi* Z-score") +
    coord_sf(xlim = st_bbox(pts)[c("xmin","xmax")],
             ylim = st_bbox(pts)[c("ymin","ymax")],
             expand = TRUE) +
    labs(title = paste(region_name, "— Getis-Ord Gi*"),
         subtitle = "Significant clusters only (|Z| > 1.96, p < 0.05)",
         x = NULL, y = NULL) +
    theme_minimal()
  print(p_gi)

  # LISA map
  p_lisa <- ggplot() +
    geom_sf(data = world_proj, fill = "grey85", colour = "white", linewidth = 0.2) +
    geom_sf(data = pts, aes(colour = lisa_quad), size = 2.5, alpha = 0.9) +
    scale_colour_manual(values = quad_colours, name = "LISA Cluster") +
    coord_sf(xlim = st_bbox(pts)[c("xmin","xmax")],
             ylim = st_bbox(pts)[c("ymin","ymax")],
             expand = TRUE) +
    labs(title = paste(region_name, "— LISA Clusters"),
         subtitle = "Local Moran's I, p < 0.05",
         x = NULL, y = NULL) +
    theme_minimal()
  print(p_lisa)
}

```

### Plotting the Regions

```{r}
# ── Projections ──────────────────────────────────────────────────────────────
crs_robin <- "+proj=robin +lon_0=0 +datum=WGS84 +units=m +no_defs"

# ── Base geodata ─────────────────────────────────────────────────────────────
world_wgs84 <- ne_countries(scale = "medium", returnclass = "sf")

# ── Globe outline (bounding box → Robinson) ───────────────────────────────────
globe_outline <- st_as_sfc(
  st_bbox(c(xmin = -180, ymin = -90, xmax = 180, ymax = 90),
          crs = st_crs(4326))
) |>
  st_transform(crs_robin)

# ── Graticule ────────────────────────────────────────────────────────────────
grat <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = seq(-60,   60, by = 30),
  crs = st_crs(4326)
) |>
  st_transform(crs_robin)

# ── Top / bottom boundary lines ───────────────────────────────────────────────
lon_seq <- seq(-180, 180, by = 0.25)

make_parallel <- function(lat) {
  st_sfc(
    st_linestring(cbind(lon_seq, rep(lat, length(lon_seq)))),
    crs = 4326
  ) |>
    st_transform(crs_robin)
}

bottom_boundary <- make_parallel(-90)
top_boundary    <- make_parallel( 90)

# ── Project your data layers to Robinson ─────────────────────────────────────
# (avulsion_sf and region_polys are assumed to be in WGS84 / EPSG:4326)
avulsion_robin  <- st_transform(avulsion_sf,  crs_robin)
region_robin    <- st_transform(region_polys, crs_robin)

# ── Region label centroids (in Robinson space) ────────────────────────────────
region_labels <- region_robin |>
  mutate(
    lon = map_dbl(geometry, ~ st_centroid(.)[[1]]),
    lat = map_dbl(geometry, ~ st_centroid(.)[[2]])
  )

p_overview <- ggplot() +

  # Graticule behind everything
  geom_sf(data = grat,
          color = "grey70", linewidth = 0.25) +

  # Continents
  geom_sf(data = world_wgs84,
          fill = "grey75", colour = "grey60", linewidth = 0.2) +

  # Study-region polygons
  geom_sf(data = region_robin,
          fill = "steelblue", alpha = 0.15,
          colour = "steelblue", linewidth = 0.6) +

  # Avulsion points coloured by p95_zscore
  geom_sf(data = avulsion_robin,
          aes(colour = p95_zscore),
          size = 2, alpha = 0.85) +

  # Top / bottom boundary arcs
  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  # Globe frame
  geom_sf(data = globe_outline,
          fill = NA, colour = "grey70", linewidth = 0.7) +

  # Region labels — increased size 3 → 5
  geom_label_repel(
    data        = region_labels,
    aes(x = lon, y = lat, label = region),
    size        = 5,          # ← was 3
    fill        = "white",
    alpha       = 0.85,
    label.size  = 0.2,
    box.padding = 0.4,
    seed        = 42
  ) +

  # Colour scale
  scale_colour_viridis_c(option = "plasma", name = "p95 Z-score") +

  # Robinson CRS
  coord_sf(crs = crs_robin, expand = FALSE) +

  labs(title = "Avulsion Points – Study Regions") +

  theme_minimal(base_size = 14) +
  theme(
    axis.title        = element_blank(),
    axis.text         = element_blank(),
    axis.ticks        = element_blank(),
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank(),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    plot.title        = element_text(face = "bold", hjust = 0.5),

    legend.position   = c(0.08, 0.25),
    legend.direction  = "vertical",
    legend.title      = element_text(face = "bold", size = 14),  # ← was 9
    legend.text       = element_text(size = 13),                 # ← was 8
    legend.background = element_rect(fill = "white", colour = "white"),
    legend.key.height = unit(1.8, "lines"),                      # ← was 0.6 (taller bar)
    legend.key.width  = unit(1.5, "lines")                       # ← new (wider bar)
  ) +

  guides(
    colour = guide_colorbar(
      barheight    = unit(5, "cm"),   # ← taller colour bar
      barwidth     = unit(0.6, "cm"), # ← wider colour bar
      title.hjust  = 0.5
    )
  )

print(p_overview)

ggsave(
  filename = here("plots_2.0", "study_regions_meshgrid.png"),
  plot = p_overview,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300
)

```

```{r}

# ── Combine all regions into one sf object ────────────────────────────────────
all_pts <- bind_rows(
  lapply(stats_results, function(res) {
    res$points |> mutate(region = res$region)
  })
)

# ── Project to Robinson ───────────────────────────────────────────────────────
all_pts_robin <- st_transform(all_pts, crs_robin)

# ── Colours ───────────────────────────────────────────────────────────────────
quad_colours <- c(
  "HH"              = "#d7191c",
  "HL"              = "#fdae61",
  "LH"              = "#abd9e9",
  "LL"              = "#2c7bb6",
  "Not significant" = "black"    # overridden by alpha below
)

# ══════════════════════════════════════════════════════════════════════════════
# LISA World Map
# ══════════════════════════════════════════════════════════════════════════════

# Split into significant and not-significant
lisa_sig    <- all_pts_robin |> filter(lisa_quad != "Not significant")
lisa_nonsig <- all_pts_robin |> filter(lisa_quad == "Not significant")

p_lisa_world <- ggplot() +

  geom_sf(data = grat,
          color = "grey70", linewidth = 0.25) +

  geom_sf(data = world_wgs84,
          fill = "grey75", colour = "grey60", linewidth = 0.2) +

  # Non-significant points: black, high transparency
  geom_sf(data = lisa_nonsig,
          colour = "black", fill = "black",
          size = 3, alpha = 0.45, shape = 16) +

  # Significant points: quad colours, fully visible
  geom_sf(data = lisa_sig,
          aes(colour = lisa_quad),
          size = 3, alpha = 0.9, shape = 16) +

  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  geom_sf(data = globe_outline,
          fill = NA, colour = "grey70", linewidth = 0.7) +

  scale_colour_manual(
    values = c(
      "HH" = "#d7191c",
      "HL" = "#fdae61",
      "LH" = "#abd9e9",
      "LL" = "#2c7bb6"
    ),
    name   = "LISA Cluster",
    drop   = FALSE
  ) +

  coord_sf(crs = crs_robin, expand = FALSE) +

  theme_minimal(base_size = 14) +
  theme(
    axis.title        = element_blank(),
    axis.text         = element_blank(),
    axis.ticks        = element_blank(),
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank(),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    plot.title        = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle     = element_text(hjust = 0.5, size = 11, color = "grey40"),
    legend.position   = c(0.08, 0.25),
    legend.direction  = "vertical",
    legend.title      = element_text(face = "bold", size = 14),
    legend.text       = element_text(size = 13),
    legend.background = element_rect(fill = "white", colour = "white"),
    legend.key.height = unit(1.8, "lines"),
    legend.key.width  = unit(1.5, "lines")
  ) +

  guides(
    colour = guide_legend(
      override.aes = list(size = 4, alpha = 0.9)
    )
  )

print(p_lisa_world)

ggsave(
  filename = here("plots_2.0", "lisa_world_meshgrid.png"),
  plot = p_lisa_world,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300
)

# ══════════════════════════════════════════════════════════════════════════════
# Gi* World Map
# ══════════════════════════════════════════════════════════════════════════════

# Split significant vs not (|Z| > 1.96)
gi_sig    <- all_pts_robin |> filter(abs(gi_star) > 1.96)
gi_nonsig <- all_pts_robin |> filter(abs(gi_star) <= 1.96)

p_gi_world <- ggplot() +

  geom_sf(data = grat,
          color = "grey70", linewidth = 0.25) +

  geom_sf(data = world_wgs84,
          fill = "grey75", colour = "grey60", linewidth = 0.2) +

  # Non-significant: black, high transparency
  geom_sf(data = gi_nonsig,
          colour = "black", fill = "black",
          size = 3, alpha = 0.45, shape = 16) +

  # Significant: blue–white–red diverging scale
  geom_sf(data = gi_sig,
          aes(colour = gi_star),
          size = 3, alpha = 0.9, shape = 16) +

  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  geom_sf(data = globe_outline,
          fill = NA, colour = "grey70", linewidth = 0.7) +

  scale_colour_gradient2(
    low      = "blue",
    mid      = "white",
    high     = "red",
    midpoint = 0,
    name     = "Gi* Z-score"
  ) +

  coord_sf(crs = crs_robin, expand = FALSE) +

  theme_minimal(base_size = 14) +
  theme(
    axis.title        = element_blank(),
    axis.text         = element_blank(),
    axis.ticks        = element_blank(),
    panel.grid.major  = element_blank(),
    panel.grid.minor  = element_blank(),
    plot.background   = element_rect(fill = "white", color = NA),
    panel.background  = element_rect(fill = "white", color = NA),
    plot.title        = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle     = element_text(hjust = 0.5, size = 11, color = "grey40"),
    legend.position   = c(0.08, 0.25),
    legend.direction  = "vertical",
    legend.title      = element_text(face = "bold", size = 14),
    legend.text       = element_text(size = 13),
    legend.background = element_rect(fill = "white", colour = "white"),
    legend.key.height = unit(1.8, "lines"),
    legend.key.width  = unit(1.5, "lines")
  ) +

  guides(
    colour = guide_colorbar(
      barheight   = unit(5, "cm"),
      barwidth    = unit(0.6, "cm"),
      title.hjust = 0.5
    )
  )

print(p_gi_world)

ggsave(
  filename = here("plots_2.0", "gi_world_meshgrid.png"),
  plot = p_gi_world,
  width = 10,
  height = 7,
  units = "in",
  dpi = 300
)

# ── Optional saves ────────────────────────────────────────────────────────────
# ggsave("fig_lisa_world.png",   p_lisa_world, width = 10, height = 7, dpi = 300)
# ggsave("fig_gi_star_world.png", p_gi_world,  width = 10, height = 7, dpi = 300)

```
