---
title: "Templates for All Maps and Graphs + Descriptive Statistics"
description: "This document is meant to act as the location where all final version graphs are meant to be created, and all descriptive statistics put onto those graphs will be calcualted. This synthesizes material from other documents up to this point. "
author: "Sam Lance"
format:
  html:
    code-fold: true
    toc: true
    number-sections: true
    embed-resources: true
    theme: Lux
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r, include =FALSE}
library(dplyr)
library(tidyverse)
library(lubridate)
library(readr)
library(here)
library(ggplot2)
library(stringr)
library(purrr)
library(maps)
library(knitr)
library(kableExtra)
library(sf)
library(RColorBrewer)
library(patchwork)
library(tmap)
library(viridis)
library(sp)
library(rworldmap)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(devtools)
library(janitor)
library(svglite)
library(gridExtra)
library(grid)
```

## Reading In + Cleaning Data

### Reading in Data

```{r, output=FALSE}}
#DISCHARGE
join_dis <- read_csv(here("data","vg_dis_z.csv")) |>
  inner_join(
    read_csv(here("data","base_sites.csv")) |>
      select(site_name, site_lat, site_lon),
    by = "site_name") |> #og data missing these parts, need to join
  mutate(
    zlog_p95_q  = coalesce(zlog_p95_q, z_p95_q),
    zlog_mean_q = coalesce(zlog_mean_q, z_mean_q)
  ) #coalesce falls back to the second option if the first is not available, doing because issue with intermittent flows at 10 sites 

#REMOVED FOR PAPER- Sediment - 100 Sediment Count
#join_sed = read_csv(here("data","all_site_ssc_z.csv")) |>
  #drop_na()

#Sediment - 62 Sediment Count 
join_sed <- read_csv(here("data","ssc_1km.csv")) |>
  drop_na()



```

### Count by Type

```{r}

count_by_type <- join_dis |>
  group_by(av_type) |>
  summarise(n = n()) |>
  ungroup() 

count_by_type_ssc <- join_sed |>
  group_by(av_type) |>
  summarise(n = n()) |>
  ungroup()

```

### Cleaning Discharge Data

```{r}
#Convert DF to SF Object 
sf_dis <- st_as_sf(join_dis, 
                           coords = c("site_lon", "site_lat"), #locate long and lat in df
                           crs = 4326,
                       remove = FALSE) #setting coordinate reference system WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_dis, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_dis <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

#Export as CSV 
write_csv(sf_dis, "discharge_all_sites.csv")

```

### Cleaning Sediment Data

```{r}

#Convert DF to SF Object 
sf_sed <- st_as_sf(join_sed, coords = c("site_lon", "site_lat"), crs = 4326,remove = FALSE) #setting  WGS84

#Load Data from Natural Earth 
world <- ne_countries(scale = "medium", returnclass = "sf") |> #turn into SF object 
  select(name, continent, geometry)   # keep only country name + continent

#Spatial Join
#Assigning every point a continent based on location
continent_joined <- st_join(sf_sed, world, join = st_within)

#If nothing within then go with the nearest continent 
miss_idx <- which(is.na(continent_joined$continent))
if (length(miss_idx) > 0) {
  nearest_row_in_world <- st_nearest_feature(continent_joined[miss_idx, ], world)
  continent_joined$continent[miss_idx] <- world$continent[nearest_row_in_world]
  continent_joined$name[miss_idx]      <- world$name[nearest_row_in_world]
}

#Final Renaming and Creating New DF 
sf_sed <- continent_joined %>%
  st_drop_geometry() %>%
  rename(country = name)

```

### Stats for Paper

```{r}

# Discharge + SSC Overall
dis_stats <- join_dis |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE),
    mean_log_durz = mean(zlog_flood_dur, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_q, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_q, mu = 0)$p.value,
    p_log_durz = t.test(zlog_flood_dur, mu = 0)$p.value
  )

ssc_stats <- join_sed |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_ssc, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_ssc, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_ssc, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_ssc, mu = 0)$p.value
  )

# Summarize Same Way but for Avulsion Types
dis_stats_type <- join_dis |>
  group_by(av_type) |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_q, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_q, na.rm = TRUE),
    mean_log_durz = mean(zlog_flood_dur, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_q, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_q, mu = 0)$p.value,
    p_log_durz = t.test(zlog_flood_dur, mu = 0)$p.value
  )

ssc_stats_type <- join_sed |>
  group_by(av_type) |>
  summarise(
    n = n(),
    mean_log_95z = mean(zlog_p95_ssc, na.rm = TRUE),
    mean_log_meanz = mean(zlog_mean_ssc, na.rm = TRUE),
    p_log_95z = t.test(zlog_p95_ssc, mu = 0)$p.value,
    p_log_meanz = t.test(zlog_mean_ssc, mu = 0)$p.value
  )
```

### Reading in Full Time Series Data

```{r}

#Load Discharge Data 
discharge_ts <- read_csv(here("data", "vg_dis_z_all_years.csv"))

```

### Pulling out 5 Years Prior to Avulsion

```{r}
#Pull Out 5 Years Prior to Avulsion
pre_avulsion_5 <- discharge_ts |>
  filter(year >= (avulsion_year - 5) & year < avulsion_year) |>
  filter(avulsion_year > 1989)

```

## Figure 1 - World Map with Avulsion Types

### Establishing Colors and Shapes for All Maps

```{r}
#Loading in World Map Data
world_map <- ne_countries(scale = "medium", returnclass = "sf")

shapes <- c(
  "delta"             = 21,
  "fan"               = 22,
  "intra-lobe"        = 23,
  "retrogradational"  = 24
)

colors <- plasma(4)
av_type_colors <- setNames(colors, c("delta", "fan", "intra-lobe", "retrogradational"))
print(av_type_colors)
combined_colors <- c("All Sites" = "grey50", av_type_colors)
print(names(combined_colors))

# #devtools::install_github("an-bui/calecopal")# filled shapes
# colors <- c(
#   "delta"             = "#0D0887FF",
#   "fan"               = "#9C179EFF",
#   "intra-lobe"        = "#F89441FF",
#   "retrogradational"  = "#F0F921FF"
# )
# breaks <- c(-2, -1, 0, 1, 2, 3, 4, 5)
# colors <- brewer.pal(length(breaks) - 1, "RdYlBu")
# rev_colors <- rev(colors)  # Reverse the color order
# shapes <- c(16, 17, 15, 18)  # Different shapes for avulsion types

```

### Robinson Projection World Map

```{r}

# 1) Select sites to highlight
selected_sites <- sf_dis |>
  filter(site_name %in% c("B9", "R3.2", "JA.25.India.1990", "MANAMBOLO.2004"))

# 2) Ensure factor levels consistent
sf_dis <- sf_dis |>
  mutate(av_type = factor(av_type))

# 3) Named color vector for av_type
av_type_colors <- setNames(av_type_colors, levels(sf_dis$av_type))

# 4) Robinson CRS
crs_robin <- "+proj=robin +lon_0=0 +datum=WGS84 +units=m +no_defs"

# 5) Make points sf (critical for correct projection)
sf_dis_sf <- st_as_sf(
  sf_dis,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

selected_sites_sf <- st_as_sf(
  selected_sites,
  coords = c("site_lon", "site_lat"),
  crs = 4326,
  remove = FALSE
)

# 6) Ensure world_map has CRS (if not already)
if (is.na(st_crs(world_map))) st_crs(world_map) <- 4326

# 7) Create a "globe" outline and optional graticule
globe_outline <- st_as_sfc(
  st_bbox(c(xmin = -180, ymin = -90, xmax = 180, ymax = 90), crs = st_crs(4326))
) |>
  st_transform(crs_robin)

# Optional graticule
grat <- st_graticule(
  lon = seq(-180, 180, by = 60),
  lat = seq(-60, 60, by = 30),
  crs = st_crs(4326)
) |>
  st_transform(crs_robin)

# -------------------------
# NEW: Top + bottom boundary (latitude caps)
# -------------------------
lat_min <- -90
lat_max <- 90
lon_seq <- seq(-180, 180, by = 0.25)

make_parallel <- function(lat) {
  st_sfc(
    st_linestring(cbind(lon_seq, rep(lat, length(lon_seq)))),
    crs = 4326
  ) |>
    st_transform(crs_robin)
}

bottom_boundary <- make_parallel(lat_min)
top_boundary    <- make_parallel(lat_max)

# 8) Build plot (Robinson + globe boundary)
world_map_plot <- ggplot() +

  # Light graticule behind everything (optional; remove if you don’t want lines)
  geom_sf(data = grat, color = "grey70", linewidth = 0.25) +

  # Continents
  geom_sf(
    data = world_map,
    fill = "grey70",
    color = "grey60",
    linewidth = 0.25
  ) +

  # All sites
  geom_sf(
    data = sf_dis_sf,
    aes(fill = av_type, shape = av_type, color = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 0.5
  ) +

  # Selected sites: same fill/shape, red outline
  geom_sf(
    data = selected_sites_sf,
    aes(fill = av_type, shape = av_type),
    size = 3,
    alpha = 0.8,
    stroke = 1.0,
    color = "black",
    show.legend = FALSE
  ) +

  # -------------------------
  # NEW: draw top/bottom boundary lines
  # -------------------------
  geom_sf(data = bottom_boundary, color = "grey70", linewidth = 0.7) +
  geom_sf(data = top_boundary,    color = "grey70", linewidth = 0.7) +

  # Globe outline (draw last so it frames everything)
  geom_sf(
    data = globe_outline,
    fill = NA,
    color = "grey70",
    linewidth = 0.7
  ) +

  # Scales
  scale_color_manual(name = "Avulsion Type", values = colors) +
  scale_fill_manual(name = "Avulsion Type", values = colors) +
  scale_shape_manual(name = "Avulsion Type", values = shapes) +

  # Robinson projection
  coord_sf(crs = crs_robin, expand = FALSE) +

  # Styling
  theme_minimal(base_size = 16) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),

    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),

    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    
    #legend.position = "none",
    legend.position = c(0.2, 0.25),
    legend.direction = "vertical",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    legend.title.align = 0.5,
    legend.spacing.y = unit(-0.1, "cm"),
    legend.key.height = unit(0.3, "lines"),
    legend.background = element_rect(fill = "white", color = "white")
  ) +
  guides(
    fill  = guide_legend(byrow = TRUE),
    shape = guide_legend(byrow = TRUE)
  )

world_map_plot

ggsave(
  filename = here("plots_v3", "fig1_world_robinson_globe.svg"),
  plot = world_map_plot,
  width = 10,
  height = 7,
  device = svglite, 
  units = "in",
  dpi = 300
)

```

## Figure 2 - Stacked Time Series Plots

```{r}

# =========================================================
# Figure 2 — Combined SSC + Discharge (Top Panel) — GRL export
# Self-contained: defines mm_to_in() here so this script doesn't
# depend on any other file being sourced first.
# Saves as SVG (not PNG) -- PNG import in Affinity sizes the image
# from pixel-count-at-document-DPI, which is exactly what caused
# the oversized-import bug. SVG carries its physical size directly,
# with no DPI ambiguity.
# =========================================================

# ---- mm -> inches helper (ggsave wants inches by default) ----
mm_to_in <- function(mm) mm / 25.4

# -------------------------
# Paths
# -------------------------
av_year_updated_path <- here("data", "betsiboka_ssc_ts.csv")
compiled_output_path <- here("data", "betsiboka_raw_ssc.csv")
av_year_path_dis     <- here("data", "betsiboka_dis_ts.csv")
compiled_path_dis    <- here("data", "betsiboka_raw_dis.csv")
plots_dir <- here("plots_v3")

# -------------------------
# Load SSC data
# -------------------------
av_year_ssc  <- read_csv(av_year_updated_path, show_col_types = FALSE)
compiled_ssc <- read_csv(compiled_output_path, show_col_types = FALSE) %>%
  mutate(date = as.Date(date))

# -------------------------
# Load Discharge data
# -------------------------
av_year_dis  <- read_csv(av_year_path_dis, show_col_types = FALSE)
compiled_dis <- read_csv(compiled_path_dis, show_col_types = FALSE) %>%
  mutate(valid_time = as.POSIXct(valid_time, tz = "UTC"))

# -------------------------
# SELECT ONE SITE
# -------------------------
site <- "BETSIBOKA.2004"

site_ssc <- compiled_ssc %>% 
  filter(site_name == site) %>% 
  arrange(date)

site_dis <- compiled_dis %>% 
  filter(site_name == site) %>% 
  arrange(valid_time)

stopifnot(nrow(site_ssc) > 0, nrow(site_dis) > 0)

site_dis <- site_dis %>%
  mutate(date = as.Date(valid_time))

# -------------------------
# Avulsion year
# -------------------------
get_avulsion_year <- function(df) {
  if ("avulsion_start_date" %in% names(df)) {
    df %>% pull(avulsion_start_date) %>% as.numeric() %>% na.omit() %>% unique()
  } else {
    numeric(0)
  }
}

avs <- get_avulsion_year(site_ssc)

# -------------------------
# Shading windows
# -------------------------
shade_avulsion  <- NULL
shade_pre       <- NULL

if (length(avs) > 0) {
  av_yr <- avs[1]

  shade_avulsion <- tibble(
    xmin = as.Date(paste0(av_yr - 1, "-01-01")),
    xmax = as.Date(paste0(av_yr,     "-12-31")),
    ymin = -Inf, ymax = Inf
  )

  shade_pre <- tibble(
    xmin = as.Date(paste0(av_yr - 6, "-01-01")),
    xmax = as.Date(paste0(av_yr - 2, "-12-31")),
    ymin = -Inf, ymax = Inf
  )
}

# -------------------------
# p95 points within avulsion window
# -------------------------
get_p95_pt_ssc <- function(df, shade) {
  if (is.null(shade)) return(NULL)
  w <- df %>% filter(date >= shade$xmin, date <= shade$xmax, !is.na(y_pred))
  if (nrow(w) == 0) return(NULL)
  p95_val <- as.numeric(quantile(w$y_pred, 0.95, na.rm = TRUE, type = 7))
  max_row <- w %>% filter(y_pred == max(y_pred, na.rm = TRUE)) %>% arrange(date) %>% slice(1)
  tibble(date = max_row$date, y_pred = p95_val)
}

get_p95_pt_dis <- function(df, shade) {
  if (is.null(shade)) return(NULL)
  w <- df %>% filter(date >= shade$xmin, date <= shade$xmax, !is.na(dis24))
  if (nrow(w) == 0) return(NULL)
  p95_val <- as.numeric(quantile(w$dis24, 0.95, na.rm = TRUE, type = 7))
  max_row <- w %>% filter(dis24 == max(dis24, na.rm = TRUE)) %>% arrange(date) %>% slice(1)
  tibble(date = max_row$date, dis24 = p95_val)
}

p95_pt_ssc <- get_p95_pt_ssc(site_ssc, shade_avulsion)
p95_pt_dis <- get_p95_pt_dis(site_dis, shade_avulsion)

# -------------------------
# Scaling factor for dual y-axis
# -------------------------
ssc_range <- range(site_ssc$y_pred, na.rm = TRUE)
dis_range <- range(site_dis$dis24,  na.rm = TRUE)

scale_dis <- function(x) {
  (x - dis_range[1]) / diff(dis_range) * diff(ssc_range) + ssc_range[1]
}
unscale_dis <- function(x) {
  (x - ssc_range[1]) / diff(ssc_range) * diff(dis_range) + dis_range[1]
}

site_dis <- site_dis %>% mutate(dis_scaled = scale_dis(dis24))

# -------------------------
# Shared shading helper
# -------------------------
shade_layers <- function() {
  list(
    if (!is.null(shade_pre))
      geom_rect(
        data = shade_pre,
        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
        inherit.aes = FALSE,
        fill  = "#fff7bc",
        alpha = 0.6
      ),
    if (!is.null(shade_avulsion))
      geom_rect(
        data = shade_avulsion,
        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
        inherit.aes = FALSE,
        fill  = "#ffea4d",
        alpha = 0.45
      )
  )
}

# -------------------------
# Shared x scale + theme (6pt text per GRL spec; original line weights)
# -------------------------
GRL_BASE_SIZE <- 10
GRL_LINEWIDTH <- 1.2

shared_x <- scale_x_date(
  limits = as.Date(c("1983-01-01", "2020-12-31")),
  breaks = seq(as.Date("1983-01-01"), as.Date("2020-01-01"), by = "5 years"),
  labels = date_format("%Y"),
  expand = c(0, 0)
)

shared_theme <- theme_minimal(base_size = GRL_BASE_SIZE) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = 0.25, color = "grey90"),
    axis.title.y       = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    axis.title.x       = element_blank(),
    axis.line.x        = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    axis.line.y        = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    axis.text.x        = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    axis.text.y        = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    axis.ticks.y       = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    axis.ticks.x       = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    legend.position    = "none",
    plot.background    = element_rect(fill = "white", colour = NA)
  )

# -------------------------
# Combined SSC + Discharge plot (dual y-axis)
# -------------------------
p_combined <- ggplot(site_ssc, aes(x = date, y = y_pred)) +

  shade_layers() +

  geom_line(data  = site_dis, aes(x = date, y = dis_scaled),
            inherit.aes = FALSE, color = "#2f6fa5", linewidth = 0.75) +
  geom_point(data = site_dis, aes(x = date, y = dis_scaled),
             inherit.aes = FALSE, color = "#2f6fa5", size = 0.75) +

  geom_line(color  = "#9bbf30", linewidth = 0.75, alpha = 0.7) +
  geom_point(color = "#9bbf30", size = 0.75, alpha = 0.7) +

  scale_y_continuous(
    name   = "Discharge (m\u00b3/s)",
    expand = c(0, 0.05),
    labels = function(x) round(unscale_dis(x), 0),
    sec.axis = sec_axis(
      transform = ~ .,
      name      = "SSC (mg/L)"
    )
  ) +

  shared_x +
  labs(x = "Year") +
  shared_theme +
  theme(
    plot.margin = margin(t = 1, r = 4, b = 1, l = 1, unit = "mm")
  )

# -------------------------
# Save combined plot -- SVG, not PNG, to avoid DPI-based import sizing
# -------------------------
ggsave(
  filename = file.path(plots_dir, "fig2_combined_timeseries_GRL.svg"),
  plot     = p_combined,
  width    = mm_to_in(170),
  height   = mm_to_in(70),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

p_combined

```

```{r}

# =========================================================
# Figure 2 — Z-score Time Series (Bottom Panel) — GRL export
# Self-contained: defines mm_to_in() and the GRL theme here so this
# script doesn't depend on any other file being sourced first.
# Sized to match the top panel (170 x 45 mm) so the two stack cleanly
# with aligned x-axes.
# ========================================================

# ---- mm -> inches helper (ggsave wants inches by default) ----
mm_to_in <- function(mm) mm / 25.4

# ---- GRL text/line settings (must match the top panel) ----
GRL_BASE_SIZE <- 10      # pt, per AGU/GRL spec -- do not increase
GRL_LINEWIDTH <- 1.2    # axis/tick lines -- original weight
GRL_GRIDWIDTH <- 0.25   # gridlines -- original weight

plots_dir <- here("plots_v3")

# ---- shared theme, redefined here so this script is standalone ----
shared_theme <- theme_minimal(base_size = GRL_BASE_SIZE) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(linetype = "solid", linewidth = GRL_GRIDWIDTH, color = "grey90"),
    axis.title.y       = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    axis.title.x       = element_blank(),
    axis.line.x        = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    axis.line.y        = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    axis.text.x        = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    axis.text.y        = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    axis.ticks.y       = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    axis.ticks.x       = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
    legend.position    = "none",
    plot.background    = element_rect(fill = "white", colour = NA)
  )

# =========================================================
# Data prep
# =========================================================

malawi_dis_zscores <- read_csv(here("data", "betsiboka_dis_ts.csv"))
malawi_ssc_zscores <- read_csv(here("data", "betsiboka_ssc_ts.csv"))

prep_df <- function(df, label) {
  df %>%
    mutate(
      year       = as.integer(year),
      zlog_p95_q = as.numeric(zlog_p95_q),
      dataset    = label,
      date       = as.Date(paste0(year, "-01-01"))
    ) %>%
    filter(!is.na(year), !is.na(zlog_p95_q))
}

dis_red  <- prep_df(malawi_dis_zscores, "Discharge (m^3/s)")
ssc_blue <- prep_df(malawi_ssc_zscores, "SSC (mg/L)")
df_all   <- bind_rows(dis_red, ssc_blue)
stopifnot(nrow(df_all) > 0)

# ---- Shading windows ----
av_yr_z <- 2004L

shade_avulsion_z <- tibble(
  xmin = as.Date(paste0(av_yr_z - 1, "-01-01")),
  xmax = as.Date(paste0(av_yr_z,     "-12-31")),
  ymin = -Inf, ymax = Inf
)

shade_pre_z <- tibble(
  xmin = as.Date(paste0(av_yr_z - 6, "-01-01")),
  xmax = as.Date(paste0(av_yr_z - 2, "-12-31")),
  ymin = -Inf, ymax = Inf
)

# ---- Marker points ----
mark_year_dis <- 2004L
mark_year_ssc <- 2003L

mark_pts_dis <- df_all %>% filter(dataset == "Discharge (m^3/s)", year == mark_year_dis)
mark_pts_ssc <- df_all %>% filter(dataset == "SSC (mg/L)",        year == mark_year_ssc)

# =========================================================
# Plot
# =========================================================

p_z <- ggplot(df_all, aes(x = date, y = zlog_p95_q, color = dataset)) +

  geom_rect(
    data = shade_pre_z,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill  = "#fff7bc",
    alpha = 0.55
  ) +

  geom_rect(
    data = shade_avulsion_z,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill  = "#ffea4d",
    alpha = 0.45
  ) +

  geom_line(linewidth = 0.9, alpha = 0.7) +
  geom_point(size = 1.8, alpha = 0.7) +
  geom_hline(color = "grey30", linewidth = GRL_LINEWIDTH, yintercept = 0, linetype = "dashed") +
  
  coord_cartesian(ylim = c(-3, 2)) +
  

  { if (nrow(mark_pts_dis) > 0)
    geom_point(data = mark_pts_dis, aes(x = date, y = zlog_p95_q),
               inherit.aes = FALSE, shape = 21, size = 2.5, stroke = 1,
               fill = NA, color = "red", alpha = 0.8)
  } +

  { if (nrow(mark_pts_ssc) > 0)
    geom_point(data = mark_pts_ssc, aes(x = date, y = zlog_p95_q),
               inherit.aes = FALSE, shape = 21, size = 2.5, stroke = 1,
               fill = NA, color = "red", alpha = 0.8)
  } +

  scale_color_manual(values = c(
    "Discharge (m^3/s)" = "#2f6fa5",
    "SSC (mg/L)"        = "#9bbf30"
  )) +

  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +

  scale_x_date(
    limits = c(as.Date("1983-01-01"), as.Date("2020-12-31")),
    breaks = seq(as.Date("1983-01-01"), as.Date("2020-01-01"), by = "5 years"),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +

  labs(x = "Year", y = "Z-Score (-)") +

  shared_theme +
  theme(
    # 6pt everywhere, matching the top panel -- no leftover 20/16/14 values
    axis.title.x      = element_text(size = GRL_BASE_SIZE, face = "bold", color = "grey30"),
    axis.title.y      = element_text(size = GRL_BASE_SIZE, face = "bold", color = "grey30"),
    axis.text         = element_text(size = GRL_BASE_SIZE, color = "grey30", face = "bold"),
    legend.position   = "none",
    legend.direction  = "horizontal",
    legend.title      = element_blank(),
    legend.text       = element_text(size = GRL_BASE_SIZE, face = "bold", color = "grey30"),
    legend.background = element_rect(fill = scales::alpha("white", 0.7), colour = NA),
    legend.key        = element_rect(fill = NA, colour = NA),
    # Same right margin as the top panel, so the two x-axes (Year)
    # align vertically once stacked in Affinity
    plot.margin = margin(t = 1, r = 15, b = 1, l = 3, unit = "mm")
  )

# =========================================================
# Save -- 170 x 45 mm, matching the top panel exactly
# =========================================================
ggsave(
  filename = file.path(plots_dir, "fig2_z_timeseries_GRL.svg"),
  plot     = p_z,
  width    = mm_to_in(170),
  height   = mm_to_in(70),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

p_z

```

## Figure 3- Boxplots

```{r}

# =========================================================
# Figure 3 — Boxplots of Discharge/SSC Z-Score — GRL export
# =========================================================
.objs_to_clear <- c("make_panel", "grl_theme_classic", "p_p95_q", "p_mean_q",
                     "p_flood_dur", "ssc_overall", "combined_grob")
rm(list = Filter(exists, .objs_to_clear))
rm(.objs_to_clear)

mm_to_in <- function(mm) mm / 25.4

plots_dir <- here("plots_v3")

# =========================================================
# GRL text/line constants — the ONLY place these are defined
# =========================================================
GRL_BASE_SIZE  <- 10      # pt, AGU/GRL spec baseline
GRL_TITLE_SIZE <- 10      # axis titles
GRL_XTEXT_SIZE <- 10     # x-axis tick text
GRL_YTEXT_SIZE <- 10     # y-axis tick text -- explicit override
GRL_SIG_SIZE   <- 12     # significance stars (pt) -- explicit override
GRL_LINEWIDTH  <- 1.2    # axis/tick lines

# =========================================================
# Theme — defined ONCE. Explicit element_text() on every text slot,
# nothing left to inherit from theme_classic()'s defaults, so there is
# no rel()-multiplier ambiguity and no possibility of a later +theme()
# silently winning over this on size alone.
# =========================================================
grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) %+replace%
    theme(
      axis.title.y     = element_text(color = "grey30", size = GRL_TITLE_SIZE,
                                       face = "bold", angle = 90, margin = margin(r = 2)),
      axis.title.x     = element_blank(),
      axis.line.x      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x      = element_text(color = "grey30", size = GRL_XTEXT_SIZE, face = "bold"),
      axis.text.y      = element_text(color = "grey30", size = GRL_YTEXT_SIZE, face = "bold"),
      axis.ticks.y     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",
      plot.background  = element_rect(fill = "white", colour = NA),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.margin      = margin(t = 1.5, r = 1.5, b = 1.5, l = 1.5, unit = "mm"))
}

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar <- function(p) {
  if (p < 0.001)     "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else               "ns"
}

# ── HELPER: BUILD PLOT DATA + SIG LABELS FOR ONE METRIC ──────────────────
# (av_type_colors / combined_colors dead-code at top level removed --
#  this function already builds its own colors internally from plasma(4))
make_panel <- function(data, metric_col, y_label, exclude_types = NULL,
                       all_sites_color = "grey40") {

  if (!is.null(exclude_types)) {
    data <- data |> filter(!av_type %in% exclude_types)
  }

  all_sites <- data |> mutate(av_type_plot = "All Sites")
  by_type   <- data |> mutate(av_type_plot = as.character(av_type))

  remaining_types <- unique(as.character(data$av_type))
  all_levels <- c("All Sites", intersect(c("delta", "fan", "intra-lobe", "retrogradational"), remaining_types))

  plot_data <- bind_rows(all_sites, by_type) |>
    mutate(av_type_plot = factor(av_type_plot, levels = all_levels))

  t_all     <- t.test(data[[metric_col]], mu = 0)
  t_by_type <- lapply(remaining_types, function(typ) {
    t.test(data |> filter(av_type == typ) |> pull(metric_col), mu = 0)
  })
  names(t_by_type) <- remaining_types

  sig_labels <- data.frame(
    av_type_plot = factor(all_levels, levels = all_levels),
    label = c(
      pstar(t_all$p.value),
      sapply(remaining_types, function(typ) pstar(t_by_type[[typ]]$p.value))
    )
  )

  y_positions <- plot_data |>
    group_by(av_type_plot) |>
    summarise(y_pos = max(.data[[metric_col]], na.rm = TRUE) + 0.2)

  sig_labels <- sig_labels |> left_join(y_positions, by = "av_type_plot")

  base_colors <- c("delta" = plasma(4)[1], "fan" = plasma(4)[2],
                   "intra-lobe" = plasma(4)[3], "retrogradational" = plasma(4)[4])
  combined_colors <- c("All Sites" = all_sites_color, base_colors[remaining_types])

  ggplot(plot_data, aes(x = av_type_plot, y = .data[[metric_col]],
                         fill = av_type_plot, color = av_type_plot)) +

    geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
    geom_boxplot(show.legend = FALSE, alpha = 0.4,
                 outlier.shape = NA, color = "grey30", linewidth = 0.7) +
    stat_summary(fun = mean, geom = "crossbar", width = 0.5,
                 color = scales::alpha("red", 0.50), linewidth = 0.3) +
    geom_hline(yintercept = 0, linetype = "dashed",
               color = "grey30", linewidth = GRL_LINEWIDTH) +
    geom_vline(xintercept = 1.5, linetype = "dotted",
               color = "grey50", linewidth = 0.4) +
    geom_text(data = sig_labels,
              aes(x = av_type_plot, y = y_pos, label = label),
              inherit.aes = FALSE, color = "grey30", size = GRL_SIG_SIZE / .pt,
              fontface = "bold") +
    scale_fill_manual(values = combined_colors) +
    scale_color_manual(values = combined_colors) +
    scale_x_discrete(labels = c(
      "All Sites"        = "All Sites",
      "delta"            = "Delta",
      "fan"              = "Fan",
      "intra-lobe"       = "Intra-lobe",
      "retrogradational" = "Retro."
    )) +
    labs(y = y_label) +
    grl_theme_classic()
}

# ── ALL SITES GREYS — one distinct shade per discharge panel ─────────────
# These are also reused for the SSC panel's Peak/Mean boxes below, so
# the P95 and Mean greys visually "translate" across both figures.
GREY_P95   <- "grey75"   # darkest -- P95 / Peak
GREY_MEAN  <- "grey55"   # mid -- Mean
GREY_FLOOD <- "grey25"   # lightest -- Flood Duration (no SSC equivalent)

# ── BUILD THE 4 PANELS FOR THE GRID ───────────────────────────────────────
p_p95_q     <- make_panel(join_dis, "zlog_p95_q",     "P95 Discharge Z (-)",
                           all_sites_color = GREY_P95)
p_mean_q    <- make_panel(join_dis, "zlog_mean_q",    "Mean Discharge Z (-)",
                           all_sites_color = GREY_MEAN)
p_flood_dur <- make_panel(join_dis, "zlog_flood_dur", "Flood Duration Z (-)",
                           all_sites_color = GREY_FLOOD)

# ── SSC OVERALL PANEL (Peak vs Mean) ──────────────────────────────────────
plot_data_ssc <- join_sed |>
  select(zlog_p95_ssc, zlog_mean_ssc) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    variable = case_when(
      variable == "zlog_p95_ssc"  ~ "Peak",
      variable == "zlog_mean_ssc" ~ "Mean"
    ),
    variable = factor(variable, levels = c("Peak", "Mean"))
  )

t_p95  <- t.test(join_sed$zlog_p95_ssc,  mu = 0)
t_mean <- t.test(join_sed$zlog_mean_ssc, mu = 0)

y_positions_ssc <- plot_data_ssc |>
  group_by(variable) |>
  summarise(y_pos = max(value, na.rm = TRUE) + 0.2)

sig_labels_ssc <- data.frame(
  variable = factor(c("Peak", "Mean"), levels = c("Peak", "Mean")),
  label = c(pstar(t_p95$p.value), pstar(t_mean$p.value))
) |>
  left_join(y_positions_ssc, by = "variable")

ssc_overall <- plot_data_ssc |>
  ggplot(aes(x = variable, y = value, fill = variable, color = variable)) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 1) +
  geom_boxplot(width = 0.35, alpha = 0.6, outlier.shape = NA,
               color = "grey30", linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "gray30", linewidth = GRL_LINEWIDTH) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.35,
               color = scales::alpha("red", 0.50), linewidth = 0.3) +
  geom_text(data = sig_labels_ssc,
            aes(x = variable, y = y_pos, label = label),
            inherit.aes = FALSE, color = "grey30", size = GRL_SIG_SIZE / .pt,
            fontface = "bold") +
  scale_fill_manual(values  = c("Peak" = GREY_P95, "Mean" = GREY_MEAN)) +
  scale_color_manual(values = c("Peak" = GREY_P95, "Mean" = GREY_MEAN)) +
  labs(x = NULL, y = "SSC Z-Score (-)") +
  grl_theme_classic() +
  theme(axis.title.x = element_text(color = "grey30", size = GRL_TITLE_SIZE, face = "bold"))

# =========================================================
# Assemble 2x2 grid: p95_q | mean_q
#                    flood_dur | ssc_overall
# Final size: 170 mm wide x 170 mm tall -- square overall grid,
# each cell ~85 x 85 mm with 4mm internal padding per panel
# =========================================================

combined_grob <- arrangeGrob(
  p_p95_q, p_mean_q,
  p_flood_dur, ssc_overall,
  ncol = 2, nrow = 2
)

svglite(
  file.path(plots_dir, "fig3_grid_GRL.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(130),
  bg     = "white"
)
grid.newpage()
grid.draw(combined_grob)
dev.off()

grid.newpage()
grid.draw(combined_grob)

# =========================================================
# Affinity import notes:
# - Place the .svg (not a .png) at exactly 170 x 170 mm (100% scale).
# - Check the W field after placing -- switch units to mm if showing
#   px, and type 170 manually if it didn't auto-size correctly.
# - Text is 6pt per AGU/GRL spec, EXCEPT y-axis text which is set to
#   10pt (explicit deviation from the strict 6pt minimum, per request).
# - Line/point/jitter weights were kept close to original visual scale.
# - Each panel has 4mm internal padding so plotted content doesn't
#   touch the cell edges in the 2x2 grid.
# - For final submission, export the assembled figure from Affinity
#   as PDF or EPS with fonts outlined/converted to curves.
# =========================================================

```

## Figure 4 - Avulsion Leadup

```{r}

# =========================================================
# Figure 4 — Leadup to Avulsions — GRL-compliant export
# Target: 170 mm wide, 6 pt text at final size (per AGU/GRL spec)
# 3 panels side by side (~55 mm each + gutters), shared bottom legend
# Line/point weights kept close to original visual weight (not capped by
# the 6pt rule, which applies to TEXT only).
#
# ADDED: per-(av_type, years_before) one-sample t-tests (mean vs 0),
# with significance stars colored to match each av_type's line color
# (same `colors` discrete scale used by the lines/points). Error bars
# remain STANDARD ERROR of the mean (sd/sqrt(n)), same as originally.
# =========================================================

library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(svglite)

mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL settings ----
GRL_BASE_SIZE  <- 10      # pt, per AGU/GRL spec -- do not increase
GRL_LINEWIDTH  <- 1.2    # axis lines/ticks -- original weight
GRL_LINE_LW    <- 2      # geom_line weight -- original
GRL_HLINE_LW   <- 1.2    # dashed zero-line -- original
GRL_ERRBAR_LW  <- 0.6    # error bar weight -- original
GRL_POINT_SIZE <- 5      # point size -- original
GRL_POINT_STROKE <- 1    # point stroke -- original
GRL_SIG_SIZE   <- 12     # significance star size (pt)
SIG_GAP        <- 0.04   # vertical gap between error bar tip and star

grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) +
    theme(
      axis.title.y = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.title.x = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.line.x  = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y  = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x  = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.text.y  = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.ticks.y = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",
      plot.background  = element_rect(fill = "transparent", colour = NA),
      plot.margin      = margin(2, 2, 2, 2)
    )
}

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

plots_dir <- here("plots_v3")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
# Vectorized: case_when() evaluates element-wise across a whole column,
# unlike a plain if/else function (which only accepts a single scalar
# and errors with "condition has length > 1" inside mutate()).
pstar_vec <- function(p) {
  dplyr::case_when(
    is.na(p)     ~ "",
    p < 0.001    ~ "***",
    p < 0.01     ~ "**",
    p < 0.05     ~ "*",
    TRUE         ~ "ns"
  )
}

# ── REUSABLE HELPER: per-(av_type, years_before) t-tests + star placement ──
build_sig_data <- function(value_col, summary_df, mean_col, se_col) {
  value_sym <- rlang::sym(value_col)
  mean_sym  <- rlang::sym(mean_col)
  se_sym    <- rlang::sym(se_col)

  pre_avulsion_5 %>%
    mutate(years_before = avulsion_year - year) %>%
    group_by(av_type, years_before) %>%
    summarise(
      p_value = {
        vals <- (!!value_sym)[!is.na(!!value_sym)]
        if (length(vals) >= 2) t.test(vals, mu = 0)$p.value else NA_real_
      },
      .groups = "drop"
    ) %>%
    mutate(sig_label = pstar_vec(p_value)) %>%
    # drop "ns" so the panel only shows stars for actual significance --
    # remove this filter if you'd rather see "ns" marked explicitly too
    filter(sig_label != "" & sig_label != "ns") %>%
    left_join(summary_df, by = c("av_type", "years_before")) %>%
    mutate(
      star_y = ifelse(
        !!mean_sym >= 0,
        !!mean_sym + !!se_sym + SIG_GAP,
        !!mean_sym - !!se_sym - SIG_GAP
      )
    )
}

# =========================================================
# Panel 1 — Log P95 Annual Discharge
# =========================================================

plot_data <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_p95_q = mean(zlog_p95_q, na.rm = TRUE),
    se_zlog_p95_q   = sd(zlog_p95_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_p95_q))),
    n               = sum(!is.na(zlog_p95_q)),
    .groups = "drop"
  ) %>%
  mutate(
    ci_mult  = qt(0.975, df = pmax(n - 1, 1)),
    ci_lower = mean_zlog_p95_q - ci_mult * se_zlog_p95_q,
    ci_upper = mean_zlog_p95_q + ci_mult * se_zlog_p95_q
  )

sig_data_p95 <- build_sig_data("zlog_p95_q", plot_data, "mean_zlog_p95_q", "se_zlog_p95_q")

p95_line <- ggplot(plot_data, aes(x = years_before, y = mean_zlog_p95_q,
                                   color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_p95_q - se_zlog_p95_q,
                     ymax = mean_zlog_p95_q + se_zlog_p95_q),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.5) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  # significance stars -- colored per av_type to match that line, using
  # the same discrete color scale as the rest of the panel
  geom_text(
    data = sig_data_p95,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.6, 0.6)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Peak Discharge Z (-)",
    color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type"
  ) +
  grl_theme_classic()

# =========================================================
# Panel 2 — Log Mean Annual Discharge
# =========================================================

plot_data1 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_mean_q = mean(zlog_mean_q, na.rm = TRUE),
    se_zlog_mean_q   = sd(zlog_mean_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_mean_q))),
    .groups = "drop"
  )

sig_data_mean <- build_sig_data("zlog_mean_q", plot_data1, "mean_zlog_mean_q", "se_zlog_mean_q")

mean_line <- ggplot(plot_data1, aes(x = years_before, y = mean_zlog_mean_q,
                                     color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_mean_q - se_zlog_mean_q,
                     ymax = mean_zlog_mean_q + se_zlog_mean_q),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.5) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  geom_text(
    data = sig_data_mean,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.6, 0.6)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Mean Discharge Z (-)",
    color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type"
  ) +
  grl_theme_classic()

# =========================================================
# Panel 3 — Log Flood Duration
# =========================================================

plot_data2 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_flood_dur = mean(zlog_flood_dur, na.rm = TRUE),
    se_zlog_flood_dur   = sd(zlog_flood_dur, na.rm = TRUE) / sqrt(sum(!is.na(zlog_flood_dur))),
    .groups = "drop"
  )

sig_data_flood <- build_sig_data("zlog_flood_dur", plot_data2, "mean_zlog_flood_dur", "se_zlog_flood_dur")

flooddur_line <- ggplot(plot_data2, aes(x = years_before, y = mean_zlog_flood_dur,
                                         color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_flood_dur - se_zlog_flood_dur,
                     ymax = mean_zlog_flood_dur + se_zlog_flood_dur),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.6) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  geom_text(
    data = sig_data_flood,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.6, 0.6)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Flood Duration Z (-)",
    color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type"
  ) +
  grl_theme_classic()

# =========================================================
# Shared legend (extracted, centered, bottom row)
# =========================================================

legend_source <- ggplot(plot_data, aes(x = years_before, y = mean_zlog_p95_q,
                                        color = av_type, fill = av_type, shape = av_type)) +
  geom_point(size = GRL_POINT_SIZE, alpha = 0.7) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  guides(color = guide_legend(override.aes = list(size = GRL_POINT_SIZE), nrow = 1)) +
  labs(color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type") +
  theme_classic(base_size = GRL_BASE_SIZE) +
  theme(
    legend.position      = "bottom",
    legend.justification = "center",
    legend.text          = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    legend.title         = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
    legend.key.size      = unit(0.8, "lines")
  )

g <- ggplotGrob(legend_source)
legend_grob <- g$grobs[[which(sapply(g$grobs, function(x) x$name) == "guide-box")]]

# =========================================================
# Assemble: 3 panels in a row + legend strip below
# Final size: 170 mm wide x 100 mm tall (panels + legend)
# =========================================================

combined_grob <- arrangeGrob(
  arrangeGrob(p95_line, mean_line, flooddur_line, ncol = 3),
  legend_grob,
  nrow = 2,
  heights = c(5, 1)
)

svglite(
  file.path(plots_dir, "fig4_combined_GRL.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(65),
  bg     = "white"
)
grid.newpage()
grid.draw(combined_grob)
dev.off()

grid.newpage()
grid.draw(combined_grob)

# =========================================================
# Notes for Affinity assembly:
# - Place this SVG at exactly 170 x 100 mm (100% scale) -- do not resize
#   further, or the 6pt text drifts off the GRL spec.
# - Line/point weights were kept at original visual scale on purpose;
#   only TEXT is constrained to 6pt by the AGU/GRL rule.
# - Significance stars: one-sample t-test (mean vs 0) per av_type per
#   years_before value, colored to match that av_type's line via the
#   same `colors` scale. "ns" results are filtered out -- only
#   significant points (p < 0.05) get a star. Remove the
#   filter(sig_label != "ns") line in build_sig_data() if you want
#   "ns" explicitly labeled too.
# - Error bars remain STANDARD ERROR of the mean (sd/sqrt(n)), matching
#   the original convention -- consistent with what the t-tests use
#   internally, so overlapping-zero bars roughly track non-significant
#   points.
# - For final submission, export the assembled figure from Affinity as
#   PDF or EPS with fonts outlined/converted to curves (AGU requirement).
# =========================================================

```

## Figure 5 - All in One

```{r}

# =========================================================
# Figure 5 — GRL-compliant export
# Target: 170 mm wide, 6 pt text at final size (per AGU/GRL spec)
# Panel A: full width, 170 x 55 mm
# Panels B & C: side by side, 82 x 70 mm each (6 mm gutter)
# =========================================================

# ---- helper: mm -> inches for ggsave ----
mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL theme settings ----
# NOTE: GRL_BASE_SIZE must stay at 6pt to satisfy the AGU/GRL text-size
# requirement at final print size. Line widths, point sizes, and gridlines
# are NOT covered by that rule and can be kept close to original/bold
# values for visual punch without violating the spec.
GRL_BASE_SIZE   <- 10      # pt, per AGU/GRL spec -- do not increase
GRL_LINEWIDTH   <- 1.2    # axis/tick lines -- back to original weight
GRL_GRIDWIDTH   <- 0.3    # gridlines, slightly lighter than axis lines
GRL_POINT_SIZE  <- 3      # scatter point size -- back to original
GRL_SMOOTH_LW   <- 1.5    # trend line width -- back to original

grl_theme <- function(legend = TRUE) {
  theme_minimal(base_size = GRL_BASE_SIZE) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_line(linetype = "solid", linewidth = GRL_GRIDWIDTH, color = "grey90"),
      axis.title.y  = element_text(face = "bold", color = "grey30", size = GRL_BASE_SIZE,
                                    margin = margin(r = 2)),
      axis.title.x  = element_text(face = "bold", color = "grey30", size = GRL_BASE_SIZE,
                                    margin = margin(t = 2)),
      axis.text     = element_text(face = "bold", color = "grey30", size = GRL_BASE_SIZE),
      axis.line     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks    = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      plot.margin   = margin(t = 1, r = 2, b = 1, l = 2),
      legend.text   = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      legend.title  = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      legend.key.size = unit(0.6, "lines"),
      legend.margin   = margin(1, 1, 1, 1),
      legend.background = element_rect(fill = alpha("white", 0.7), color = NA),
      plot.background   = element_rect(fill = "white", colour = NA)
    )
}

# =========================================================
# PANEL A — Avulsion counts by site/year (stacked bars)
# =========================================================


# -------------------------
# Prep avulsion counts
# -------------------------
avulsion_counts <- join_dis %>%
  filter(!is.na(avulsion_year), !is.na(site_area)) %>%
  mutate(year = as.integer(avulsion_year)) %>%
  filter(site_area %in% c("Himalaya", "Pacific Islands", "South America")) %>%
  group_by(site_area, year) %>%
  summarise(n = n(), .groups = "drop")

all_sites <- join_dis %>%
  filter(!is.na(avulsion_year)) %>%
  mutate(year = as.integer(avulsion_year)) %>%
  group_by(year) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(site_area = "All Sites")

# -------------------------
# Poisson test per year
# -------------------------
# Total number of years observed
n_years  <- nrow(all_sites)
# Mean annual count = background Poisson rate
lambda   <- mean(all_sites$n)
mean_all <- lambda

# Run poisson.test() for each year's count against the background rate
# T = n_years so the expected value is lambda * 1 (per year)
poisson_results <- all_sites %>%
  rowwise() %>%
  mutate(
    p_value = poisson.test(n, T = 1, r = lambda)$p.value,
    sig     = p_value < 0.05,
    sig_label = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) %>%
  ungroup() %>%
  filter(sig)

# ---- Combine for plotting ----
df_plot <- bind_rows(avulsion_counts, all_sites)

site_areas <- c("South America", "Himalaya", "Pacific Islands")
all_levels <- c(site_areas, "All Sites")
df_plot    <- df_plot %>% mutate(site_area = factor(site_area, levels = all_levels))

n_sites <- length(site_areas)


df_stack <- df_plot %>%
  filter(site_area != "All Sites") %>%
  mutate(site_area = droplevels(factor(site_area)))

site_levels <- sort(unique(as.character(df_stack$site_area)))
grey_vals <- setNames(
  grey.colors(length(site_levels), start = 0.75, end = 0.15),
  site_levels
)
df_stack <- df_stack %>%
  mutate(site_area = factor(site_area, levels = site_levels))

df_totals <- df_stack %>%
  group_by(year) %>%
  summarise(n = sum(n), .groups = "drop")

poisson_results <- poisson_results %>%
  select(year, sig_label) %>%
  left_join(df_totals, by = "year")

p_av <- ggplot(
  df_stack,
  aes(x = year, y = n, fill = site_area, group = site_area)
) +
  geom_col(width = 0.8, color = "white", linewidth = 0.2, alpha = 0.7) +
  scale_fill_manual(values = grey_vals, name = NULL) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  scale_x_continuous(
    breaks = seq(1984, 2018, by = 4),
    limits = c(1983, 2019),
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.0, 0.3))) +
  labs(x = "Year", y = "Avulsion\nCount") +
  grl_theme() +
  theme(
    legend.position      = c(0.5, 0.85),
    legend.justification = c(0.5, 1)
  )

ggsave(
  filename = file.path(plots_dir, "fig5_avulsion_counts_by_site_v2.svg"),
  plot     = p_av,
  width    = mm_to_in(170),
  height   = mm_to_in(75),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

# =========================================================
# PANELS B & C — Beta / Gamma vs discharge Z-score
# =========================================================

alpha_beta <- read.csv(here("data", "nature_alpha_beta.csv")) |>
  clean_names() |>
  dplyr::select(avulsion_name, geomorphology, gamma, a, b, beta) |>
  rename(site_name = avulsion_name) |>
  mutate(site_name = str_replace_all(site_name, "_", "."))

alpha_beta_av <- inner_join(join_dis, alpha_beta, by = c("site_name"))

alpha_beta_stats <- alpha_beta_av |>
  group_by(av_type) |>
  summarise(n = n()) |>
  ungroup()

corr_beta  <- cor.test(alpha_beta_av$zlog_mean_q, alpha_beta_av$beta)
corr_gamma <- cor.test(alpha_beta_av$zlog_mean_q, alpha_beta_av$gamma)
print(corr_beta)
print(corr_gamma)

# ---- Panel B: Beta ----
trend_colors <- scales::alpha(colors, 0.7)   # bake 50% transparency into trend-line colors

beta_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = beta, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = GRL_SMOOTH_LW) +
  geom_point(size = GRL_POINT_SIZE, stroke = 0.1, color = "white", alpha = 0.7) +
  labs(
    x = "Mean Discharge Z (-)",
    y = "Beta (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = trend_colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  grl_theme() +
  theme(
    legend.position  = "none",
    legend.direction = "horizontal"
  )

ggsave(
  filename = file.path(plots_dir, "fig5_beta.svg"),
  plot     = beta_dis,
  width    = mm_to_in(82),
  height   = mm_to_in(55),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

# ---- Panel C: Gamma ----
gamma_dis <- alpha_beta_av |>
  ggplot(aes(x = zlog_mean_q, y = gamma, color = av_type, fill = av_type, shape = av_type)) +
  geom_smooth(method = "lm", se = FALSE, linewidth = GRL_SMOOTH_LW) +
  geom_point(size = GRL_POINT_SIZE, stroke = 0.1, color = "white", alpha = 0.7) +
  labs(
    x = "Mean Discharge Z (-)",
    y = "Gamma (-)",
    color = "Avulsion type",
    fill  = "Avulsion type",
    shape = "Avulsion type"
  ) +
  scale_color_manual(values = trend_colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  grl_theme() +
  theme(
    legend.position  = "none",
    legend.direction = "horizontal"
  )

ggsave(
  filename = file.path(plots_dir, "fig5_gamma.svg"),
  plot     = gamma_dis,
  width    = mm_to_in(82),
  height   = mm_to_in(55),
  dpi      = 600,
  device   = svglite,
  bg       = "white"
)

```

## Supplemental

### Discharge vs. SSC

```{r}

# =========================================================
# Discharge vs SSC correlation (Pearson)
# Joins join_dis (discharge) and join_sed (SSC) by site_name, then
# correlates the summary z-scores -- one point per avulsion site.
# =========================================================

library(dplyr)
library(ggplot2)

# -------------------------
# Join discharge and SSC summary stats by site
# -------------------------
dis_ssc_joined <- inner_join(
  join_dis %>% select(site_name, av_type, zlog_mean_q, zlog_p95_q),
  join_sed %>% select(site_name, zlog_mean_ssc, zlog_p95_ssc),
  by = "site_name"
)

cat("Sites with both discharge and SSC data:", nrow(dis_ssc_joined), "\n\n")

# -------------------------
# Pearson correlations
# -------------------------
corr_mean <- cor.test(dis_ssc_joined$zlog_mean_q, dis_ssc_joined$zlog_mean_ssc,
                       method = "pearson")


corr_p95  <- cor.test(dis_ssc_joined$zlog_p95_q, dis_ssc_joined$zlog_p95_ssc,
                       method = "pearson")


cat("Mean Discharge Z vs Mean SSC Z:\n")
print(corr_mean)

cat("\nP95 Discharge Z vs P95 SSC Z:\n")
print(corr_p95)

# -------------------------
# Scatter plot: Mean Discharge Z vs Mean SSC Z
# -------------------------
r_label <- sprintf("r = %.2f, p = %.3g", corr_mean$estimate, corr_mean$p.value)

p_corr <- ggplot(dis_ssc_joined, aes(x = zlog_mean_q, y = zlog_mean_ssc)) +
  geom_point(size = 3, alpha = 0.6, color = "#2f6fa5") +
  geom_smooth(method = "lm", se = TRUE, color = "#9bbf30", fill = "#9bbf30", alpha = 0.15) +
  annotate("text", x = -Inf, y = Inf, label = r_label,
           hjust = -0.1, vjust = 1.5, size = 5, fontface = "bold", color = "grey30") +
  labs(
    x = "Mean Discharge Z (-)",
    y = "Mean SSC Z (-)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title = element_text(color = "grey30", face = "bold"),
    axis.text  = element_text(color = "grey30", face = "bold")
  )

p_corr

```

## Getting Freaky 

### Leadup Mean Only 

```{r}
# =========================================================
# Figure 4 — Leadup to Avulsions, Mean Discharge only — GRL export
# Single panel (was 3-panel + shared legend). Adds per-(av_type,
# years_before) t-tests against 0, with significance stars colored
# to match each av_type's line color, positioned above/below the
# error bar depending on the sign of the mean. No legend.
# =========================================================

library(ggplot2)
library(dplyr)
library(svglite)

mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL settings ----
GRL_BASE_SIZE    <- 10     # pt
GRL_LINEWIDTH    <- 1.2    # axis lines/ticks
GRL_LINE_LW      <- 2      # geom_line weight
GRL_HLINE_LW     <- 1.2    # dashed zero-line
GRL_ERRBAR_LW    <- 0.6    # error bar weight
GRL_POINT_SIZE   <- 5      # point size
GRL_POINT_STROKE <- 1      # point stroke
GRL_SIG_SIZE     <- 14     # significance star size (pt)
SIG_GAP          <- 0.04   # vertical gap between error bar tip and star

grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) +
    theme(
      axis.title.y     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.title.x     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.line.x      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.text.y      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.ticks.y     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",   # legend removed
      plot.background  = element_rect(fill = "transparent", colour = NA),
      plot.margin      = margin(4, 4, 4, 4, unit = "mm")
    )
}

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

plots_dir <- here("plots_v3")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
# Vectorized version: case_when() evaluates element-wise across a whole
# column, unlike a plain if/else function (which only accepts a single
# scalar and errors with "condition has length > 1" when handed a vector
# inside mutate()/ifelse()).
pstar_vec <- function(p) {
  dplyr::case_when(
    is.na(p)     ~ "",
    p < 0.001    ~ "***",
    p < 0.01     ~ "**",
    p < 0.05     ~ "*",
    TRUE         ~ "ns"
  )
}

# =========================================================
# Mean Annual Discharge — summary stats + t-tests per group
# =========================================================

plot_data1 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_mean_q = mean(zlog_mean_q, na.rm = TRUE),
    se_zlog_mean_q   = sd(zlog_mean_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_mean_q))),
    n                = sum(!is.na(zlog_mean_q)),
    .groups = "drop"
  )

# One-sample t-test (mean vs 0) for every av_type x years_before group.
# Pulls the raw values for each group from pre_avulsion_5 directly so
# the test uses the actual underlying data, not just the summary stats.
sig_data <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    p_value = {
      vals <- zlog_mean_q[!is.na(zlog_mean_q)]
      if (length(vals) >= 2) t.test(vals, mu = 0)$p.value else NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(sig_label = pstar_vec(p_value)) %>%
  # drop "ns" so the panel only shows stars for actual significance,
  # not a label on every single point -- remove this filter if you'd
  # rather see "ns" marked explicitly too
  filter(sig_label != "" & sig_label != "ns") %>%
  left_join(plot_data1, by = c("av_type", "years_before")) %>%
  mutate(
    # place the star just above the upper error-bar tip when the mean
    # is positive, or just below the lower error-bar tip when negative,
    # so it never overlaps the point/error bar
    star_y = ifelse(
      mean_zlog_mean_q >= 0,
      mean_zlog_mean_q + se_zlog_mean_q + SIG_GAP,
      mean_zlog_mean_q - se_zlog_mean_q - SIG_GAP
    )
  )

# =========================================================
# Plot
# =========================================================

mean_line <- ggplot(plot_data1, aes(x = years_before, y = mean_zlog_mean_q,
                                     color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_mean_q - se_zlog_mean_q,
                     ymax = mean_zlog_mean_q + se_zlog_mean_q),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.5) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  # significance stars -- colored per av_type to match that line,
  # using the same discrete color scale rather than a fixed color
  geom_text(
    data = sig_data,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Mean Discharge Z (-)"
  ) +
  grl_theme_classic()

# =========================================================
# Save — single panel, 170 x 60 mm (no legend strip needed)
# =========================================================
svglite(
  file.path(plots_dir, "fig4_mean_discharge_GRL.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(60),
  bg     = "transparent"
)
print(mean_line)
dev.off()

mean_line

# =========================================================
# Notes:
# - This single-panel figure replaces the old 3-panel + legend layout.
# - Significance stars use a one-sample t-test (mean vs 0) for each
#   av_type at each years_before value, using the raw underlying data
#   (not just the summary mean/SE), and are colored via the same
#   discrete `colors` scale as the lines/points so they visually match
#   each avulsion type.
# - "ns" results are filtered out so only significant points (p < 0.05)
#   get a star. Remove the `filter(sig_label != "ns")` line above if
#   you want "ns" explicitly labeled on non-significant points too.
# - Star placement: above the error bar when the mean is positive,
#   below it when negative, so stars never overlap the data.
# - No legend (legend.position = "none" in the theme, and the star
#   geom_text layer also has show.legend = FALSE as a safeguard).
# - Place the exported .svg at exactly 170 x 60 mm in Affinity (100%
#   scale); check the W field and switch units to mm if showing px.
# =========================================================

```

### Leadup P95 Only

```{r}
# =========================================================
# Figure 4 — Leadup to Avulsions, P95 Discharge only — GRL export
# Mirrors the Mean Discharge version: single panel, per-(av_type,
# years_before) t-tests against 0, significance stars colored to
# match each av_type's line color, positioned above/below the
# error bar depending on the sign of the mean. No legend.
# =========================================================

library(ggplot2)
library(dplyr)
library(svglite)

mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL settings ----
GRL_BASE_SIZE    <- 10     # pt
GRL_LINEWIDTH    <- 1.2    # axis lines/ticks
GRL_LINE_LW      <- 2      # geom_line weight
GRL_HLINE_LW     <- 1.2    # dashed zero-line
GRL_ERRBAR_LW    <- 0.6    # error bar weight
GRL_POINT_SIZE   <- 5      # point size
GRL_POINT_STROKE <- 1      # point stroke
GRL_SIG_SIZE     <- 14     # significance star size (pt)
SIG_GAP          <- 0.04   # vertical gap between error bar tip and star

grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) +
    theme(
      axis.title.y     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.title.x     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.line.x      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.text.y      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.ticks.y     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",   # legend removed
      plot.background  = element_rect(fill = "transparent", colour = NA),
      plot.margin      = margin(4, 4, 4, 4, unit = "mm")
    )
}

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

plots_dir <- here("plots_svg")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
# Vectorized version: case_when() evaluates element-wise across a whole
# column, unlike a plain if/else function (which only accepts a single
# scalar and errors with "condition has length > 1" when handed a vector
# inside mutate()/ifelse()).
pstar_vec <- function(p) {
  dplyr::case_when(
    is.na(p)     ~ "",
    p < 0.001    ~ "***",
    p < 0.01     ~ "**",
    p < 0.05     ~ "*",
    TRUE         ~ "ns"
  )
}

# =========================================================
# P95 Annual Discharge — summary stats + t-tests per group
# =========================================================

plot_data_p95 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_p95_q = mean(zlog_p95_q, na.rm = TRUE),
    se_zlog_p95_q   = sd(zlog_p95_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_p95_q))),
    n               = sum(!is.na(zlog_p95_q)),
    .groups = "drop"
  )

# One-sample t-test (mean vs 0) for every av_type x years_before group.
# Pulls the raw values for each group from pre_avulsion_5 directly so
# the test uses the actual underlying data, not just the summary stats.
sig_data_p95 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    p_value = {
      vals <- zlog_p95_q[!is.na(zlog_p95_q)]
      if (length(vals) >= 2) t.test(vals, mu = 0)$p.value else NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(sig_label = pstar_vec(p_value)) %>%
  # drop "ns" so the panel only shows stars for actual significance,
  # not a label on every single point -- remove this filter if you'd
  # rather see "ns" marked explicitly too
  filter(sig_label != "" & sig_label != "ns") %>%
  left_join(plot_data_p95, by = c("av_type", "years_before")) %>%
  mutate(
    # place the star just above the upper error-bar tip when the mean
    # is positive, or just below the lower error-bar tip when negative,
    # so it never overlaps the point/error bar
    star_y = ifelse(
      mean_zlog_p95_q >= 0,
      mean_zlog_p95_q + se_zlog_p95_q + SIG_GAP,
      mean_zlog_p95_q - se_zlog_p95_q - SIG_GAP
    )
  )

# =========================================================
# Plot
# =========================================================

p95_line <- ggplot(plot_data_p95, aes(x = years_before, y = mean_zlog_p95_q,
                                       color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_p95_q - se_zlog_p95_q,
                     ymax = mean_zlog_p95_q + se_zlog_p95_q),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.5) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  # significance stars -- colored per av_type to match that line,
  # using the same discrete color scale rather than a fixed color
  geom_text(
    data = sig_data_p95,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "P95 Discharge Z (-)"
  ) +
  grl_theme_classic()

# =========================================================
# Save — single panel, 170 x 60 mm (no legend strip needed)
# =========================================================
svglite(
  file.path(plots_dir, "fig4_p95_discharge_GRL.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(60),
  bg     = "transparent"
)
print(p95_line)
dev.off()

p95_line

# =========================================================
# Notes:
# - Same structure as the Mean Discharge version, swapped to
#   zlog_p95_q throughout (data, summary stats, t-tests, axis label,
#   output filename).
# - Significance stars use a one-sample t-test (mean vs 0) for each
#   av_type at each years_before value, using the raw underlying data
#   (not just the summary mean/SE), and are colored via the same
#   discrete `colors` scale as the lines/points so they visually match
#   each avulsion type.
# - "ns" results are filtered out so only significant points (p < 0.05)
#   get a star. Remove the `filter(sig_label != "ns")` line above if
#   you want "ns" explicitly labeled on non-significant points too.
# - Star placement: above the error bar when the mean is positive,
#   below it when negative, so stars never overlap the data.
# - No legend (legend.position = "none" in the theme, and the star
#   geom_text layer also has show.legend = FALSE as a safeguard).
# - Place the exported .svg at exactly 170 x 60 mm in Affinity (100%
#   scale); check the W field and switch units to mm if showing px.
# =========================================================
```

### Leadup Flood Dur Only

```{r}
# =========================================================
# Figure 4 — Leadup to Avulsions, Flood Duration only — GRL export
# Mirrors the Mean/P95 Discharge versions: single panel, per-(av_type,
# years_before) t-tests against 0, significance stars colored to
# match each av_type's line color, positioned above/below the
# error bar depending on the sign of the mean. No legend.
# =========================================================

library(ggplot2)
library(dplyr)
library(svglite)

mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL settings ----
GRL_BASE_SIZE    <- 10     # pt
GRL_LINEWIDTH    <- 1.2    # axis lines/ticks
GRL_LINE_LW      <- 2      # geom_line weight
GRL_HLINE_LW     <- 1.2    # dashed zero-line
GRL_ERRBAR_LW    <- 0.6    # error bar weight
GRL_POINT_SIZE   <- 5      # point size
GRL_POINT_STROKE <- 1      # point stroke
GRL_SIG_SIZE     <- 14     # significance star size (pt)
SIG_GAP          <- 0.04   # vertical gap between error bar tip and star

grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) +
    theme(
      axis.title.y     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.title.x     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.line.x      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.text.y      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.ticks.y     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",   # legend removed
      plot.background  = element_rect(fill = "transparent", colour = NA),
      plot.margin      = margin(4, 4, 4, 4, unit = "mm")
    )
}

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

plots_dir <- here("plots_svg")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
# Vectorized version: case_when() evaluates element-wise across a whole
# column, unlike a plain if/else function (which only accepts a single
# scalar and errors with "condition has length > 1" when handed a vector
# inside mutate()/ifelse()).
pstar_vec <- function(p) {
  dplyr::case_when(
    is.na(p)     ~ "",
    p < 0.001    ~ "***",
    p < 0.01     ~ "**",
    p < 0.05     ~ "*",
    TRUE         ~ "ns"
  )
}

# =========================================================
# Flood Duration — summary stats + t-tests per group
# =========================================================

plot_data_flood <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_flood_dur = mean(zlog_flood_dur, na.rm = TRUE),
    se_zlog_flood_dur   = sd(zlog_flood_dur, na.rm = TRUE) / sqrt(sum(!is.na(zlog_flood_dur))),
    n                   = sum(!is.na(zlog_flood_dur)),
    .groups = "drop"
  )

# One-sample t-test (mean vs 0) for every av_type x years_before group.
# Pulls the raw values for each group from pre_avulsion_5 directly so
# the test uses the actual underlying data, not just the summary stats.
sig_data_flood <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    p_value = {
      vals <- zlog_flood_dur[!is.na(zlog_flood_dur)]
      if (length(vals) >= 2) t.test(vals, mu = 0)$p.value else NA_real_
    },
    .groups = "drop"
  ) %>%
  mutate(sig_label = pstar_vec(p_value)) %>%
  # drop "ns" so the panel only shows stars for actual significance,
  # not a label on every single point -- remove this filter if you'd
  # rather see "ns" marked explicitly too
  filter(sig_label != "" & sig_label != "ns") %>%
  left_join(plot_data_flood, by = c("av_type", "years_before")) %>%
  mutate(
    # place the star just above the upper error-bar tip when the mean
    # is positive, or just below the lower error-bar tip when negative,
    # so it never overlaps the point/error bar
    star_y = ifelse(
      mean_zlog_flood_dur >= 0,
      mean_zlog_flood_dur + se_zlog_flood_dur + SIG_GAP,
      mean_zlog_flood_dur - se_zlog_flood_dur - SIG_GAP
    )
  )

# =========================================================
# Plot
# =========================================================

flooddur_line <- ggplot(plot_data_flood, aes(x = years_before, y = mean_zlog_flood_dur,
                                              color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_flood_dur - se_zlog_flood_dur,
                     ymax = mean_zlog_flood_dur + se_zlog_flood_dur),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.6) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  # significance stars -- colored per av_type to match that line,
  # using the same discrete color scale rather than a fixed color
  geom_text(
    data = sig_data_flood,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Flood Duration Z (-)"
  ) +
  grl_theme_classic()

# =========================================================
# Save — single panel, 170 x 60 mm (no legend strip needed)
# =========================================================
svglite(
  file.path(plots_dir, "fig4_flood_duration_GRL.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(60),
  bg     = "transparent"
)
print(flooddur_line)
dev.off()

flooddur_line

# =========================================================
# Notes:
# - Same structure as the Mean/P95 Discharge versions, swapped to
#   zlog_flood_dur throughout (data, summary stats, t-tests, axis
#   label, output filename).
# - Significance stars use a one-sample t-test (mean vs 0) for each
#   av_type at each years_before value, using the raw underlying data
#   (not just the summary mean/SE), and are colored via the same
#   discrete `colors` scale as the lines/points so they visually match
#   each avulsion type.
# - "ns" results are filtered out so only significant points (p < 0.05)
#   get a star. Remove the `filter(sig_label != "ns")` line above if
#   you want "ns" explicitly labeled on non-significant points too.
# - Star placement: above the error bar when the mean is positive,
#   below it when negative, so stars never overlap the data.
# - No legend (legend.position = "none" in the theme, and the star
#   geom_text layer also has show.legend = FALSE as a safeguard).
# - Place the exported .svg at exactly 170 x 60 mm in Affinity (100%
#   scale); check the W field and switch units to mm if showing px.
# =========================================================
```


### Three Panel All Type 

```{r}

# =========================================================
# Figure 4 — Leadup to Avulsions, ALL TYPES COMBINED — GRL export
# THREE PANELS ASSEMBLED INTO ONE FIGURE: P95, Mean, Flood Duration.
# Each panel: one pooled line/point series (no av_type split), with
# significance stars from a one-sample t-test (mean vs 0) at each
# years_before value. No legend needed (single series per panel).
#
# FIX: panel.background was NOT explicitly set to transparent before --
# only plot.background was. theme_classic() defaults panel.background
# to white, which is the actual axes/data area; plot.background only
# covers the margin around it. That mismatch is why the panels showed
# as solid white boxes against your black Affinity canvas even though
# plot.background = "transparent" was already set. Both are now
# transparent, so the whole figure (panel + margin) sits cleanly on
# whatever background color you place it on in Affinity.
# =========================================================

library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(svglite)

mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL settings ----
GRL_BASE_SIZE    <- 10     # pt
GRL_LINEWIDTH    <- 1.2    # axis lines/ticks
GRL_LINE_LW      <- 2      # geom_line weight
GRL_HLINE_LW     <- 1.2    # dashed zero-line
GRL_ERRBAR_LW    <- 0.6    # error bar weight
GRL_POINT_SIZE   <- 5      # point size
GRL_POINT_STROKE <- 1      # point stroke
GRL_SIG_SIZE     <- 12     # significance star size (pt) -- matches per-type version
SIG_GAP          <- 0.04   # vertical gap between error bar tip and star
LINE_COLOR       <- "#2f6fa5"   # fallback default, no longer used directly below

# ── PANEL GREYS — reversed from the Figure 3 boxplot scheme ──────────────
# Figure 3 used P95=grey25 (darkest), Mean=grey55 (mid), Flood=grey75
# (lightest). Here that order is flipped: P95 lightest, Flood darkest.
GREY_P95   <- "grey75"   # lightest -- P95
GREY_MEAN  <- "grey55"   # mid -- Mean
GREY_FLOOD <- "grey25"   # darkest -- Flood Duration

grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) +
    theme(
      axis.title.y     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.title.x     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.line.x      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.text.y      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.ticks.y     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",   # no legend needed -- single series per panel
      # BOTH panel.background and plot.background must be transparent --
      # panel.background covers the actual data/axes area, which is
      # white by default in theme_classic() and was the real culprit
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background  = element_rect(fill = "transparent", colour = NA),
      plot.margin      = margin(2, 2, 2, 2)
    )
}

plots_dir <- here("plots_v3")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar_vec <- function(p) {
  dplyr::case_when(
    is.na(p)     ~ "",
    p < 0.001    ~ "***",
    p < 0.01     ~ "**",
    p < 0.05     ~ "*",
    TRUE         ~ "ns"
  )
}

# ── REUSABLE BUILDER: pooled mean/SE + t-test + plot for one variable ────
build_pooled_panel <- function(value_col, y_label, errbar_alpha = 0.5,
                                line_color = LINE_COLOR) {

  value_sym <- rlang::sym(value_col)

  plot_data_all <- pre_avulsion_5 %>%
    mutate(years_before = avulsion_year - year) %>%
    group_by(years_before) %>%
    summarise(
      mean_val = mean(!!value_sym, na.rm = TRUE),
      se_val   = sd(!!value_sym, na.rm = TRUE) / sqrt(sum(!is.na(!!value_sym))),
      n        = sum(!is.na(!!value_sym)),
      .groups = "drop"
    )

  sig_data_all <- pre_avulsion_5 %>%
    mutate(years_before = avulsion_year - year) %>%
    group_by(years_before) %>%
    summarise(
      p_value = {
        vals <- (!!value_sym)[!is.na(!!value_sym)]
        if (length(vals) >= 2) t.test(vals, mu = 0)$p.value else NA_real_
      },
      .groups = "drop"
    ) %>%
    mutate(sig_label = pstar_vec(p_value)) %>%
    filter(sig_label != "" & sig_label != "ns") %>%
    left_join(plot_data_all, by = "years_before") %>%
    mutate(
      star_y = ifelse(
        mean_val >= 0,
        mean_val + se_val + SIG_GAP,
        mean_val - se_val - SIG_GAP
      )
    )

  ggplot(plot_data_all, aes(x = years_before, y = mean_val)) +
    geom_line(linewidth = GRL_LINE_LW, alpha = 0.7, color = line_color) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
    geom_errorbar(aes(ymin = mean_val - se_val, ymax = mean_val + se_val),
                  width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = errbar_alpha, color = line_color) +
    geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white",
               fill = line_color, shape = 21, alpha = 0.7) +
    geom_text(
      data = sig_data_all,
      aes(x = years_before, y = star_y, label = sig_label),
      inherit.aes = FALSE, color = line_color, size = GRL_SIG_SIZE / .pt,
      fontface = "bold"
    ) +
    scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
    coord_cartesian(ylim = c(-0.5, 0.5)) +
    labs(x = "Time to Avulsion (yrs)", y = y_label) +
    grl_theme_classic()
}

# =========================================================
# Build all three panels
# =========================================================
p95_line_all      <- build_pooled_panel("zlog_p95_q",     "Peak Discharge Z (-)",
                                         errbar_alpha = 0.5, line_color = GREY_P95)
mean_line_all     <- build_pooled_panel("zlog_mean_q",    "Mean Discharge Z (-)",
                                         errbar_alpha = 0.5, line_color = GREY_MEAN)
flooddur_line_all <- build_pooled_panel("zlog_flood_dur", "Flood Duration Z (-)",
                                         errbar_alpha = 0.6, line_color = GREY_FLOOD)

# =========================================================
# Assemble: 3 panels side by side, no legend strip needed
# Final size: 170 mm wide x 65 mm tall
# =========================================================

combined_grob <- arrangeGrob(
  p95_line_all, mean_line_all, flooddur_line_all,
  ncol = 3,
  widths = unit(c(1, 1, 1), "null")  # force exactly equal column widths,
                                       # overriding ggplot's default of
                                       # sizing each column based on its
                                       # own axis label text width
)

svglite(
  file.path(plots_dir, "fig4_all_types_combined_3panel_GRL.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(65),
  bg     = "transparent"
)
grid.newpage()
grid.draw(combined_grob)
dev.off()

grid.newpage()
grid.draw(combined_grob)

```
## Trials 

### Trial 1

```{r}

# =========================================================
# Figure 4 — Leadup to Avulsions, ALL TYPES COMBINED — GRL export
# THREE PANELS ASSEMBLED INTO ONE FIGURE: P95, Mean, Flood Duration.
# Each panel: one pooled line/point series (no av_type split), with
# significance stars from a one-sample t-test (mean vs 0) at each
# years_before value. No legend needed (single series per panel).
#
# FIX: panel.background was NOT explicitly set to transparent before --
# only plot.background was. theme_classic() defaults panel.background
# to white, which is the actual axes/data area; plot.background only
# covers the margin around it. That mismatch is why the panels showed
# as solid white boxes against your black Affinity canvas even though
# plot.background = "transparent" was already set. Both are now
# transparent, so the whole figure (panel + margin) sits cleanly on
# whatever background color you place it on in Affinity.
# =========================================================

library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(svglite)

mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL settings ----
GRL_BASE_SIZE    <- 10     # pt
GRL_LINEWIDTH    <- 1.2    # axis lines/ticks
GRL_LINE_LW      <- 2      # geom_line weight
GRL_HLINE_LW     <- 1.2    # dashed zero-line
GRL_ERRBAR_LW    <- 0.6    # error bar weight
GRL_POINT_SIZE   <- 5      # point size
GRL_POINT_STROKE <- 1      # point stroke
GRL_SIG_SIZE     <- 12     # significance star size (pt) -- matches per-type version
SIG_GAP          <- 0.04   # vertical gap between error bar tip and star
LINE_COLOR       <- "#2f6fa5"   # single series color, used across all 3 panels

grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) +
    theme(
      axis.title.y     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.title.x     = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.line.x      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y      = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.text.y      = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.ticks.y     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x     = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",   # no legend needed -- single series per panel
      # BOTH panel.background and plot.background must be transparent --
      # panel.background covers the actual data/axes area, which is
      # white by default in theme_classic() and was the real culprit
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background  = element_rect(fill = "transparent", colour = NA),
      plot.margin      = margin(2, 2, 2, 2)
    )
}

plots_dir <- here("plots_v3")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
pstar_vec <- function(p) {
  dplyr::case_when(
    is.na(p)     ~ "",
    p < 0.001    ~ "***",
    p < 0.01     ~ "**",
    p < 0.05     ~ "*",
    TRUE         ~ "ns"
  )
}

# ── REUSABLE BUILDER: pooled mean/SE + t-test + plot for one variable ────
build_pooled_panel <- function(value_col, y_label, errbar_alpha = 0.5) {

  value_sym <- rlang::sym(value_col)

  plot_data_all <- pre_avulsion_5 %>%
    mutate(years_before = avulsion_year - year) %>%
    group_by(years_before) %>%
    summarise(
      mean_val = mean(!!value_sym, na.rm = TRUE),
      se_val   = sd(!!value_sym, na.rm = TRUE) / sqrt(sum(!is.na(!!value_sym))),
      n        = sum(!is.na(!!value_sym)),
      .groups = "drop"
    )

  sig_data_all <- pre_avulsion_5 %>%
    mutate(years_before = avulsion_year - year) %>%
    group_by(years_before) %>%
    summarise(
      p_value = {
        vals <- (!!value_sym)[!is.na(!!value_sym)]
        if (length(vals) >= 2) t.test(vals, mu = 0)$p.value else NA_real_
      },
      .groups = "drop"
    ) %>%
    mutate(sig_label = pstar_vec(p_value)) %>%
    filter(sig_label != "" & sig_label != "ns") %>%
    left_join(plot_data_all, by = "years_before") %>%
    mutate(
      star_y = ifelse(
        mean_val >= 0,
        mean_val + se_val + SIG_GAP,
        mean_val - se_val - SIG_GAP
      )
    )

  ggplot(plot_data_all, aes(x = years_before, y = mean_val)) +
    geom_line(linewidth = GRL_LINE_LW, alpha = 0.7, color = LINE_COLOR) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
    geom_errorbar(aes(ymin = mean_val - se_val, ymax = mean_val + se_val),
                  width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = errbar_alpha, color = LINE_COLOR) +
    geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white",
               fill = LINE_COLOR, shape = 21, alpha = 0.7) +
    geom_text(
      data = sig_data_all,
      aes(x = years_before, y = star_y, label = sig_label),
      inherit.aes = FALSE, color = LINE_COLOR, size = GRL_SIG_SIZE / .pt,
      fontface = "bold"
    ) +
    scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
    coord_cartesian(ylim = c(-0.5, 0.5)) +
    labs(x = "Time to Avulsion (yrs)", y = y_label) +
    grl_theme_classic()
}

# =========================================================
# Build all three panels
# =========================================================
p95_line_all      <- build_pooled_panel("zlog_p95_q",     "Peak Discharge Z (-)",   errbar_alpha = 0.5)
mean_line_all     <- build_pooled_panel("zlog_mean_q",    "Mean Discharge Z (-)",   errbar_alpha = 0.5)
flooddur_line_all <- build_pooled_panel("zlog_flood_dur", "Flood Duration Z (-)",   errbar_alpha = 0.6)

# =========================================================
# Assemble: 3 panels side by side, no legend strip needed
# Final size: 170 mm wide x 65 mm tall
# =========================================================

combined_grob <- arrangeGrob(
  p95_line_all, mean_line_all, flooddur_line_all,
  ncol = 3,
  widths = unit(c(1, 1, 1), "null")  # force exactly equal column widths,
                                       # overriding ggplot's default of
                                       # sizing each column based on its
                                       # own axis label text width
)

svglite(
  file.path(plots_dir, "t1.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(65),
  bg     = "transparent"
)
grid.newpage()
grid.draw(combined_grob)
dev.off()

grid.newpage()
grid.draw(combined_grob)

# =========================================================
# Notes:
# - panel.background is now explicitly transparent (the actual fix for
#   the white-box-on-black-canvas issue). plot.background remains
#   transparent too. Both must be set -- they cover different areas.
# - Stacking with fig4_per_type_with_sig_stars_GRL.svg: confirm that
#   script ALSO has panel.background = element_rect(fill =
#   "transparent", colour = NA) explicitly set in its theme function --
#   if it was only relying on theme_minimal()/theme_classic() defaults
#   plus plot.background, it likely has the same white-panel issue,
#   just masked if it was exported with bg = "white" at the svglite()
#   call level (which fills the WHOLE canvas white, panel and margin
#   alike, hiding the contrast). Check that file's grl_theme_classic()
#   and svglite() bg argument and apply the same fix if needed.
# - Place the exported .svg at exactly 170 x 65 mm in Affinity (100%
#   scale); check the W field and switch units to mm if showing px.
# =========================================================

```


```{r}

# =========================================================
# Figure 4 — Leadup to Avulsions — GRL-compliant export
# Target: 170 mm wide, 6 pt text at final size (per AGU/GRL spec)
# 3 panels side by side (~55 mm each + gutters), shared bottom legend
# Line/point weights kept close to original visual weight (not capped by
# the 6pt rule, which applies to TEXT only).
#
# ADDED: per-(av_type, years_before) one-sample t-tests (mean vs 0),
# with significance stars colored to match each av_type's line color
# (same `colors` discrete scale used by the lines/points). Error bars
# remain STANDARD ERROR of the mean (sd/sqrt(n)), same as originally.
# =========================================================

library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(svglite)

mm_to_in <- function(mm) mm / 25.4

# ---- shared GRL settings ----
GRL_BASE_SIZE  <- 10      # pt, per AGU/GRL spec -- do not increase
GRL_LINEWIDTH  <- 1.2    # axis lines/ticks -- original weight
GRL_LINE_LW    <- 2      # geom_line weight -- original
GRL_HLINE_LW   <- 1.2    # dashed zero-line -- original
GRL_ERRBAR_LW  <- 0.6    # error bar weight -- original
GRL_POINT_SIZE <- 5      # point size -- original
GRL_POINT_STROKE <- 1    # point stroke -- original
GRL_SIG_SIZE   <- 12     # significance star size (pt)
SIG_GAP        <- 0.04   # vertical gap between error bar tip and star

grl_theme_classic <- function() {
  theme_classic(base_size = GRL_BASE_SIZE) +
    theme(
      axis.title.y = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.title.x = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.line.x  = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.line.y  = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.text.x  = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.text.y  = element_text(color = "grey30", size = GRL_BASE_SIZE, face = "bold"),
      axis.ticks.y = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      axis.ticks.x = element_line(color = "grey30", linewidth = GRL_LINEWIDTH),
      legend.position  = "none",
      plot.background  = element_rect(fill = "transparent", colour = NA),
      plot.margin      = margin(2, 2, 2, 2)
    )
}

colors <- plasma(4)
shapes <- c(21, 22, 23, 24)

plots_dir <- here("plots_v3")

# ── SIGNIFICANCE HELPER ───────────────────────────────────────────────────
# Vectorized: case_when() evaluates element-wise across a whole column,
# unlike a plain if/else function (which only accepts a single scalar
# and errors with "condition has length > 1" inside mutate()).
pstar_vec <- function(p) {
  dplyr::case_when(
    is.na(p)     ~ "",
    p < 0.001    ~ "***",
    p < 0.01     ~ "**",
    p < 0.05     ~ "*",
    TRUE         ~ "ns"
  )
}

# ── REUSABLE HELPER: per-(av_type, years_before) t-tests + star placement ──
build_sig_data <- function(value_col, summary_df, mean_col, se_col) {
  value_sym <- rlang::sym(value_col)
  mean_sym  <- rlang::sym(mean_col)
  se_sym    <- rlang::sym(se_col)

  pre_avulsion_5 %>%
    mutate(years_before = avulsion_year - year) %>%
    group_by(av_type, years_before) %>%
    summarise(
      p_value = {
        vals <- (!!value_sym)[!is.na(!!value_sym)]
        if (length(vals) >= 2) t.test(vals, mu = 0)$p.value else NA_real_
      },
      .groups = "drop"
    ) %>%
    mutate(sig_label = pstar_vec(p_value)) %>%
    # drop "ns" so the panel only shows stars for actual significance --
    # remove this filter if you'd rather see "ns" marked explicitly too
    filter(sig_label != "" & sig_label != "ns") %>%
    left_join(summary_df, by = c("av_type", "years_before")) %>%
    mutate(
      star_y = ifelse(
        !!mean_sym >= 0,
        !!mean_sym + !!se_sym + SIG_GAP,
        !!mean_sym - !!se_sym - SIG_GAP
      )
    )
}

# =========================================================
# Panel 1 — Log P95 Annual Discharge
# =========================================================

plot_data <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_p95_q = mean(zlog_p95_q, na.rm = TRUE),
    se_zlog_p95_q   = sd(zlog_p95_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_p95_q))),
    n               = sum(!is.na(zlog_p95_q)),
    .groups = "drop"
  ) %>%
  mutate(
    ci_mult  = qt(0.975, df = pmax(n - 1, 1)),
    ci_lower = mean_zlog_p95_q - ci_mult * se_zlog_p95_q,
    ci_upper = mean_zlog_p95_q + ci_mult * se_zlog_p95_q
  )

sig_data_p95 <- build_sig_data("zlog_p95_q", plot_data, "mean_zlog_p95_q", "se_zlog_p95_q")

p95_line <- ggplot(plot_data, aes(x = years_before, y = mean_zlog_p95_q,
                                   color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_p95_q - se_zlog_p95_q,
                     ymax = mean_zlog_p95_q + se_zlog_p95_q),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.5) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  # significance stars -- colored per av_type to match that line, using
  # the same discrete color scale as the rest of the panel
  geom_text(
    data = sig_data_p95,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.6, 0.6)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Peak Discharge Z (-)",
    color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type"
  ) +
  grl_theme_classic()

# =========================================================
# Panel 2 — Log Mean Annual Discharge
# =========================================================

plot_data1 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_mean_q = mean(zlog_mean_q, na.rm = TRUE),
    se_zlog_mean_q   = sd(zlog_mean_q, na.rm = TRUE) / sqrt(sum(!is.na(zlog_mean_q))),
    .groups = "drop"
  )

sig_data_mean <- build_sig_data("zlog_mean_q", plot_data1, "mean_zlog_mean_q", "se_zlog_mean_q")

mean_line <- ggplot(plot_data1, aes(x = years_before, y = mean_zlog_mean_q,
                                     color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_mean_q - se_zlog_mean_q,
                     ymax = mean_zlog_mean_q + se_zlog_mean_q),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.5) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  geom_text(
    data = sig_data_mean,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.6, 0.)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Mean Discharge Z (-)",
    color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type"
  ) +
  grl_theme_classic()

# =========================================================
# Panel 3 — Log Flood Duration
# =========================================================

plot_data2 <- pre_avulsion_5 %>%
  mutate(years_before = avulsion_year - year) %>%
  group_by(av_type, years_before) %>%
  summarise(
    mean_zlog_flood_dur = mean(zlog_flood_dur, na.rm = TRUE),
    se_zlog_flood_dur   = sd(zlog_flood_dur, na.rm = TRUE) / sqrt(sum(!is.na(zlog_flood_dur))),
    .groups = "drop"
  )

sig_data_flood <- build_sig_data("zlog_flood_dur", plot_data2, "mean_zlog_flood_dur", "se_zlog_flood_dur")

flooddur_line <- ggplot(plot_data2, aes(x = years_before, y = mean_zlog_flood_dur,
                                         color = av_type, fill = av_type, shape = av_type)) +
  geom_line(linewidth = GRL_LINE_LW, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", linewidth = GRL_HLINE_LW) +
  geom_errorbar(aes(ymin = mean_zlog_flood_dur - se_zlog_flood_dur,
                     ymax = mean_zlog_flood_dur + se_zlog_flood_dur),
                width = 0.15, linewidth = GRL_ERRBAR_LW, alpha = 0.6) +
  geom_point(size = GRL_POINT_SIZE, stroke = GRL_POINT_STROKE, color = "white", alpha = 0.7) +
  geom_text(
    data = sig_data_flood,
    aes(x = years_before, y = star_y, label = sig_label, color = av_type),
    inherit.aes = FALSE, size = GRL_SIG_SIZE / .pt, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_reverse(breaks = 1:5, labels = paste0(-1:-5)) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  coord_cartesian(ylim = c(-0.6, 0.6)) +
  labs(
    x = "Time to Avulsion (yrs)",
    y = "Flood Duration Z (-)",
    color = "Avulsion Type", fill = "Avulsion Type", shape = "Avulsion Type"
  ) +
  grl_theme_classic()

# =========================================================
# Assemble: 3 panels in a row, NO legend
# Final size: 170 mm wide x 65 mm tall -- matches the all-types-
# combined script exactly, so the two figures stack with identical
# panel heights (the previous version's panels were sized against an
# 83mm-tall region within a 100mm canvas, due to the legend's
# heights = c(5, 1) split -- that's why "chopping off" the legend
# visually didn't actually fix the size mismatch. This version lays
# the panels out against a 65mm canvas from the start, so there's no
# leftover legend-shaped gap in the math.)
# =========================================================

combined_grob <- arrangeGrob(
  p95_line, mean_line, flooddur_line,
  ncol = 3,
  widths = unit(c(1, 1, 1), "null")  # force exactly equal column widths
)

svglite(
  file.path(plots_dir, "t2.svg"),
  width  = mm_to_in(170),
  height = mm_to_in(65),
  bg     = "transparent"
)
grid.newpage()
grid.draw(combined_grob)
dev.off()

grid.newpage()
grid.draw(combined_grob)

# =========================================================
# Notes for Affinity assembly:
# - Place this SVG at exactly 170 x 65 mm (100% scale) -- do not resize
#   further, or the 6pt text drifts off the GRL spec.
# - This now matches fig4_all_types_combined_3panel_GRL.svg exactly in
#   both width AND height (170 x 65 mm both), with equal column widths
#   in both -- the two figures should stack with identically-sized
#   panels and aligned x-axes.
# - Line/point weights were kept at original visual scale on purpose;
#   only TEXT is constrained to 6pt by the AGU/GRL rule.
# - Significance stars: one-sample t-test (mean vs 0) per av_type per
#   years_before value, colored to match that av_type's line via the
#   same `colors` scale. "ns" results are filtered out -- only
#   significant points (p < 0.05) get a star. Remove the
#   filter(sig_label != "ns") line in build_sig_data() if you want
#   "ns" explicitly labeled too.
# - Error bars remain STANDARD ERROR of the mean (sd/sqrt(n)), matching
#   the original convention -- consistent with what the t-tests use
#   internally, so overlapping-zero bars roughly track non-significant
#   points.
# - No legend in this export. If you need the avulsion-type color key
#   for the reader, add it separately in Affinity as a small text/swatch
#   legend, or ask me to generate one as its own standalone SVG sized
#   independently (so it never affects the panel dimensions again).
# - For final submission, export the assembled figure from Affinity as
#   PDF or EPS with fonts outlined/converted to curves (AGU requirement).
# =========================================================
#   PDF or EPS with fonts outlined/converted to curves (AGU requirement).
# =========================================================

```

































